### 2022-09-03 paper.tex

\documentclass[12pt,a4paper,man]{apa7}
\usepackage{listings}

\author{Florian Donnelly}
\title{The Creation of a Programming Language}
\date{August 2022}
\affiliation{Gymnasium Burgdorf}
\shorttitle{Creation of a Programming Language}
\abstract{
    This is an abstract...
}

% Defining custom commands
\newcommand{\code}[3] {
    \lstset{language=C,numbers=left,basicstyle=\scriptsize,
        frame=single,caption=src/#1,keywordstyle=\color{blue},
        stringstyle=\color{green},commentstyle=\color{gray},
        morecomment=[l][\color{magenta}]{\#}}
    \lstinputlisting{src/#1}
}

\begin{document}
\maketitle

\tableofcontents\newpage

\section{Introduction}

\code{main.c}

\section{Theory}
\section{Implementation}
\section{Results}
\section{Discussion}
\section{Index}
\section{Closing Words}
\section{Declaration of Autonomy}
\section{Appendix}

\end{document}

### 2022-09-03 paper.tex
paper make target, continue writing paper, add bibliography

@@
 \usepackage{listings}
+\usepackage{setspace}
+\usepackage{times} % Use Times New Roman font
+\usepackage[style=apa]{biblatex}
+
+\addbibresource{bibliography.bib}
 
@@
 \title{The Creation of a Programming Language}
-\date{August 2022}
-\affiliation{Gymnasium Burgdorf}
+\authorsaffiliations{Gymnasium Burgdorf}
 \shorttitle{Creation of a Programming Language}
@@
 % Defining custom commands
+
+% include C code from file #1 from lines #2 to #3
 \newcommand{\code}[3] {
@@
         morecomment=[l][\color{magenta}]{\#}}
-    \lstinputlisting{src/#1}
+    \lstinputlisting[firstnumber=#2,firstline=#2,lastline=#3]{src/#1}
 }
@@
 \begin{document}
-\maketitle
 
-\tableofcontents\newpage
+\onehalfspacing
+\maketitle\tableofcontents\newpage
+
+% the name of the programming language
+\newcommand{\name}{-the language- }
 
 \section{Introduction}
+\subsection{Relevance of the Topic}
+\subsection{Goals}
 
-\code{main.c}
+\code{main.c}{2}{6}
 
 \section{Theory}
+\subsection{What this is and what this is not}
+The final product in the process of creating this programming language will 
+not be perfect, or in fact, be used or adopted by anyone at all, and it is not
+intended to be. Creating a good, or even usable, programming language 
+implementation by today's  standards means to create an entire ecosystem of 
+tools, not only including a compiler or interpreter, but also a package manager
+for dealing with dependencies and a platform for people to share their work.
+
+The process used in this paper is more so a very core implementation of what
+a programming language, only containing the very necessary features.
+
 \section{Implementation}
+\subsection{Used Software}
+At the heart of creating a program in C stands the C compiler. This project uses
+\emph{gcc}, the GNU Compiler Collection, which, as the name suggests, also 
+supports different languages than C. Substantiating the development process is 
+\emph{GNU Make}, a general-purpose build system. It handles tasks like building
+the executable step-by-step from the project source code and directory clean-up.
+The C ecosystem is very easy to use and fundamentally supported by 
+\emph{GNU+Linux}, the operating system used to program on.
+The main code editor used was \emph{Microsoft Visual Studio Code}, providing great
+features like built-in git integration.
+\emph{Git} is the project management and tracking software used to back-up and share
+the project files on \emph{GitHub.com}, a Microsoft hosted git repository server to
+store and collaborate on source code.
+
 \section{Results}
 \section{Discussion}
+
+\subsection{Interesting Ideas untouched}
+Creating the ideal programming language takes a lot of planning and sketching
+out. This section of the paper gives insight on interesting ideas or concepts
+that are used in other programming languages, not \name.
+
+\subsubsection{Foreign Function Interfaces}
+A foreign function interface\footcite{FFI} often abbreviated as FFI,
+is a mechanism by which one programming language can call functions written 
+in another. This makes a language more attractive, as it can be used with
+existing libraries from another language, instead of functions having to be
+re-written. Enabling C function calls often allows a certain language to
+directly interact with system libraries or the operating system itsself.
+
+Foreign function interfaces are available in many modern languages.
+Examples are Java, Python and Rust.
+Java enables users to call C, C++ and even assembly code with the Java Native
+Interface\footcite{JNI}, JNI for short, whereas within Python C function can
+be run with the standard ctypes\footcite{ctypes} library.
+
+Foreign function interfaces are specially difficult to implement, or implement
+right. Let us assume a programming language interpreter \name written in C that
+would like to call some functions that are part of yet another C program or
+library. The problem lies in the fact that at runtime, C does not know
+structure of the functions to be called, say return- or argument-types, which
+is by the way where the term foreign functions originates from. This is why the
+\name developer would have to provide that information by re-declaring the
+function inside \name, so that, with some trickery, data can be passed to and
+received from the function in the correct format, preventing a segmentation
+fault. The libffi\footcite{libffi} C-library, which is also used by Java and
+Python for this purpose, can be used to load and call such foreign functions.
+
+\subsubsection{Reflection}
+Reflection\footcite{reflection}, also called reflective programming, describes
+the ability of a program to examine and modify its own structure and behaviour 
+at runtime. An example would be self-modifying code. This can easily be
+achieved in assembly language, which inherently does not differenciate between
+data and instructions.
+Another example is Java, providing methods and classes in the java.lang.reflect
+package that enable developers to examine and change properties or functions
+of objects.
+The advantage of reflection over not having it is that it allows for non-static
+programs and program flows, effectively making smaller executables and more
+resource efficient programs, as those programs can adjust themselfs to run
+differently under different conditions at runtime.
+The easiest and simplest way to go about implementing Reflection in \name is
+certainly to allow the interpretation of strings at runtime. That way, strings
+holding program instructions can be created and modified and later executed.
+
 \section{Index}
@@
 \section{Declaration of Autonomy}
+Hereby I, Florian Donnelly, declare to have written this paper, also including
+the entire source code of \name by myself using resources listed under the
+references section.
+
 \section{Appendix}
 
+\newpage\printbibliography[heading=bibintoc]
+
 \end{document}

### 2022-09-03 src/interpreter/bytecode.c
fix function can use deleted variable, untrack pdf

@@ -13,6 +13,9 @@ static inline mementry *_register(bytecode *code, stackptr ptr, typing type) {
     return ret;
 }
 
+// TODO: does not work for multi-thread
+table_level max_level;
+
 // see protocol specifications in header file
 void _recursiveconsume(bytecode *dst, stackptr ptr, stnode *subtree) {
     switch(subtree->type){
@@ -29,11 +32,17 @@ void _recursiveconsume(bytecode *dst, stackptr ptr, stnode *subtree) {
 
             break;
 
-        case FIELD:
+        case FIELD: {
             // just copy the mementry pointer
-            _register(dst, ptr, REFERENCE)->value = subtree->data.leaf.value;
+            mementry *fielddst = _register(dst, ptr, REFERENCE);
+            fielddst->value = subtree->data.leaf.value;
+            fielddst->level = ((mementry *)fielddst->value)->level;
 
-            break;
+            // reassign max_field
+            if(fielddst->level > max_level)
+                max_level = fielddst->level;
+            
+            break; }
 
         case STRING:
             // push pointer and don't free the orignial memory
@@ -74,6 +83,9 @@ void _recursiveconsume(bytecode *dst, stackptr ptr, stnode *subtree) {
 
             size_t offset = 0;
 
+            // reset max level
+            max_level = 0;
+
             // Consume a member chain into the stack
             stnode *tree;
             while(subtree != NULL && subtree->data.parent.left != NULL) {
@@ -92,7 +104,10 @@ void _recursiveconsume(bytecode *dst, stackptr ptr, stnode *subtree) {
             fit_stack(function);
 
             // push function pointer to stack
-            _register(dst, ptr, FUNCTION)->value = function;
+            mementry *fundst = _register(dst, ptr, FUNCTION);
+            fundst->value = function;
+            // set functions maximally used level
+            fundst->level = max_level;
 
             break;
         }

### 2022-09-03 src/interpreter/error/error.c

@@ -22,6 +22,8 @@ static const char *errformats[] = {
     "The specified operator cannot be used on of given type(s)",
     "The specified operand is not a function, therefore it cannot be called",
     "The specified operand must be mutable",
+    "You can only assign the value from one field to another, if the source "
+        "field was created in a scope equal or lower than the destination field",
 
     "Loading of a shared object failed",
     "Loading of foreign function failed",

### 2022-09-03 src/interpreter/error/error.h

@@ -26,6 +26,7 @@ enum errorformat {
     EI_WRONG_TYPE,
     EI_NOT_CALLABLE,
     EI_REQ_MUTABLE,
+    EI_UNLEVELED,
 
     // Foreign function errors
     EF_DL_LOADING_FAILED,

### 2022-09-03 src/interpreter/localizer.c

@@ -51,7 +51,7 @@ void localize(stnode *subtree, hashtable *table, errorinfo *info) {
             ht_down(table);
             break;
 
-        case FIELD:
+        case FIELD: {
             // resolve field into reference
             char *key = subtree->data.leaf.value;
             subtree->data.leaf.value = find(table, key);
@@ -59,7 +59,7 @@ void localize(stnode *subtree, hashtable *table, errorinfo *info) {
             // free the key
             free(key);
 
-            break;
+            break; }
 
         default:
             // Ignore

### 2022-09-03 src/interpreter/mappings/mapop.c

@@ -45,7 +45,7 @@
 #line 23 "./mapop.gperf"
 struct s_op;
 
-#define TOTAL_KEYWORDS 34
+#define TOTAL_KEYWORDS 25
 #define MIN_WORD_LENGTH 1
 #define MAX_WORD_LENGTH 2
 #define MIN_HASH_VALUE 1
@@ -67,16 +67,16 @@ hash (register const char *str, register size_t len)
       63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
       63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
       63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
-      63, 63, 63, 22, 63, 63, 63, 17, 25, 63,
-      63, 63, 12, 20, 40,  5, 63,  7, 63, 63,
-      63, 63, 63, 63, 63, 63, 63, 63, 30, 10,
-       2,  0, 55, 63, 63, 63, 63, 63, 63, 63,
+      63, 63, 63,  8, 63, 63, 63, 31, 30, 63,
+      63, 63, 26, 25,  6,  5, 63, 28, 63, 63,
+      63, 63, 63, 63, 63, 63, 63, 63, 23, 18,
+       3,  0, 15, 63, 63, 63, 63, 63, 63, 63,
       63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
       63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
-      63, 63, 63, 63, 45, 63, 63, 63, 63, 63,
+      63, 63, 63, 63, 13, 63, 63, 63, 63, 63,
       63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
       63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
-      63, 63, 63, 63, 15, 63, 35, 63, 63, 63,
+      63, 63, 63, 63, 20, 63, 10, 63, 63, 63,
       63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
       63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
       63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
@@ -115,82 +115,65 @@ lookup (register const char *str, register size_t len)
       {"=",    OP_ASS,     INFIX,   RTL, 3     /*  Assign                    */},
 #line 39 "./mapop.gperf"
[…]

### 2022-09-03 src/interpreter/processing/implementations.c

@@ -232,13 +232,29 @@ mementry *_lambda(opargs *args){
 }
 
 mementry *_ass(opargs *args){
-    mementry *src = _recursiveprocess(args, DEREFERENCE); // Load right operand into src
-    mementry *dst = _recursiveprocess(args, MUTABLE); // Load left operand into dst
+    mementry *src = _recursiveprocess(args, 0); // Load right operand into src
+    mementry *dst = _recursiveprocess(args, 0); // Load left operand into dst
 
-    if(dst == NULL) {
+    if(dst == NULL || dst->type != REFERENCE) {
         throw(EI_REQ_MUTABLE, args->info);
         return NULL;
     }
+
+    if(src->type == REFERENCE || src->type == FUNCTION) {
+        // make sure that dst exists higher than source
+        if(src->value != NULL && dst->level < src->level) {
+            throw(EI_UNLEVELED, args->info);
+            return NULL;
+        }
+
+    }
+
+    if(src->type == REFERENCE)
+        // dereference field
+        src = src->value;
+
+    // dereference field
+    dst = dst->value;
 
     // copy type
     dst->type = src->type;

### 2022-09-04 src/interpreter/error/error.h
implement std.if

@@ -48,8 +48,6 @@ typedef struct {
         unsigned int line, character;
     } fileinfo;
 
-    // TODO: add thread info & name
-
 } errorinfo;
 
 // Throw selected error, print it to stderr

### 2022-09-04 src/interpreter/libraries/stdlib/stdlib.c

@@ -1,5 +1,8 @@
 #include "stdlib.h"
+#include "../../processing/implementations.h"
+#include "../../error/error.h"
 #include "../../bytecode.h"
+#include "../../memory/array.h"
 
 #define ALLOC 1
 
@@ -16,8 +19,6 @@ void _type(mementry *args, mementry *dst) {
 
     dst->type = STRING;
     dst->value = strdup(typeNames[args->type]);
-
-    return dst;
 }
 
 // print a value to the console
@@ -37,13 +38,12 @@ void _print(mementry *args, mementry *dst) {
             dst->type = NUMBER;
             dst->value = malloc(sizeof(number));
             *((number *)dst->value) = (number) 0;
-            return dst;
+            return;
     }
 
     dst->type = NUMBER;
     dst->value = malloc(sizeof(number));
     *((number *)dst->value) = (number) 1;
-    return dst;
 }
 
 // execute a system shell comand (sh on UNIX, cmd on Windows)
@@ -53,13 +53,68 @@ void _exec(mementry *args, mementry *dst) {
 
     if(args->type != STRING){
         *((number *)dst->value) = (number) 0;
-        return dst;
+        return;
     }
     
     system((char*) args->value);
     
     *((number *)dst->value) = (number) 1;
-    return dst;
+}
+
+void _if(mementry *args, mementry *dst) {
+    if(args->type == TUPLE){
+        array *arr = ((array *)args->value);
+        mementry *fun = NULL;
+        // can be (condition, if) or (condition, if, else)
+        if(arr->size == 2 || arr->size == 3) {
+            mementry *cond = arr->arr[0];
+            if(cond->type == NUMBER && *((number *)cond->value) != 0 ||
+                    cond->type == STRING && *((char *)args->value) != '\0') {
+                // perform if case
+                fun = arr->arr[1];
+            } else if(arr->size == 3) {
+                // perform else case 
+                fun = arr->arr[2];
+            }        
+
+            if(fun != NULL && fun->type == FUNCTION){
+                // same as in implementations.c
+                // create new opargs for virtual environment
+                opargs new_args;
+
+                // reference the function code stack
+                new_args.code = (bytecode *)fun->value;
+
+                // TODO: put arguments into table
+
+                // print any errors from pseudo-if 
+                errorinfo info = {.throwable = 0, .fileinfo=
+                    {.filename="<if-statement>", .line=0, .character=0}};
+                new_args.info = &info;
+
[…]

### 2022-09-04 src/interpreter/parser.c

@@ -89,66 +89,7 @@ void parser_init(cache *cache) {
 
 // Define the following functions
 stnode *expr(cache *cache, unsigned char rbp);
-stnode *secondary(cache *cache);
-
-
-// Builds a member tree of multiple expressions
-stnode *subexpr(cache *cache, unsigned char rbp) {
-    stnode *member;
-    
-    member = expr(cache, rbp);
-
-    // Parse in-expression code blocks
-    if(member != NULL && member->type == BLOCK){
-        // loop variable
-        stnode *left, *last;
-
-        // change type from block to member
-        member->type = MEMBER;
-
-        // Set right pointer to NULL
-        member->data.parent.right = NULL;
-
-        // keep track of level
-        unsigned int level = 1;
-
-        // left child is expression or BLOCK_END
-        // right child is member pointer or NULL
-
-        char flag = 0;
-
-        while(1) {
-            left = parse(cache);
-
-            // pass on errors / EOF and ignore sub-scope
-            if(left->type == FILE_END)
-                return left;
-
-            // keep track of level
-            else if(left->type == BLOCK)
-                level++;
-            else if(left->type == BLOCK_END){
-                level--;
-                if(level == 0) 
-                    break;
-            }
-
-            // skip this once
-            if(flag) {    
-                // reverse direction
-                last = member;
-                member = allocate_typed(MEMBER);
-                member->data.parent.right = last;                  
-            } else
-                flag = 1;
-
-            member->data.parent.left = left;
-
-        }
-
-    }
-    return member;
-}
+stnode *secondary(cache *cache, unsigned char rbp);
 
 // parser-defined operators
 operator virtualops[] = {
@@ -158,9 +99,11 @@ operator virtualops[] = {
 };
 
 // Parses a single expression
+char in_expression; // does not work for multi-thread
 stnode *expr(cache *cache, unsigned char rbp) {
 
-    stnode *left = secondary(cache);
+    stnode *left = secondary(cache, rbp);
+    in_expression = 1;
 
     // If secondary is done, just return
     if(left == NULL || left->type == FILE_END)
@@ -186,7 +129,7 @@ stnode *expr(cache *cache, unsigned char rbp) {
             advance(cache);
 
             if(op->position == INFIX){
-                new->data.parent.right = subexpr(cache, op->precedence - op->associativity);
+                new->data.parent.right = expr(cache, op->precedence - op->associativity);
             } else {
[…]

### 2022-09-04 src/interpreter/processing/implementations.c

@@ -18,7 +18,7 @@
 #define MOD(a, b) (number) fmodl(a, b)
 
 // free a synthetic mementry and its synthetic value
-static inline void _free_synth(mementry *entry) {
+void _free_synth(mementry *entry) {
     if(entry->flags.synthetic) {
         if(entry->flags.value_synthetic)
             free(entry->value);
@@ -367,7 +367,8 @@ mementry *_call(opargs *args){
             else {
                 // type and value will be set by the called function
                 dst = malloc(sizeof(mementry));
-                dst->flags = (struct s_mementry_flags) {.mutable=0, .synthetic=1, .value_synthetic=1};
+                dst->flags = (struct s_mementry_flags) 
+                    {.mutable=0, .synthetic=1, .value_synthetic=1};
             }
 
             // call function from shared object

### 2022-09-04 src/interpreter/processing/implementations.h

@@ -7,6 +7,7 @@
 // @param return_reference if set, a reference from hashtable is returned (and element is allocated)
 //  else, a copy is returned
 mementry *_recursiveprocess(opargs *args, char return_reference);
+void _free_synth(mementry *entry);
 
 // ----- Declarations -----
 mementry *_incr(opargs *args);

### 2022-09-04 src/interpreter/parser.c
change if behaviour, fix empty function

@@ -249,8 +249,9 @@ stnode *secondary(cache *cache, unsigned char rbp){
                     // loop variable
                     stnode *left, *last;
 
-                    // Set right pointer to NULL
+                    // Set right and left pointers to NULL
                     ret->data.parent.right = NULL;
+                    ret->data.parent.left = NULL;
 
                     // keep track of level
                     unsigned int level = 1;

### 2022-10-01 paper.tex
Continue writing paper

@@
 \usepackage{setspace}
+\usepackage{hyperref}
+\usepackage{dirtree}
 \usepackage{times} % Use Times New Roman font
@@
     \lstset{language=C,numbers=left,basicstyle=\scriptsize,
-        frame=single,caption=src/#1,keywordstyle=\color{blue},
+        frame=single,caption=#1,keywordstyle=\color{blue},
         stringstyle=\color{green},commentstyle=\color{gray},
         morecomment=[l][\color{magenta}]{\#}}
-    \lstinputlisting[firstnumber=#2,firstline=#2,lastline=#3]{src/#1}
+    \lstinputlisting[firstnumber=#2,firstline=#2,lastline=#3]{#1}
 }
@@
 
-\code{main.c}{2}{6}
-
 \section{Theory}
@@
 
+\subsection{Different types of interpreters}
+In computing, there are several types of interpreters that behave differently.
+Here are some examples and their use-cases:
+\begin{itemize}
+    \item \emph{Just in time Compiler.} JIT-compilers blur the line between
+        classical interpreters and compilers. As the name suggests, code is
+        being compiled at run time of the program. The JIT-compiler can optimize
+        compilation of more-often used code and therefore adapt itself to
+        behave somewhat close to optimally for a given use-case. Today, this 
+        technique is often used instead of classical interpreters, as
+        JIT-compilers come with only advantages. The most popular example of
+        a JIT-compiler would be Google's V8\footcite{V8} JavaScript engine used
+        in NodeJS and any Chromium browsers.
+    \item \emph{Bytecode Interpreter.} Such interpret the bytecode that was
+        output by a bytecode compiler from a given piece of code. The bytecode
+        is an intermediate representation of the program, used to speed up
+        both compilation and interpretation. Such an intermediary bytecode is
+        useful for ensuring platform independence or portability, as it will be
+        the same across any machine, the only difference being the interpreter.
+        The most popular example would be the Java Virtual Machine\footcite{jvm},
+        a bytecode interpreter for java bytecode, also used by many other languages
+        such as Kotlin, Scala or Groovy.
+    \item \emph{Abstract Syntax Tree interpreter.} Source code can be
+        transformed into an AST or parse tree by a parser. A simple example
+        can be found under 
+        \hyperref[simple interpreter]{'Implementation/The Art of Abstraction'}
+        on page \pageref{simple interpreter}. Compared to bytecode interpreters
+        there is a large time overhead when performing syntax related computation
+        or visiting tree nodes recursively.
+    \item \emph{Self interpreter.} These are interpreters that interpret the
+        language they themselves are written in. This requires a program
+        written in another language running an interpreter for the wanted
+        language, in which runs another interpreted for and written in said
+        language. Using a host language to initiate such system is called
+        bootstrapping.
+
+        Self-interpreters are closely related to self-hosting compilers. These
+        are in turn compilers which are written in the language they compile.
+        An example of a compiler written in itself is rustc, the Rust
+        compiler\footcite{rustc}, which was bootstrapped by the OCaml language.
+\end{itemize}
+This paper will implement a mixture form of an AST compiler and bytecode
+interpreter in-one.
+
 \section{Implementation}
@@
 features like built-in git integration.
-\emph{Git} is the project management and tracking software used to back-up and share
+\emph{Git} is the project management and tracking software used to back up and share
 the project files on \emph{GitHub.com}, a Microsoft hosted git repository server to
@@
 
+\subsection{The Art of Abstraction}
+When dealing with problems in general, a helpful method of going about can be
+to create an abstraction. This is also the case when building a programming
+language interpreter. Like that, an interpreter is often times subdivided into several
+components, each solving a discrete problem. One way to define
+such components could be the following:
+\begin{enumerate}\label{simple interpreter}
+    \item Create meaningful groups of characters from the input string. This
+        process is called tokenizing and is done with the component called
+        tokenizer or lexer. As an example, the string "aa+  bb" could result
+        in three tokens, namely "aa", "+" and "bb", in case space characters
+        are considered unimportant.
+    \item Ordering the tokens into a so-called parse-tree or syntax-tree.
+        Creating an order is important, as some calculations depend on others, e.g. multiplication
+        is always done before addition. TODO:example
+    \item Technically, the parse tree could now be interpreted directly, and
+        an output could be produced, as the interpreter know knows what to do
+        and in what order to perform computations.
+\end{enumerate}
+As seen above, these components, or stages as they will be called in this paper,
+each input and output data, where the output of one stage is the input of the
+respective next stage.
+
+\subsection{A Project Overview}
+
+\newpage\subsubsection{File Structure}
+Here is a file tree overview of the project which will be referenced
+several times in the following sections of this paper. In the depicted file tree
+only source files are shown, no build files or executables.
+\small{\setlength{\DTbaselineskip}{10pt}\dirtree{%
+    .1 project root.
+    .2 .gitignore.
+    .2 README.md.
+    .2 LICENSE.
+    .2 Makefile.
+    .2 mapop.gperf.
+    .2 paper.tex.
+    .2 bibliography.bib.
+    .2 pseudocode
+        .3 parser.c
+    .2 src.
+        .3 main.c.
+        .3 interpreter.
+            .4 bytecode.c.
+            .4 bytecode.h.
+            .4 error.
+                .5 error.c.
+                .5 error.h.
+            .4 interpreter.c.
+            .4 interpreter.h.
+            .4 libraries.
+                .5 libraries.c.
+                .5 libraries.h.
+                .5 stdlib.
+                    .6 stdlib.c.
+                    .6 stdlib.h.
+            .4 localizer.c.
+            .4 localizer.h.
+            .4 mappings.
+                .5 operations.c.
+                .5 operations.h.
+                .5 mapop.c.
+                .5 mapop.h.
+            .4 memory.
+                .5 hashtable.c.
+                .5 hashtable.h.
+                .5 array.c.
+                .5 array.h.
+                .5 stack.c.
+                .5 stack.h.
+            .4 parser.c.
+            .4 parser.h.
+            .4 processing.
+                .5 implementations.c.
+                .5 implementations.h.
+            .4 runtime.c.
+            .4 runtime.h.
+            .4 tokenizer.c.
+            .4 tokenizer.h.
+}}
+
+As visible in the file tree, most of the source-code files, ending in .c,
+have a corresponding header file ending in .h. This is due to the nature
+of the compilation process of the C language, which is the host language
+the interpreter of \name is written in.
+Each header file contains only information about functions, types or variables
+available to other source or header files. The actual implementations or values
+are stored in the respective .c files.
+TODO:explain C compilation or link example
+
+The pseudocode directory contains only examples used in this paper, no acutal,
+working code. The paper is written in \LaTeX and BibLaTeX and compiled with
+the tools pdflatex and biber. The paper's source files are paper.tex and
+bibliography.bib.
+
+The file mapop.gperf is a file to generate a perfect-hash-function\footcite{hash}
+with the GNU gperf tool. The generated mapop.c file is used to map string operators
+like '==' or '+' to a simpler number representation in constant time.
+
+\subsubsection{Interpreter Construction}
+This section will explain the process of constructing and the inner workings of
+the \name interpreter. It is therefore the most complicated yet detailed section
+in this paper. Examples will be made with both pseudo code and actual source
+code from the project to further illustrate the approaches made. File paths
+used are relative to the project's root directory.
+
+The syntax of \name is not specified in a formal programming language grammar,
+but rather implemented from scratch in the files src/interpreter/tokenizer.c 
+and src/interpreter/parser.c.
+
+Source code can be found on GitHub TODO: insert link
+
+\subsubsection{The Entry Point}
+The entry point of the project lies in the src/main.c file.
+The file implements the C main function, which is the entry-point of any
+C application in general. Since the file is so small, here is its source code:
+\code{src/main.c}{1}{999}
+
+For the readers who do not know how to read source code, C in particular, here is a short
+explanation of the basic syntax of the C language based on the src/main.c file:
+\begin{itemize}
+    \item \emph{Preprocessor Instructions.} Lines beginning with hashtags are
+        C preprocessor instructions, they are dealt with by the C preprocessor.
+        Preprocessing is what the compiler does before actually compiling the
+        code.
+
+        The most important preprocessor instructions are '\#include <header.h>'
+        to include various definitions from a header file and '\#define A B' to
+        define A as an alias for B.
+    \item \emph{Functions.} Functions in C are defined the following way:
+        '<return type> <function name>([arguments]) { <function body> }'.
+        In main.c there is one function. The function's name is 'main', its
+        return type is 'int', it has two arguments 'argc' and 'argv' and a
+        function body containing what the function does.
+    \item \emph{Pointers.} Pointers are variables that store memory addresses
+        of data. In C-like syntax they are defined with the star (*) symbol.
+        An example from src/main.c is 'FILE *input;'. That means that input is not
+        actually a FILE, but rather the memory address to where a FILE is.
+
+        To dereference the pointer, meaning to actually retrieve the value stored
+        at the memory address it contains, another star could be used as such:
+        'FILE value = *input;'.
+\end{itemize}
+
+For those who are further interested in the C programming language, there are
+many internet articles or tutorials going about reading and understanding or
+writing and compiling C code. One such example is the article 'The C Beginner's
+Handbook' provided by freeCodeCamp.org linked in the sources\footcite{freeCodeCamp}
+on page \pageref{bibliography}.
+
+Here is a summary of what happens in the src/main.c file: the src/interpreter/interpreter.h
+header file is included. In the main function, the program decides to either
+read a program from the standard input or from a file. Then, the interpret function
+from the said included header is called with the selected input.
+
+\subsubsection{The Interpreter}
+This section deals with the file src/interpreter/interpreter.c which guides
+input data through the various stages of the \name interpreter in its
+function 'interpret' which is called from the program's main function.
+This section does not go in detail on the following stages, see their own sections.
+
+The 'interpret' function calls the parser to generate a syntax-tree from tokens retrieved
+from the tokenizer implemented in src/interpreter/parser.c and 
+src/interpreter/tokenizer.c respectively.
+
+After that, variables with names are 'localized', meaning their names are
+replaced by addresses where their value is stored by src/interpreter/localizer.c.
+The advantage of performing this step so early is that mapping variable names
+to memory is a rather resource intensive task with time complexity of maximally 
+O(log n), n being the amount of different variables.
+
+The syntax tree could now be interpreted directly, but is instead converted
+to an intermediary internal bytecode format by src/interpreter/bytecode.c.
+This step is done so that code that is ran multiple times does not convey
+the problem of time overhead when traversing the syntax tree, but the current
+implementation does not take advantage of this.
+
+The final step is to process or execute said bytecode. This step is performed
+by src/interpreter/runtime.c.
+
+\subsubsection{The Tokenizer}
+The tokenizer is implemented in src/interpreter/tokenizer.c. Its job is to
+take source code in plain-text format as input and output so-called tokens.
+These tokens represent groups of characters that belong together. In this paper's
+implementation, each token has a strict type and a text content.
+The different types of tokens and other thins are defined in the header file
+src/interpreter/tokenizer.h.
+
+Here is an example of tokenization:
+An input of 'std.test==23.4' would result in the following tokens according
+to the tokenizer implementation of this paper:
+\begin{enumerate}
+    \item Token of type FIELD with content 'std.text'.
+    \item Token of type SYMBOL with content '=='.
+    \item Token of type NUMBER with content '23.4'.
+\end{enumerate}
+FIELD type tokens are later replaced by REFERENCES during the stage of localization
+and SYMBOL type tokens are mapped to EXPR by src/implementation/mappings/mapop.c
+during parsing. NUMBER type tokens are replaced by an actual number type, which
+is C's 'long double' type as specified in src/interpreter/bytecode.h.
+
+The tokenizer used in this paper is a simple finite-state machine, reading
+character by character and determining the token type as it goes.
+
+These tokens are then directly consumed by the parser as described in the next
+section.
+
+\subsubsection{The Parser}
+The parser, implemented in src/interpreter/parser.c, generates a syntax-tree
+from tokens. The algorithm used is inspired by the Pratt Parsing\footcite{pratt}
+algorithm first described by Vaughan Pratt in 1973, which is a kind of precedence
+parser based on recursive descent. There is a great simple explanation of the algorithm
+by Jonathan Apodaca on dev.io\footcite{devio} linked in the references on page 
+\pageref{bibliography} which this paper's parser takes inspiration from.
+
+The syntax tree in the implementation of \name consists of 'stnode' type
+pointers, as defined in src/interpreter/parser.h. Each such node depending
+on its 'nodetype', declared in src/interpreter/tokenizer.h,
+is either a value-node storing a pointer to some data, e.g. a number or text, or a
+parent-node referencing two child nodes and an operator, e.g. two numbers as
+children and a +-operator to signify an addition.
+
+The difficult part of implementing a parser is precedence. Presedence and
+associativity define the order of execution of statements. A simple example
+would be to perform multiplication before addition, e.g. "2+3*4" would
+result in 14 and is not the same as "(2+3)*4". Therefore, each operator is
+assigned a precedence or priority, as well as an associativity, which is
+either 'Left to Right' or 'Right to Left'. An example of associativity is
+'20 / 10 / 2'. Both in C as well as in \name, the division operator is
+left-to-right associative, meaning the stated expression is the same as
+'(20 / 10) / 2' and not as '20 / (10 / 2)'. Both precedence and associativity
+of each operator are defined in the file mapop.gperf.
+
+The parser implementation in src/interpreter/parser.c may be difficult to
+read and understand, which is why the most important functions are explained here:
+\begin{itemize}
+    \item \emph{advance.} This function reads a new token from the tokenizer.
+    \item \emph{secondary.} The 'secondary' function deals with parsing single,
+        simple values that make up the start of an expression, such as numbers,
+        fields, strings and it handles brackets, such as indexing with square
+        brackets, functions or code blocks with curly braces and sub-expressions
+        in standard brackets.
+    \item \emph{expr.} The 'expr' function calls 'secondary' which acts
+        as the basis for a new expression. It then repeatedly checks
+        if an operator follows the secondary token and creates a tree of
+        EXPR type stnodes, each having children which are recursively parsed
+        and information about the operator.
+    \item \emph{parse.} The parse function is a small wrapper which only
+        calls the 'expr' function, initiating the parsing process.
+\end{itemize}
+
+Following is a piece of pseudo-code to better illustrate the inner workings of the
+parser which is based on a Pratt-parser. 
+This is only an example and not completely accurate to the actual parser used
+by this project.
+TODO: validate code
+\code{pseudocode/parser.c}{1}{999}
+
 \section{Results}
+\subsection{Limitation}
+Token input size etc.
+Single threaded
+
 \section{Discussion}
@@
 Reflection\footcite{reflection}, also called reflective programming, describes
-the ability of a program to examine and modify its own structure and behaviour 
+the ability of a program to examine and modify its own structure and behavior 
 at runtime. An example would be self-modifying code. This can easily be
-achieved in assembly language, which inherently does not differenciate between
+achieved in assembly language, which inherently does not differentiate between
 data and instructions.
@@
 programs and program flows, effectively making smaller executables and more
-resource efficient programs, as those programs can adjust themselfs to run
+resource efficient programs, as those programs can adjust themselves to run
 differently under different conditions at runtime.
@@
 the entire source code of \name by myself using resources listed under the
-references section.
+'references' section on page \pageref{bibliography}.
 
@@
 
-\newpage\printbibliography[heading=bibintoc]
+\newpage\printbibliography[heading=bibintoc]\label{bibliography}
 
+TODO: check correctness of all file paths!! e.g. src/...
 \end{document}

### 2022-10-01 pseudocode/parser.c

@@ -1,0 +1,36 @@
+// implement functions
+stnode* parse() {
+    // just call expr
+    return expr(0);
+}
+
+stnode *expr(int precedence) {
+    // read secondary
+    stnode *left = secondary();
+    stnode *ret = left;
+
+    // reached end of file
+    if(left == NULL)
+        return NULL;
+
+    token next;
+    while(precedence_of(next = peek_token()) > precedence 
+            && next.type == SYMBOL) {
+        // allocate a new node
+        ret = allocate_stnode(EXPR);
+        ret->left_child = sec;
+        ret->right_child = expr(presedence_of(next) - associativity_of(next));
+        ret->operator = operator_of_token(next);
+
+        // throw away token
+        next = next_token();
+
+        sec = ret;
+    }
+    return ret;
+}
+
+stnode *secondary() {
+    // dealing with brackets left out, see full source code...
+    return stnode_from(next_token()); 
+}

### 2022-10-08 paper.tex
continue writing on paper

@@
 % the name of the programming language
-\newcommand{\name}{-the language- }
+\newcommand{\name}{\emph{TestScript }}
 
@@
         process is called tokenizing and is done with the component called
-        tokenizer or lexer. As an example, the string "aa+  bb" could result
-        in three tokens, namely "aa", "+" and "bb", in case space characters
+        tokenizer or lexer. As an example, the string 'aa+  bb' could result
+        in three tokens, namely 'aa', '+' and 'bb', in case space characters
         are considered unimportant.
@@
 
-Source code can be found on GitHub TODO: insert link
-
-\subsubsection{The Entry Point}
+\subsubsection{The Entry-Point}
 The entry point of the project lies in the src/main.c file.
@@
 
-Here is a summary of what happens in the src/main.c file: the src/interpreter/interpreter.h
+Here is a very quick summary of what happens in the src/main.c file: the src/interpreter/interpreter.h
 header file is included. In the main function, the program decides to either
@@
 The advantage of performing this step so early is that mapping variable names
-to memory is a rather resource intensive task with time complexity of maximally 
+to memory is a rather resource-intensive task with time complexity of maximally 
 O(log n), n being the amount of different variables.
@@
 implementation, each token has a strict type and a text content.
-The different types of tokens and other thins are defined in the header file
+The different types of tokens and other things are defined in the header file
 src/interpreter/tokenizer.h.
@@
 \end{enumerate}
-FIELD type tokens are later replaced by REFERENCES during the stage of localization
+FIELD type tokens are later replaced by REFERENCE tokens during the stage of localization
 and SYMBOL type tokens are mapped to EXPR by src/implementation/mappings/mapop.c
@@
 parent-node referencing two child nodes and an operator, e.g. two numbers as
-children and a +-operator to signify an addition.
+children and a plus-operator to signify an addition.
 
-The difficult part of implementing a parser is precedence. Presedence and
+The difficult part of implementing a parser is precedence. Precedence and
 associativity define the order of execution of statements. A simple example
-would be to perform multiplication before addition, e.g. "2+3*4" would
-result in 14 and is not the same as "(2+3)*4". Therefore, each operator is
-assigned a precedence or priority, as well as an associativity, which is
+would be to perform multiplication before addition, e.g. '2+3*4' would
+result in 14 and is not the same as '(2+3)*4'. Therefore, each operator is
+assigned a precedence, or priority, as well as an associativity, which is
 either 'Left to Right' or 'Right to Left'. An example of associativity is
-'20 / 10 / 2'. Both in C as well as in \name, the division operator is
+\emph{'20 / 10 / 2'}. Both in C as well as in \name, the division operator is
 left-to-right associative, meaning the stated expression is the same as
-'(20 / 10) / 2' and not as '20 / (10 / 2)'. Both precedence and associativity
-of each operator are defined in the file mapop.gperf.
+\emph{'(20 / 10) / 2'} and not as \emph{'20 / (10 / 2)'}. Both precedence 
+and associativity are properties that are defined in the file mapop.gpref for
+each operator.
 
@@
 
-Following is a piece of pseudo-code to better illustrate the inner workings of the
+Following is a piece of pseudocode to better illustrate the inner workings of the
 parser which is based on a Pratt-parser. 
@@
 
+\subsubsection{The Bytecode Compiler}
+The bytecode compiler is part of the interpreter. Its job is to create a list
+of byte-sized instructions and data from the syntax-tree the parser created.
+Having such an intermediary bytecode format, instead of directly interpreting
+the parse-tree can be beneficial. This is the case when the same piece of code
+has to be run several times. Instead of generating a time and memory overhead 
+having to walk along tree-nodes recursively, only the list of instructions in
+the bytecode can be traversed, which is much faster.
+
+Not to be confused with machine-native bytecode, the bytecode compiler in \name
+generates byte-sequences in an internal format, that are then interpreted. A
+JIT compiler would instead compile the AST down to machine-native bytecode, which
+could run directly on the system's hardware and would not have to be interpreted.
+
+To generate such a sequence of bytes, \name recursively reads through the
+binary tree created by the parser and pushes found instructions onto a stack,
+which is in this case used similarly to a variable-sized list.
+The algorithm used is inspired by Edsger Dijkstra's 'Shunting yard 
+algorithm'\footcite{shunting_yard}, traversing the tree post-order to create
+a prefix notation.
+
+A prefix notation is a way to write down mathematical expressions, except
+operators precede their operands, in contrast to the more common infix notation,
+where operands surround the operator.
+
+In src/interpreter/bytecode.c we can see, that first, the left operand and then
+the right operand are added onto the stack of bytecode. Therefore, when
+popping from the stack and evaluating instructions, the operands will have to
+be read in reverse.
+Here is an example:
+\begin{enumerate}
+    \item Assume the following mathematical expression in infix notation:
+        \begin{center}
+            \emph{'1 - 2 * 4 / (5 + 6)'}
+        \end{center}
+    \item In common prefix notation, also called Polish notation, the
+        operators would now precede the operands as such:
+        \begin{center}
+            \emph{'- 1 (/ (* 2 4) (+ 5 6))'}
+        \end{center}
+    \item But in \name the operands are pushed onto the stack is reverse order.
+        This would look like the following:
+        \begin{center}
+            \emph{'- (/ (+ 6 5) (* 4 2)) 1'}
+        \end{center}
+    \item The brackets are not needed, because for every operator, the amount
+        of operands is known. This is the case as e.g. the subtraction operator
+        and the negative operator, which are both a minus sign, are distinguished
+        beforehand. This means the above statement would look like this:
+        \begin{center}
+            \emph{'- / + 6 5 * 4 2 1'}
+        \end{center} 
+        This is an ideal abstraction of how data 
+        is represented internally in \name.
+\end{enumerate}
+
+\subsubsection{The Runtime}
+This is the last stage of interpreting program code as implemented in \name.
+The runtime steps through the intermediary bytecode created by the bytecode
+compiler and evaluates expressions by plugging in the provided values.
+All available expressions are implemented in 
+src/interpreter/processing/implementations.c.
+
+The evaluation of expressions also works in a recursive manner. Here is an 
+example of \name's bytecode notation and how it would be evaluated at runtime:
+\begin{itemize}
+    \item Assume the input in \name's bytecode to resemble the following expression: 
+        \begin{center}
+            \emph{'- 6 * 3 4'}
+        \end{center}
+        In infix notation, this would be equivalent to 
+        \begin{center}
+            \emph{'(4 * 3) - 6'}.
+        \end{center}
+    \item The runtime will read through the bytecode left-to-right. First it will
+        notice the subtraction-operator and accordingly assume that two
+        operands must follow.
+    \item The processing function will call itsself
+        and notice the 6, a number, and therefore a value. The function simply returns the 6
+        to the subtraction expression. This 6 will be the right-hand operator
+        of the subtraction.
+    \item The subtraction will then call the processing function again to find
+        the left operand. As the processing function reads a multiplication operator,
+        it now recursively reads through that expression's right and then left
+        operator, which would be 3 and for respectively and compute and return
+        the result of the multiplication, which is 12.
+    \item Now the subtraction knows its right- and left-hand operators, 6 and 12,
+        and again, calculate and return the result, which is 6.
+\end{itemize}
+
+That's it! With all the possible expressions available, entire programs
+can now be interpreted without a problem.
+
 \section{Results}
 \subsection{Limitation}
-Token input size etc.
-Single threaded
+\name is a programming language without a doubt. But it is not perfect in any
+way. There are limitations that make it practically unusable for actual
+professional projects. In this section, a few shortcomings of \name are listed
+and reasoned over.
+
+\subsubsection{Memory Leaks}
+Memory leaks happen, when a program reserves memory to store some data, but
+never frees up the space for use for other programs, even though the data
+has long been discarded and is of no further use.
+In \name, memory leaks are a major problem. Even though barely noticable for small
+programs it interprets, \name does not have proper setups in place to free up
+all used memory that is not used any longer.
+
+Other programming languages, notably object-oriented languages such as Java,
+use Garbage-Collection, which automatically tracks unused allocated memory and
+frees it up automaticadlly. While this is also the case with \name, there are
+many internal flaws that have nothing to do with variables used in a \name
+program, but rather internally.
+
+A major reason for the memory leaks, but not an excuse, is the nature of the
+C programming language the \name interpreter was written in. The C language and
+compiler do not keep track of states in which certain blocks of memory are
+forgotten about and warn the developer, instead, memory leaks happen.
+More modern C alternatives such as Rust have a ownership-system in place,
+in which any allocated memory has a function as an owner, and when the function
+is done processing, the memory is automatically freed. Such systems are a
+helpful tool for developers to create better programs with low mental overhead,
+and are what programming language design should be all about.
+
+\subsubsection{Single-Threaded}
+The \name interpreter does not implement any multi-threaded, meaning parallel
+running, programs as most other languages do. Instead, any \name program can
+only be interpreted and run sequentially. Since most modern computer processors
+are equipped with a handful of processing cores and threads, not being able to
+make use of those is a major downside and in itsself a reason not to use this
+language.
+
+\subsubsection{Token input size}
+In the current iteration, each token's content's size is limited to 256 bytes,
+which makes it impossible to use e.g. variable names longer thatn 256 ASCII
+characters. This would probably be the easyest thing to fix of here-listed 
+problems of \name, but since 256 characters are long enough for 
+illustration purposes, fixing this is not worth spending time on.
+
+\subsubsection{Lack of a Development Environment}
+This is certainly the greatest problem of \name, aside the countless smaller
+bothering aspects.
+
+Pretty much any good, modern programming language does not only consist
+of it's specification and a compiler or interpreter, but of many more
+tools surrounding it.
+Such tools often include a package-manager, with which people around the world
+can share code. Examples would be npm\footcite{npm} or cargo\footcite{cargo}, 
+the 'Node Package Manager' for JavaScript applications and and Rust's package 
+manager respectively.
 
@@
 out. This section of the paper gives insight on interesting ideas or concepts
-that are used in other programming languages, not \name.
+that are great to have and are used in other programming languages, but not 
+could not be implemented into \name.
 

### 2022-10-10 paper.tex
add source code to paper appendix

@@
 \usepackage{dirtree}
+\usepackage{catchfile}
 \usepackage{times} % Use Times New Roman font
@@
 
-% Defining custom commands
+% Read contents of files.txt into \files
+\CatchFileDef{\files}{files.txt}{\endlinechar=-1}
 
-% include C code from file #1 from lines #2 to #3
+% Include C code from file #1 from lines #2 to #3
 \newcommand{\code}[3] {
     \lstset{language=C,numbers=left,basicstyle=\scriptsize,
-        frame=single,caption=#1,keywordstyle=\color{blue},
+        frame=single,title=#1,keywordstyle=\color{blue},stepnumber=5,
         stringstyle=\color{green},commentstyle=\color{gray},
@@
 }
+
+% Center text on newline
+\newcommand{\expr}[1] {
+    \begin{center}
+        #1
+    \end{center}
+}
+
+\newcommand{\pagelabel}[1]{\phantomsection\label{#1}}
 
@@
 such components could be the following:
-\begin{enumerate}\label{simple interpreter}
+\begin{enumerate}\pagelabel{simple interpreter}
     \item Create meaningful groups of characters from the input string. This
@@
         Creating an order is important, as some calculations depend on others, e.g. multiplication
-        is always done before addition. TODO:example
+        is always done before addition. % TODO:example
     \item Technically, the parse tree could now be interpreted directly, and
@@
 only source files are shown, no build files or executables.
-\small{\setlength{\DTbaselineskip}{10pt}\dirtree{%
+\begin{flushleft}
+    \small\setlength{\DTbaselineskip}{10pt}
+    \dirtree{%
         .1 project root.
@@
         .2 bibliography.bib.
-    .2 pseudocode
-        .3 parser.c
+        .2 pseudocode.
+            .3 src/interpreter/parser.c.
         .2 src.
@@
                 .4 tokenizer.h.
-}}
+    }
+\end{flushleft}
 
@@
 are stored in the respective .c files.
-TODO:explain C compilation or link example
+% TODO:explain C compilation or link example
 
@@
 This section will explain the process of constructing and the inner workings of
-the \name interpreter. It is therefore the most complicated yet detailed section
+the \name interpreter. It is therefore the most complicated and detailed section
 in this paper. Examples will be made with both pseudo code and actual source
-code from the project to further illustrate the approaches made. File paths
+code from the project to further illustrate the decisions made during
+interpreter design. File paths
 used are relative to the project's root directory.
@@
 
-        The most important preprocessor instructions are '\#include <header.h>'
-        to include various definitions from a header file and '\#define A B' to
-        define A as an alias for B.
+        The most common preprocessor instructions are the following.
+        On one hand there is \expr{\#include <header.h>}
+        to include various definitions from a header file, i.e. to paste
+        its content at where the 'include' statement lies.
+        On the other hand exists \expr{\#define A B} used to
+        define A as an alias for B and replace every following occurrence of A
+        with B.
+
     \item \emph{Functions.} Functions in C are defined the following way:
-        '<return type> <function name>([arguments]) { <function body> }'.
+        \expr{<return type> <function name>([arguments]) { <function body> }}
+        or with a concrete, working example:
+        \expr{int myFunction(int *arg) \{ printf("arg: \%p, arg); \}.}
         In main.c there is one function. The function's name is 'main', its
@@
         function body containing what the function does.
+
     \item \emph{Pointers.} Pointers are variables that store memory addresses
@@
 
-        To dereference the pointer, meaning to actually retrieve the value stored
+        To dereference the pointer, i.e. to actually retrieve the value stored
         at the memory address it contains, another star could be used as such:
@@
 For those who are further interested in the C programming language, there are
-many internet articles or tutorials going about reading and understanding or
+many internet articles or tutorials about reading and understanding or
 writing and compiling C code. One such example is the article 'The C Beginner's
@@
 by this project.
-TODO: validate code
+% TODO: validate code
 \code{pseudocode/parser.c}{1}{999}
@@
     \item Assume the following mathematical expression in infix notation:
-        \begin{center}
-            \emph{'1 - 2 * 4 / (5 + 6)'}
-        \end{center}
+        \expr{\emph{'1 - 2 * 4 / (5 + 6)'}}
     \item In common prefix notation, also called Polish notation, the
         operators would now precede the operands as such:
-        \begin{center}
-            \emph{'- 1 (/ (* 2 4) (+ 5 6))'}
-        \end{center}
+        \expr{\emph{'- 1 (/ (* 2 4) (+ 5 6))'}}
     \item But in \name the operands are pushed onto the stack is reverse order.
         This would look like the following:
-        \begin{center}
-            \emph{'- (/ (+ 6 5) (* 4 2)) 1'}
-        \end{center}
+        \expr{\emph{'- (/ (+ 6 5) (* 4 2)) 1'}}
     \item The brackets are not needed, because for every operator, the amount
@@
         beforehand. This means the above statement would look like this:
-        \begin{center}
-            \emph{'- / + 6 5 * 4 2 1'}
-        \end{center} 
+        \expr{\emph{'- / + 6 5 * 4 2 1'}}
         This is an ideal abstraction of how data 
@@
     \item Assume the input in \name's bytecode to resemble the following expression: 
-        \begin{center}
-            \emph{'- 6 * 3 4'}
-        \end{center}
+        \expr{\emph{'- 6 * 3 4'}}
         In infix notation, this would be equivalent to 
-        \begin{center}
-            \emph{'(4 * 3) - 6'}.
-        \end{center}
+        \expr{\emph{'(4 * 3) - 6'}.}
     \item The runtime will read through the bytecode left-to-right. First it will
@@
 
-\section{Appendix}
+\newpage\printbibliography[heading=bibintoc]\pagelabel{bibliography}
 
-\newpage\printbibliography[heading=bibintoc]\label{bibliography}
+% TODO: check correctness of all file paths!! e.g. src/...
 
-TODO: check correctness of all file paths!! e.g. src/...
+\newpage\section{Appendix}
+The appendix contains the complete source code of \name.
+
+% Append all soruce files
+\foreach \f in \files{\pagelabel{\f}\code{\f}{1}{9999}}
+
 \end{document}

### 2022-10-10 src/interpreter/bytecode.c

@@ -8,7 +8,8 @@
 #define DOUBLE(string) (number) strtod(string, NULL)
 
 static inline mementry *_register(bytecode *code, stackptr ptr, typing type) {
-    mementry *ret = *(mementry **)push(code, ptr, sizeof(mementry *)) = malloc(sizeof(mementry));
+    mementry *ret = 
+        *(mementry **)push(code, ptr, sizeof(mementry *)) = malloc(sizeof(mementry));
     ret->type = type;
     return ret;
 }
@@ -74,7 +75,8 @@ void _recursiveconsume(bytecode *dst, stackptr ptr, stnode *subtree) {
                 _register(dst, ptr, UNDEFINED);
 
             // then push operation onto the stack
-            _register(dst, ptr, EXPR)->value = implementationof(subtree->data.parent.op->opcode);
+            _register(dst, ptr, EXPR)->value = 
+                implementationof(subtree->data.parent.op->opcode);
             break;
 
         case MEMBER: {

### 2022-10-10 src/interpreter/interpreter.c

@@ -81,7 +81,8 @@ void interpret(FILE *stream, char* filename) {
     printf("Reached EOF\n");
     // Print unmatched brackets
     if(cac.bracketstack_offset != 0) {
-        printf("\nNot all brackets matched, %u left!\n", (unsigned int) cac.bracketstack_offset);
+        printf("\nNot all brackets matched, %u left!\n", 
+                (unsigned int) cac.bracketstack_offset);
         printf("Assuming: ");
         while(cac.bracketstack_offset != 0)
             putchar(*(char *)pop(cac.bracketstack, &cac.bracketstack_offset, 1));

### 2022-10-10 src/interpreter/mappings/mapop.c

@@ -25,8 +25,7 @@
       && ('s' == 115) && ('t' == 116) && ('u' == 117) && ('v' == 118) \
       && ('w' == 119) && ('x' == 120) && ('y' == 121) && ('z' == 122) \
       && ('{' == 123) && ('|' == 124) && ('}' == 125) && ('~' == 126))
-/* The character set is not based on ISO-646.  */
-#error "gperf generated tables don't work with this execution character set. Please report a bug to <bug-gperf@gnu.org>."
+#error "gperf generated tables don't work with this execution character set."
 #endif
 
 #line 8 "./mapop.gperf"

### 2022-10-10 src/interpreter/memory/hashtable.c

@@ -103,7 +103,8 @@ void ht_down(hashtable *table) {
     // decrease level
     table->level--;
 
-    while((stackptr = pop(s, &table->offset, sizeof(tableentry **))) != NULL && (entryptr = *stackptr) != NULL) {
+    while((stackptr = pop(s, &table->offset, 
+                    sizeof(tableentry **))) != NULL && (entryptr = *stackptr) != NULL) {
         // copy the entry
         entry = *entryptr; // never null
 
@@ -199,13 +200,15 @@ static inline tableentry *_find(hashtable *table, char *key) {
             bottom_value = value->hash;
 
             // the new range is now ]index, top[
-            index = (top - index -1) * (hash - value->hash) / (top_value - value->hash) + index +1;
+            index = (top - index -1) * 
+                (hash - value->hash) / (top_value - value->hash) + index +1;
         } else if(value->hash > hash) {
             top = index;
             top_value = value->hash;
 
             // the new range is now [bottom, index[
-            index = (index - bottom) * (hash - bottom_value) / (value->hash - bottom_value) + bottom;
+            index = (index - bottom) *
+                (hash - bottom_value) / (value->hash - bottom_value) + bottom;
         } else {
 
         #else
@@ -236,7 +239,8 @@ static inline tableentry *_find(hashtable *table, char *key) {
                     value->alternative = new_value = malloc(sizeof(tableentry));
 
                     // register into stack
-                    *(tableentry ***)push(table->stack, &table->offset, sizeof(tableentry **)) = &value->alternative;
+                    *(tableentry ***)push(table->stack, &table->offset,
+                            sizeof(tableentry **)) = &value->alternative;
 
                     // initialize entry
                     goto init_entry;
@@ -254,9 +258,11 @@ static inline tableentry *_find(hashtable *table, char *key) {
                 *source = malloc(sizeof(tableentry));
 
                 // put it onto the stack
-                // since the entry to be shadowed already exists, it will be removed from stack after this,
+                // since the entry to be shadowed already exists, 
+                // it will be removed from stack after this
                 // so ht_down is actually safe
-                *(tableentry ***)push(table->stack, &table->offset, sizeof(tableentry **)) = source;
+                *(tableentry ***)push(table->stack, &table->offset,
+                        sizeof(tableentry **)) = source;
 
                 // set the alternative
[…]

### 2022-10-10 src/interpreter/memory/hashtable.h

@@ -42,7 +42,8 @@ typedef struct {
         // set if this mementry is a mutable field (mutable entry in hashtable)
         unsigned char mutable: 1;
 
-        // > used by implementations.c < to see if a mementry only exists during expression-evaluation
+        // > used by implementations.c < to see if
+        // a mementry only exists during expression-evaluation
         unsigned char synthetic: 1; 
         // used to check if the value this mementry is pointing to is synthetic
         // this will only be set, if synthetic is set as well

### 2022-10-10 src/interpreter/parser.c

@@ -60,7 +60,8 @@ static inline void advance(cache *cache) {
             default: {
                 char *popped;
 
-                if((popped = (char *)pop(cache->bracketstack, &cache->bracketstack_offset, 1)) == NULL) {
+                if((popped = (char *)pop(cache->bracketstack,
+                                &cache->bracketstack_offset, 1)) == NULL) {
                     // Throw error, no closing bracket expected
                     throw(EP_INVALID_CLOSING_BRACKET, &cache->info);
 

### 2022-10-10 src/interpreter/processing/implementations.c

@@ -158,8 +158,8 @@ mementry *_bnot(opargs *args) {
 #define a (*(number *)lo->value)
 #define b (*(number *)ro->value)
 #define CALCULATE(calculation)\
-    mementry *ro = _recursiveprocess(args, DEREFERENCE); /* Load right operand into tmp mementry */\
-    mementry *lo = _recursiveprocess(args, DEREFERENCE); /* Load left operand into dst mementry */\
+    mementry *ro = _recursiveprocess(args, DEREFERENCE); /* Load right into tmp mementry */\
+    mementry *lo = _recursiveprocess(args, DEREFERENCE); /* Load left into dst mementry */\
     \
     /* make sure only to calculate numbers */\
     if(ro->type != NUMBER || lo->type != NUMBER) {\
@@ -177,7 +177,8 @@ mementry *_bnot(opargs *args) {
         dst = malloc(sizeof(mementry));\
         dst->value = malloc(sizeof(number));\
         dst->type = NUMBER;\
-        dst->flags = (struct s_mementry_flags) {.mutable=0, .synthetic=1, .value_synthetic=1};\
+        dst->flags = \
+        (struct s_mementry_flags) {.mutable=0, .synthetic=1, .value_synthetic=1};\
     }\
     \
     number *valdst = dst->flags.value_synthetic ? dst->value : malloc(sizeof(number));\
@@ -216,8 +217,8 @@ mementry *_grt(opargs *args){ CALCULATE(a > b) }
 mementry *_geq(opargs *args){ CALCULATE(a >= b) }
 
 mementry *_equ(opargs *args){
-    mementry *ro = _recursiveprocess(args, DEREFERENCE); /* Load right operand into tmp mementry */
-    mementry *lo = _recursiveprocess(args, DEREFERENCE); /* Load left operand into dst mementry */
+    mementry *ro = _recursiveprocess(args, DEREFERENCE); /* Load right into tmp mementry */
+    mementry *lo = _recursiveprocess(args, DEREFERENCE); /* Load left into dst mementry */
     
     /* determine the dst entry (or allocate a new one if ro and lo are not synthetics) */
     mementry *dst = NULL;
@@ -229,7 +230,8 @@ mementry *_equ(opargs *args){
         dst = malloc(sizeof(mementry));
         dst->value = malloc(sizeof(number));
         dst->type = NUMBER;
-        dst->flags = (struct s_mementry_flags) {.mutable=0, .synthetic=1, .value_synthetic=1};
+        dst->flags = 
[…]

### 2022-10-10 src/interpreter/processing/implementations.h

@@ -4,7 +4,8 @@
 
 // Define a function to process the output of some code into destination mementry
 // to be used by operator implementations and runtime.c
-// @param return_reference if set, a reference from hashtable is returned (and element is allocated)
+// @param return_reference if set,
+// a reference from hashtable is returned (and element is allocated)
 //  else, a copy is returned
 mementry *_recursiveprocess(opargs *args, char return_reference);
 void _free_synth(mementry *entry);

### 2022-10-11 paper.tex
add examples; continue writing

@@
 \section{Results}
+
+% TODO: add examples in /examples/
+\name is built upon a well structured implementation of a recursive Pratt
+parser and a basic tokenizer. Most actual features are reliant upon this
+foundation, such as operators and the standard library built into \name.
+
+\subsection{Syntax}
+The syntax of \name can be compared to the syntax of JavaScript, except
+there are no keywords at all, and everything is controlled via operators,
+which are non-alphanumeric characters. Such operators include the plus-sign 
+for addition, or the equals-sign for variable assignment.
+
+Following is a piece of example \name code, printing out the first ten numbers
+of the fibonacci sequence, as well as a disection of how that same code is interpreted.
+\code{examples/fibonacci.nts}{1}{999}
+\begin{itemize}
+\item The source code begins by declaring two variables \emph{a} and \emph{b}, which
+    will be utilized to keep track of the data used to calculate fibonacci numbers.
+    The memory allocation is done by creating a tuple of \emph{a} and \emph{b}, as suggested by 
+    the comma seperating the two. Since both variables \emph{a} and \emph{b} are not yet
+    known to the interpreter, they are automatically created and assigned a value
+    which has no content and the \emph{undefined} typing.
+\item The two variables \emph{a} and \emph{b} are then reassigned new values on line
+    2 of the program. In this case it is numbers with values zero and one respectively.
+    In this example, the two expressions are seperated by a space, but they can
+    also be separated by newline-characters or semicolons.
+\item Another variable \emph{count} is declared and defined as number in one step on line 4.
+    This variable will be used to limit the amount of numbers the program outputs.
+\item Yet another variable, \emph{f}, is declared and assigned a function as its
+    value on line 6. In \name, normal functions do not have any arguments, 
+    but they can interact with variables that were created outside their scope.
+\item Lines 7 to 10 hold mathematical expressions. At first, a is assigned its current value
+    plus the current value of b. Then in line 8, b is assigned the now new value of a minus the
+    current value of b, which corresponds to the value of a before it was just reassigned.
+    Lastly the variable \emph{count} is decremented by one.
+\item Now come the interesting parts of this program.
+    On lines 12 and 13, the function \emph{std.print} is referenced. This function is implemented
+    in plain C code, which internally is not handled the same as regular functions in \name,
+    and is wrapper to C's function \emph{printf}. The \emph{std.print} function
+    takes any \name variable or value and prints it to standard output. On line 13, the "\n"
+    is replaced by a newline-character, as \name can interpret Java-like escape-sequences.
+\item Another standard-library function is \emph{std.if}. It takes two or three arguments.
+    If the first argument is true, then the second, else the the third argument or 
+    \emph{undefined} is returned. A value is true, only if it is a number not equal to zero
+    or a non-empty string. In the case of this example, \emph{std.if} returns \emph{f} if 
+    count is greater than zero, and an empty function if count reaches zero. The two
+    brackets following the call of the \emph{std.if} function signal yet another call to the
+    returned value.
+\item The final thing left to do is to get the ball roling by calling the function \emph{f} for
+    the first time. As stated above, it will now print numbers and recursively call itsself,
+    until the variable \emph{count} reaches a value of zero.
+\end{itemize}
+
+\subsection{Security}
+Here is another example, which in contrast to the example above, focuses on a
+security concept implemented in \name.
+\code{examples/exec.nts}{1}{999}
+The code begins by creating the variable \emph{var}, which holds a string, and
+\emph{f}, which calls the function \emph{std.exec} with var as argument.
+Line number six holds a comment, indicated by the two slashes at the start of the line.
+Comments are ignored by the \name interpreter and their purpose is to contain notes
+and explanations inside the source code itself. Another type of commends are block-comments
+which span across multiple lines, such as on lines 9 to 12.
+On line seven, the \emph{std.exec} function is called for the first time. It runs the
+command it is given as argument in the user's shell, which is most commonly \emph{cmd} on
+Microsoft Windows and \emph{bash} on most Linux distributions. In this example, \emph{std.exec}
+launches the Firefox browser, but it is capable of all kinds of things that can otherwise be
+done in the shell, such as delete files or shut down the computer.
+
+That is why in most cases, the \emph{std.exec} function can be a risk waiting to be exploited
+by malicious \name programs. To solve this problem, there is the hashtag-operator, whose purpose
+it is to delete access to a variable through the specified name. It is used on line 13 of the
+example to remove access to the \emph{std.exec} function, which is why the call to that function
+on line 16, trying to shut down the computer, will result in an error. Unexpectedly though, the
+call on line 20 to the function \emph{f} which was created as an as to \emph{std.exec} will work.
+
+This security concept enforces the use of a top-down programming model. First, safe-to-use aliases
+to dangerous functions have to be created. Then, access to all those exploitable functions is
+revoked. This way, running unknown code can only access data and functions that are being provided.
+This contrasts the models of languages like Python. In Python, each seperate source file can
+make use of the \emph{import} keyword to include other files. Files imported this way do not
+have access or knowledge about the variables that exist in the file that includes them.
+In \name, each imported file would instead have access to all variables and functions that
+are made available to it, instead of having to include or import them seperately.
+This leads to another problem, but that is what the next section will be all about.
+
+\subsection{Name Shadowing}
+% TODO
+
+\section{Discussion}
+
+\subsection{Distribution}
+% TODO
+
 \subsection{Limitation}
@@
 and reasoned over.
+
+\subsubsection{Numbers}
+% TODO
 
@@
 manager respectively.
-
-\section{Discussion}
 

### 2022-10-19 paper.tex
continue paper

@@
-\documentclass[12pt,a4paper,man]{apa7}
-\usepackage{listings}
-\usepackage{setspace}
-\usepackage{hyperref}
-\usepackage{dirtree}
-\usepackage{catchfile}
+\documentclass[12pt,a4paper]{report}
+\usepackage{listings,xcolor,setspace,hyperref,dirtree}
+\usepackage{catchfile,pgffor}
 \usepackage{times} % Use Times New Roman font
-\usepackage[style=apa]{biblatex}
+\usepackage[style=alphabetic,backend=biber]{biblatex}
 
@@
 \title{The Creation of a Programming Language}
-\authorsaffiliations{Gymnasium Burgdorf}
-\shorttitle{Creation of a Programming Language}
-\abstract{
-    This is an abstract...
-}
 
@@
 
+% Cite text
+\newcommand{\cte}[1] {
+    \cite{#1}
+}
+
 % Center text on newline
@@
 
+% Set line spacing
+\onehalfspacing
+
 \begin{document}
 
-\onehalfspacing
 \maketitle\tableofcontents\newpage
+
+\abstract{
+    This is an abstract...
+}
 
@@
 \subsection{What this is and what this is not}
-The final product in the process of creating this programming language will 
-not be perfect, or in fact, be used or adopted by anyone at all, and it is not
-intended to be. Creating a good, or even usable, programming language 
-implementation by today's  standards means to create an entire ecosystem of 
-tools, not only including a compiler or interpreter, but also a package manager
-for dealing with dependencies and a platform for people to share their work.
+The programming language and the tools that surround it are not meant to be
+viable options for real-world projects. Instead, the final product, \name,
+should be a programming language interpreter which includes techniques and
+algorithms already present in other, well known interpreters.
+The goal of this paper is to further understand the inner workings of
+interpreters, transforming a human-readable programming language
+into a computer-executable program.
+
+Modern programming languages often times include additional tools, aside from
+the interpreter or compiler. This includes dependency management tools to
+easily create projects exdending upon existing ones, or a web-plattform for
+people to share their code.
 
@@
         JIT-compilers come with only advantages. The most popular example of
-        a JIT-compiler would be Google's V8\footcite{V8} JavaScript engine used
+        a JIT-compiler would be Google's V8\cte{V8} JavaScript engine used
         in NodeJS and any Chromium browsers.
@@
         the same across any machine, the only difference being the interpreter.
-        The most popular example would be the Java Virtual Machine\footcite{jvm},
+        The most popular example would be the Java Virtual Machine\cte{jvm},
         a bytecode interpreter for java bytecode, also used by many other languages
@@
         An example of a compiler written in itself is rustc, the Rust
-        compiler\footcite{rustc}, which was bootstrapped by the OCaml language.
+        compiler\cte{rustc}, which was bootstrapped by the OCaml language.
 \end{itemize}
 This paper will implement a mixture form of an AST compiler and bytecode
-interpreter in-one.
+interpreter. This is not practical in terms of improving \name, but serves
+the purpose of combining different features of different interpreters into one
+and therefore extending the scope of the project.
 
@@
 \subsection{Used Software}
-At the heart of creating a program in C stands the C compiler. This project uses
+At the heart of making any project in the C programming language
+stands the C compiler. This project uses
 \emph{gcc}, the GNU Compiler Collection, which, as the name suggests, also 
-supports different languages than C. Substantiating the development process is 
+supports more languages than solely C. Substantiating the development process is 
 \emph{GNU Make}, a general-purpose build system. It handles tasks like building
-the executable step-by-step from the project source code and directory clean-up.
-The C ecosystem is very easy to use and fundamentally supported by 
+the final executable program step-by-step from the project source code and directory clean-up.
+The C ecosystem is very easy to use on and fundamentally supported by 
 \emph{GNU+Linux}, the operating system used to program on.
-The main code editor used was \emph{Microsoft Visual Studio Code}, providing great
-features like built-in git integration.
+The code editor used was \emph{Microsoft Visual Studio Code}, providing great
+help with built-in git integration and auto-completion features.
 \emph{Git} is the project management and tracking software used to back up and share
-the project files on \emph{GitHub.com}, a Microsoft hosted git repository server to
-store and collaborate on source code.
+the project files on \emph{GitHub.com}, a Microsoft hosted git repository and
+web front-end to store and collaborate on source code.
 
 \subsection{The Art of Abstraction}
-When dealing with problems in general, a helpful method of going about can be
-to create an abstraction. This is also the case when building a programming
-language interpreter. Like that, an interpreter is often times subdivided into several
-components, each solving a discrete problem. One way to define
-such components could be the following:
+When dealing with problems in general, it is helpful to map out and think about
+the way of approaching the problem first. A common approach is to distill the
+problem situation down to a simpler one and extending the solution to take into
+account all important aspects later.
+
+The problem to solve in this paper is concisely to make a computer understand
+human-written and -readable text, called source code, and follow the instructions
+it proposes.
+Most interpreters follow a plain scheme to go about this problem, but the
+actual implementation of each of the steps can be heavily customized.
+Here is the problem, broken down into three steps as commonly done in other
+popular programming languages:
 \begin{enumerate}\pagelabel{simple interpreter}
-    \item Create meaningful groups of characters from the input string. This
-        process is called tokenizing and is done with the component called
-        tokenizer or lexer. As an example, the string 'aa+  bb' could result
-        in three tokens, namely 'aa', '+' and 'bb', in case space characters
-        are considered unimportant.
-    \item Ordering the tokens into a so-called parse-tree or syntax-tree.
+    \item To create meaningful groups of characters from the input string. This
+        process is called 'tokenization' and is done with the component called
+        tokenizer or lexer. As an example, the string 'aa + bb' could result
+        in three tokens, namely 'aa', '+' and 'bb'. In this case, space characters
+        are ignored. 
+        This is only an example and the grouping of characters could be done 
+        in any thinkable, logical way.
+    \item To order the tokens into parse-tree, also called syntax-tree, as
+        specified by rules. These rules are called a grammar.
         Creating an order is important, as some calculations depend on others, e.g. multiplication
-        is always done before addition. % TODO:example
-    \item Technically, the parse tree could now be interpreted directly, and
-        an output could be produced, as the interpreter know knows what to do
-        and in what order to perform computations.
+        is always done before addition. An IT data structure that can accommodate
+        such requirements is the binary tree. A binary tree consists of tree-nodes
+        that each have a maximum of two other tree nodes as their children, and
+        a tree node whose child they are, the parent node.
+    \item To interpret the parse tree, traversing through it and following the
+        instructions each of the tree-nodes represent. The outcome
+        will be the result of the computation the program describes.
 \end{enumerate}
+
 As seen above, these components, or stages as they will be called in this paper,
@@
 
-\newpage\subsubsection{File Structure}
+\subsubsection{File Structure}
 Here is a file tree overview of the project which will be referenced
 several times in the following sections of this paper. In the depicted file tree
-only source files are shown, no build files or executables.
+only source files are shown. No build files, executable files or files that
+are of no further importance are depicted.
 \begin{flushleft}
-    \small\setlength{\DTbaselineskip}{10pt}
+    \scriptsize\singlespacing\setlength{\DTbaselineskip}{9pt}
     \dirtree{%
@@
         .2 README.md.
-        .2 LICENSE.
         .2 Makefile.
         .2 mapop.gperf.
-        .2 paper.tex.
-        .2 bibliography.bib.
         .2 pseudocode.
@@
 have a corresponding header file ending in .h. This is due to the nature
-of the compilation process of the C language, which is the host language
-the interpreter of \name is written in.
+of the compilation process of the C language\cte{c_compilation}, which is the host language
+the \name interpreter is written in.
 Each header file contains only information about functions, types or variables
-available to other source or header files. The actual implementations or values
-are stored in the respective .c files.
+available to other source or header files. The actual implementations or 
+variable values are stored in the respective .c files.
 % TODO:explain C compilation or link example
 
-The pseudocode directory contains only examples used in this paper, no acutal,
-working code. The paper is written in \LaTeX and BibLaTeX and compiled with
-the tools pdflatex and biber. The paper's source files are paper.tex and
-bibliography.bib.
-
-The file mapop.gperf is a file to generate a perfect-hash-function\footcite{hash}
-with the GNU gperf tool. The generated mapop.c file is used to map string operators
-like '==' or '+' to a simpler number representation in constant time.
+The file mapop.gperf is also a source file. It is used in conjunction with the GNU gperf
+tool. The tool creates a perfect hash function\cte{hash}, which is output to the file
+src/interpreter/mappings/mapop.c. The perfect hash function is used to translate
+single- or multi-character symbols, such as '!=' and '+', to their respective
+interpreter internal representation in constant time.
 
@@
 writing and compiling C code. One such example is the article 'The C Beginner's
-Handbook' provided by freeCodeCamp.org linked in the sources\footcite{freeCodeCamp}
+Handbook' provided by freeCodeCamp.org linked in the sources\cte{freeCodeCamp}
 on page \pageref{bibliography}.
@@
 The parser, implemented in src/interpreter/parser.c, generates a syntax-tree
-from tokens. The algorithm used is inspired by the Pratt Parsing\footcite{pratt}
+from tokens. The algorithm used is inspired by the Pratt Parsing\cte{pratt}
 algorithm first described by Vaughan Pratt in 1973, which is a kind of precedence
 parser based on recursive descent. There is a great simple explanation of the algorithm
-by Jonathan Apodaca on dev.io\footcite{devio} linked in the references on page 
+by Jonathan Apodaca on dev.io\cte{devio} linked in the references on page 
 \pageref{bibliography} which this paper's parser takes inspiration from.
@@
 The algorithm used is inspired by Edsger Dijkstra's 'Shunting yard 
-algorithm'\footcite{shunting_yard}, traversing the tree post-order to create
+algorithm'\cte{shunting_yard}, traversing the tree post-order to create
 a prefix notation.
@@
 % TODO: add examples in /examples/
-\name is built upon a well structured implementation of a recursive Pratt
+\name is built upon the implementation of a recursive Pratt
 parser and a basic tokenizer. Most actual features are reliant upon this
@@
     and is wrapper to C's function \emph{printf}. The \emph{std.print} function
-    takes any \name variable or value and prints it to standard output. On line 13, the "\n"
+    takes any \name variable or value and prints it to standard output. On line 13, the "\\n"
     is replaced by a newline-character, as \name can interpret Java-like escape-sequences.
@@
 Such tools often include a package-manager, with which people around the world
-can share code. Examples would be npm\footcite{npm} or cargo\footcite{cargo}, 
+can share code. Examples would be npm\cte{npm} or cargo\cte{cargo}, 
 the 'Node Package Manager' for JavaScript applications and and Rust's package 
@@
 \subsubsection{Foreign Function Interfaces}
-A foreign function interface\footcite{FFI} often abbreviated as FFI,
+A foreign function interface\cte{FFI} often abbreviated as FFI,
 is a mechanism by which one programming language can call functions written 
@@
 Java enables users to call C, C++ and even assembly code with the Java Native
-Interface\footcite{JNI}, JNI for short, whereas within Python C function can
-be run with the standard ctypes\footcite{ctypes} library.
+Interface\cte{JNI}, JNI for short, whereas within Python C function can
+be run with the standard ctypes\cte{ctypes} library.
 
@@
 received from the function in the correct format, preventing a segmentation
-fault. The libffi\footcite{libffi} C-library, which is also used by Java and
+fault. The libffi\cte{libffi} C-library, which is also used by Java and
 Python for this purpose, can be used to load and call such foreign functions.
@@
 \subsubsection{Reflection}
-Reflection\footcite{reflection}, also called reflective programming, describes
+Reflection\cte{reflection}, also called reflective programming, describes
 the ability of a program to examine and modify its own structure and behavior 
@@
 
-\newpage\printbibliography[heading=bibintoc]\pagelabel{bibliography}
+\newpage\section{Bibliography}
+\pagelabel{bibliography}\printbibliography[heading=none]
 

### 2022-10-22 src/interpreter/processing/implementations.c
add string concat

@@ -204,7 +204,52 @@ mementry *_div(opargs *args){ CALCULATE(a / b) }
 
 mementry *_mod(opargs *args){ CALCULATE( MOD(a, b) ) }
 
-mementry *_add(opargs *args){ CALCULATE(a + b) }
+mementry *_add(opargs *args){
+    mementry *ro = _recursiveprocess(args, DEREFERENCE); /* Load right into tmp mementry */
+    mementry *lo = _recursiveprocess(args, DEREFERENCE); /* Load left into dst mementry */
+    
+    /* make sure only to add numbers or strings (concat) */
+    if(ro->type != NUMBER && ro->type != STRING || lo->type != ro->type) {
+        throw(EI_WRONG_TYPE, args->info);
+        return NULL; 
+    }
+    
+    /* determine the dst entry (or allocate a new one if ro and lo are not synthetics) */
+    mementry *dst = NULL;
+    if(lo->flags.synthetic)
+        dst = lo;
+    else if(ro->flags.synthetic)
+        dst = ro;
+    else {
+        dst = malloc(sizeof(mementry));
+        dst->value = malloc(sizeof(number));
+        dst->type = ro->type;
+        dst->flags = 
+        (struct s_mementry_flags) {.mutable=0, .synthetic=1, .value_synthetic=1};
+    }
+    
+    char *valdst = dst->flags.value_synthetic ? dst->value : 
+        malloc(ro->type == STRING ? strlen((char *)ro->value)+strlen((char *)lo->value) + 1 : sizeof(number));
+    
+    /* perform the requested calculation */
+    if(ro->type == STRING){
+        strcpy(valdst, (char *) lo->value);
+        strcat(valdst, (char *) ro->value);
+    } else
+        *(number *)valdst = a + b;
+    
+    if(!dst->flags.value_synthetic){
+        dst->flags.value_synthetic = 1;
+        dst->value = valdst;
+    }
+    
+    /* free right operand if it's not the entry to be returned */
+    if(dst != ro)
+        _free_synth(ro);
+    
+    return dst;
+
+}
 
 mementry *_sub(opargs *args){ CALCULATE(a - b) }
 

### 2022-10-22 src/interpreter/libraries/stdlib/stdlib.c
fix bug where NULL should be undefined

@@ -23,7 +23,6 @@ void _type(mementry *args, mementry *dst) {
 
 // print a value to the console
 void _print(mementry *args, mementry *dst) {
-
     switch(args->type) {
         case NUMBER:
             printf("%Lg", *((number *)args->value));
@@ -44,6 +43,7 @@ void _print(mementry *args, mementry *dst) {
     dst->type = NUMBER;
     dst->value = malloc(sizeof(number));
     *((number *)dst->value) = (number) 1;
+    return;
 }
 
 // execute a system shell comand (sh on UNIX, cmd on Windows)

### 2022-10-22 src/interpreter/processing/implementations.c

@@ -49,8 +49,12 @@ mementry *_recursiveprocess(opargs *args, char flags) {
 
     // check if NULL
     if(memptr == NULL){
-        // return NULL
-        return NULL;
+        // return undefined instead 
+        memptr = malloc(sizeof(mementry));
+        memptr->type = UNDEFINED;
+        memptr->value = NULL;
+        memptr->flags = (struct s_mementry_flags) {.mutable = 0, .synthetic = 1, .value_synthetic = 0};
+        return memptr;
     }
 
     // TODO: handle other cases
@@ -338,7 +342,7 @@ mementry *_ass(opargs *args){
     mementry *src = _recursiveprocess(args, 0); // Load right operand into src
     mementry *dst = _recursiveprocess(args, 0); // Load left operand into dst
 
-    if(dst == NULL || dst->type != REFERENCE) {
+    if(dst->type != REFERENCE) {
         throw(EI_REQ_MUTABLE, args->info);
         return NULL;
     }
@@ -532,8 +536,6 @@ mementry *_index(opargs *args){
     mementry *index = _recursiveprocess(args, DEREFERENCE);
     mementry *arr = _recursiveprocess(args, DEREFERENCE);
 
-
-    printf("arr type: %i\n", arr->type);
     if(arr->type != ARRAY && arr->type != TUPLE || index->type != NUMBER) {
         throw(EI_WRONG_TYPE, args->info);
         return NULL;

### 2022-10-22 src/interpreter/bytecode.c
add function arguments

@@ -72,7 +72,7 @@ void _recursiveconsume(bytecode *dst, stackptr ptr, stnode *subtree) {
                 _recursiveconsume(dst, ptr, subtree->data.parent.right);
             // make sure to provide a right-hand argument if it's a function call
             else if(subtree->data.parent.op->opcode == OP_CALL)
-                _register(dst, ptr, UNDEFINED);
+                _register(dst, ptr, UNDEFINED)->value = NULL;
 
             // then push operation onto the stack
             _register(dst, ptr, EXPR)->value = 
@@ -116,7 +116,7 @@ void _recursiveconsume(bytecode *dst, stackptr ptr, stnode *subtree) {
 
         case FILE_END:
             // Do not push anything
-            break;;
+            break;
     }
     
     // free the subtree node

### 2022-10-22 src/interpreter/libraries/stdlib/stdlib.c

@@ -17,8 +17,8 @@ void _type(mementry *args, mementry *dst) {
         "block", "block_end", "member", "expr"
     };
 
-    dst->type = STRING;
     dst->value = strdup(typeNames[args->type]);
+    dst->type = STRING;
 }
 
 // print a value to the console
@@ -40,9 +40,9 @@ void _print(mementry *args, mementry *dst) {
             return;
     }
 
-    dst->type = NUMBER;
     dst->value = malloc(sizeof(number));
     *((number *)dst->value) = (number) 1;
+    dst->type = NUMBER;
     return;
 }
 

### 2022-10-22 src/interpreter/mappings/mapop.c

@@ -25,7 +25,8 @@
       && ('s' == 115) && ('t' == 116) && ('u' == 117) && ('v' == 118) \
       && ('w' == 119) && ('x' == 120) && ('y' == 121) && ('z' == 122) \
       && ('{' == 123) && ('|' == 124) && ('}' == 125) && ('~' == 126))
-#error "gperf generated tables don't work with this execution character set."
+/* The character set is not based on ISO-646.  */
+#error "gperf generated tables don't work with this execution character set. Please report a bug to <bug-gperf@gnu.org>."
 #endif
 
 #line 8 "./mapop.gperf"
@@ -44,7 +45,7 @@
 #line 23 "./mapop.gperf"
 struct s_op;
 
-#define TOTAL_KEYWORDS 25
+#define TOTAL_KEYWORDS 26
 #define MIN_WORD_LENGTH 1
 #define MAX_WORD_LENGTH 2
 #define MIN_HASH_VALUE 1
@@ -66,10 +67,10 @@ hash (register const char *str, register size_t len)
       63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
       63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
       63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
-      63, 63, 63,  8, 63, 63, 63, 31, 30, 63,
-      63, 63, 26, 25,  6,  5, 63, 28, 63, 63,
-      63, 63, 63, 63, 63, 63, 63, 63, 23, 18,
-       3,  0, 15, 63, 63, 63, 63, 63, 63, 63,
+      63, 63, 63,  8, 63, 63, 63, 14, 30, 63,
+      63, 63, 31, 25, 26,  5, 63,  6, 63, 63,
+      63, 63, 63, 63, 63, 63, 63, 63, 28, 23,
+       3,  0, 15, 63, 18, 63, 63, 63, 63, 63,
       63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
       63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
       63, 63, 63, 63, 13, 63, 63, 63, 63, 63,
@@ -110,68 +111,69 @@ lookup (register const char *str, register size_t len)
   static struct s_op wordlist[] =
     {
       {""},
-#line 46 "./mapop.gperf"
+#line 47 "./mapop.gperf"
[…]

### 2022-10-22 src/interpreter/mappings/mapop.h

@@ -17,7 +17,7 @@ typedef enum {
     OP_BAND, OP_BXOR, OP_BOR, OP_LAND, OP_LOR, 
     OP_LAMBDA, OP_ASS, OP_HARDSET,
     OP_MULASS, OP_DIVASS, OP_MODASS, 
-    OP_LIST, OP_END,
+    OP_LIST, OP_END, OP_ARGS,
 
     // Special prefix operators
     OP_PREINCR, OP_PREDECR, OP_POS, OP_NEG,

### 2022-10-22 src/interpreter/mappings/operations.c

@@ -41,6 +41,7 @@ operation implementationof(opcode code) {
             [OP_POS]       = _pos,
             [OP_NEG]       = _neg,
 
+            [OP_ARGS]       = _args,
             [OP_CALL]      = _call,
             [OP_INDEX]     = _index,
             [OP_ARRAY]     = _array,

### 2022-10-22 src/interpreter/parser.c

@@ -1,4 +1,5 @@
 #include "parser.h"
+#include "tokenizer.h"
 
 // helper function to allocate an stnode
 static inline stnode *allocate_stnode() {
@@ -351,8 +352,16 @@ stnode *secondary(cache *cache, unsigned char rbp){
                 ret->type = FILE_END;
                 return ret;
             }
+
             // advance to unload operator
             advance(cache);
+
+            // the args operator does not need any actual argument
+            if(ret->data.parent.op->opcode == OP_ARGS) {
+                // use FILE_END because left child is required
+                ret->data.parent.left = allocate_typed(FILE_END);
+                return ret;
+            }
 
             break;
 

### 2022-10-22 src/interpreter/processing/implementations.c

@@ -40,7 +40,6 @@ void _free_synth(mementry *entry) {
     }
 }
 
-
 mementry *_recursiveprocess(opargs *args, char flags) {
     // get the next mementry from the bytecode
     mementry *memptr = *(mementry **)pop(args->code, &args->offset, sizeof(mementry *));
@@ -372,6 +371,10 @@ mementry *_ass(opargs *args){
     // create value reference
     dst->value = src->value;
 
+    // reset flags
+    dst->flags = 
+        (struct s_mementry_flags) {.mutable = 0, .synthetic = 0, .value_synthetic = 0};
+
     // free src if synthetic
     if(src->flags.synthetic)
         free(src);
@@ -463,6 +466,15 @@ mementry *_neg(opargs *args){
     return res;
 }
 
+
+// global args object used for function calls
+mementry *_g_params = NULL;
+
+mementry *_args(opargs *args) {
+    // used in functions to retreive current function arguments
+    return _g_params;
+}
+
 mementry *_call(opargs *args){
     mementry *params = _recursiveprocess(args, DEREFERENCE); // Load arguments
     mementry *fun = _recursiveprocess(args, DEREFERENCE); // Load function
@@ -480,7 +492,20 @@ mementry *_call(opargs *args){
             // reference the function code stack
             new_args.code = (bytecode *)fun->value;
 
-            // TODO: put arguments into table
+            // register arguments, assert tuple 
+            mementry *reset = _g_params;
+            if(params->type != TUPLE) {
+                array *arr = create_array(1);
+
+                _g_params = malloc(sizeof(mementry));
+                _g_params->type = TUPLE;
+                _g_params->value = arr;
+                _g_params->flags = 
+                    (struct s_mementry_flags) {.mutable = 0, .synthetic = 1, .value_synthetic = 0};
+
+                set_element(arr, params, 0);
+            } else
+                _g_params = params;
 
             // copy debug info
             new_args.info = args->info;
@@ -494,6 +519,10 @@ mementry *_call(opargs *args){
                     _free_synth(dst);
                 dst = _recursiveprocess(&new_args, 0);
             }
+
+            // reset arguments to parent function
+            // ? _free_synth(_g_params);
+            _g_params = reset;
 
             break;
         }
@@ -519,15 +548,14 @@ mementry *_call(opargs *args){
         default:
             // Throw an error if its not a function
             throw(EI_NOT_CALLABLE, args->info);
-            break;
[…]

### 2022-10-22 src/interpreter/processing/implementations.h

@@ -43,6 +43,7 @@ mementry *_predecr(opargs *args);
 mementry *_pos(opargs *args);
 mementry *_neg(opargs *args);
 
+mementry *_args(opargs *args);
 mementry *_call(opargs *args);
 mementry *_index(opargs *args);
 mementry *_array(opargs *args);

### 2022-10-22 src/interpreter/tokenizer.c

@@ -221,11 +221,15 @@ void readtkn(cache *cache) {
             default:
                 // set type
                 dst->type = SYMBOL;
+                char last, check = 0;
 
                 // read all following punctuation characters
-                while (ispunct(cur) && (cur != '\'' && cur != '"' )) {
-                    *con++ = cur;
+                while (ispunct(cur) && cur != '\'' && cur != '"') {
+                    if(check && (last == '@' || cur == '@'))
+                        break;
+                    *con++ = last = cur;
                     cur = next(cache);
+                    check = 1;
                 };
 
                 // push back last read character if not space

### 2022-10-22 src/interpreter/error/error.c
add std.repeat function

@@ -26,6 +26,7 @@ static const char *errformats[] = {
         "field was created in a scope equal or lower than the destination field",
     "Index out of bounds",
 
+    "Error during execution of if-statement",
     "Loading of a shared object failed",
     "Loading of foreign function failed",
     "An error occured within libffi when creating a typed foreign function",

### 2022-10-22 src/interpreter/error/error.h

@@ -30,6 +30,7 @@ enum errorformat {
     EI_INDEX_OOB,
 
     // Foreign function errors
+    EF_IF_STATEMENT,
     EF_DL_LOADING_FAILED,
     EF_FF_LOADING_FAILED,
     EF_FFI_ERROR,

### 2022-10-22 src/interpreter/libraries/libraries.c

@@ -1,7 +1,7 @@
 #include "libraries.h"
 
-char callfunc(cfunction *func, mementry *params, mementry *dst) {
+char callfunc(cfunction *func, mementry *params, mementry *dst, errorinfo *info) {
     
     // call wfunction and return
-    ((wfunction)(func->funptr))(params, dst);
+    ((wfunction)(func->funptr))(params, dst, info);
 }

### 2022-10-22 src/interpreter/libraries/libraries.h

@@ -1,7 +1,8 @@
 #include "../memory/hashtable.h"
+#include "../error/error.h"
 
 // type of a 'wrapped' function
-typedef void (*wfunction)(mementry *args, mementry *dst);
+typedef void (*wfunction)(mementry *args, mementry *dst, errorinfo* info);
 
 // definition of a c-function
 typedef struct {
@@ -13,4 +14,4 @@ typedef struct {
 
 
 // function to call a cfunction
-char callfunc(cfunction *func, mementry *params, mementry *dst);
+char callfunc(cfunction *func, mementry *params, mementry *dst, errorinfo *info);

### 2022-10-22 src/interpreter/libraries/stdlib/stdlib.c

@@ -61,7 +61,7 @@ void _exec(mementry *args, mementry *dst) {
     *((number *)dst->value) = (number) 1;
 }
 
-void _if(mementry *args, mementry *dst) {
+void _if(mementry *args, mementry *dst, errorinfo *info) {
     if(args->type == TUPLE){
         array *arr = ((array *)args->value);
         mementry *fun = NULL;
@@ -94,7 +94,34 @@ void _if(mementry *args, mementry *dst) {
         }
     }
     // error during if statement
-    printf("[error during execution of if-statement]\n");
+    throw(EF_IF_STATEMENT, info);
+}
+
+// while loop
+void _repeat(mementry *args, mementry *dst, errorinfo *info) {
+    if(args->type == REFERENCE)
+        args = args->value;
+    if(args->type != FUNCTION) {
+        dst->type = NUMBER;
+        dst->value = malloc(sizeof(number));
+        *((number *)dst->value) = (number) 0;
+        return; 
+    }
+
+    // prepare undefined mementry
+    mementry *params = malloc(sizeof(mementry));
+    params->type = UNDEFINED;
+
+    // call function
+    mementry *tmp;
+    do {
+        tmp = call(args, params, info);
+        dst->type = tmp->type;
+        dst->value = tmp->value;
+    } while(truth_of(dst));
+    
+    // free params
+    free(params);
 }
 
 // -----
@@ -112,6 +139,7 @@ void loadstd(hashtable *table){
         { _print, "print" },
         { _type,  "type" },
         { _exec,  "exec" },
+        { _repeat,"repeat" },
         { _if,    "if" },
 
         // ---

### 2022-10-22 src/interpreter/processing/implementations.c

@@ -475,10 +475,7 @@ mementry *_args(opargs *args) {
     return _g_params;
 }
 
-mementry *_call(opargs *args){
-    mementry *params = _recursiveprocess(args, DEREFERENCE); // Load arguments
-    mementry *fun = _recursiveprocess(args, DEREFERENCE); // Load function
-
+mementry *call(mementry *fun, mementry *params, errorinfo *info){
     // mementry for the result
     mementry *dst = NULL;
 
@@ -508,7 +505,7 @@ mementry *_call(opargs *args){
                 _g_params = params;
 
             // copy debug info
-            new_args.info = args->info;
+            new_args.info = info;
 
             // create a stack pointer, starting at the end
             new_args.offset = new_args.code->elements;
@@ -540,14 +537,14 @@ mementry *_call(opargs *args){
             }
 
             // call function from shared object
-            callfunc((cfunction *) fun->value, params, dst);
+            callfunc((cfunction *) fun->value, params, dst, info);
 
             // return the result that's stored in dst now
             break;
 
         default:
             // Throw an error if its not a function
-            throw(EI_NOT_CALLABLE, args->info);
+            throw(EI_NOT_CALLABLE, info);
             _free_synth(fun);
             return NULL;
     }
@@ -558,6 +555,13 @@ mementry *_call(opargs *args){
         _free_synth(fun);
 
     return dst;
+}
+
+mementry *_call(opargs *args){
+    mementry *params = _recursiveprocess(args, DEREFERENCE); // Load arguments
+    mementry *fun = _recursiveprocess(args, DEREFERENCE); // Load function
+
+    return call(fun, params, args->info);
 }
 
 mementry *_index(opargs *args){

### 2022-10-22 src/interpreter/processing/implementations.h

@@ -11,6 +11,8 @@ mementry *_recursiveprocess(opargs *args, char return_reference);
 void _free_synth(mementry *entry);
 char truth_of(mementry *entry);
 
+mementry *call(mementry *fun, mementry *params, errorinfo *info);
+
 // ----- Declarations -----
 mementry *_incr(opargs *args);
 mementry *_decr(opargs *args);

### 2022-10-22 paper.tex
continue writing paper

@@
-\documentclass[12pt,a4paper]{report}
+\documentclass[12pt,a4paper]{article}
 \usepackage{listings,xcolor,setspace,hyperref,dirtree}
 \usepackage{catchfile,pgffor}
+\usepackage[margin=2.5cm]{geometry} % Set page margin to 2.5cm
 \usepackage{times} % Use Times New Roman font
@@
 
-\maketitle\tableofcontents\newpage
+\begin{titlepage}\begin{center}
+
+    \vspace*{2.5cm}
+    \Huge
+    \textbf{The Creation of a Programming Language}
+
+    \Large
+
+    \vspace{1.5cm}
+    Matura Paper by\\ \textbf{Florian Donnelly}
+
+    \vspace{0.5cm}
+    With Supervision of\\ \textbf{Adrian Lüthi}
+
+    \vfill
+    Gymnasium Burgdorf\\ October 31, 2022
+
+\end{center}\end{titlepage}
 
@@
 
+\tableofcontents\newpage
+
 % the name of the programming language
@@
 \section{Introduction}
-\subsection{Relevance of the Topic}
+The importance of computers in today's world is unquestioned.
+But computers are not just the metals and semiconductors they are built from,
+but would be unusable without any programs running on them.
+Creating such programs, no matter if games, web browsers or even operating systems,
+involve the use of programming languages.
+Those serve as a bridge between the human mind and its imagination, and
+the capabilities of computer hardware and are a huge time-saver when compared
+to writing machine-native code, i.e. not using a programming language.
+
+Since the year 1973, when the first programming language was invented, the
+research in programming languages has fastly advanced. But the goal has stayed
+the same, to make programming as intuitive and easy, yet fast and efficient as
+possible.
+
+A programming language is itself only a description or specification, i.e. a
+set of grammar rules defining how single instructions and entire programs
+can be written. Source code, which is a program written in a certain language,
+has to be understood by the computer hardware in order to run. It can either
+be translated to machine-native bytecode with the help of a compiler, or
+read line-by-line and interpreted by another program known as interpreter.
+Such compilers or interpreters are programs on their own that are classified
+as programming language implementation. The focus of this paper will be to
+create such a programming language implementation, more specific an interpreter
+of the custom-made \name language.
+
 \subsection{Goals}
+The focus of this project is learning about the internal design of interpreters
+by creating a custom interpreter named \name interpreter, building upon
+known strategies and algorithms used in existing programming languages.
+The goal of this paper is to further understand the inner workings of
+interpreters, stepping through human-readable source code and computing
+desired results.
+
+Modern programming languages often times provide additional tools aside from
+the implementation. This includes dependency management tools to
+easily create projects extending upon existing ones, or a web platforms for
+people to share their code, all of which is not subject of this paper.
+The \name implementation will also not include many features seen in modern
+languages, as those are often created in years of time spent planning and
+programming.
 
 \section{Theory}
-\subsection{What this is and what this is not}
-The programming language and the tools that surround it are not meant to be
-viable options for real-world projects. Instead, the final product, \name,
-should be a programming language interpreter which includes techniques and
-algorithms already present in other, well known interpreters.
-The goal of this paper is to further understand the inner workings of
-interpreters, transforming a human-readable programming language
-into a computer-executable program.
-
-Modern programming languages often times include additional tools, aside from
-the interpreter or compiler. This includes dependency management tools to
-easily create projects exdending upon existing ones, or a web-plattform for
-people to share their code.
-
-The process used in this paper is more so a very core implementation of what
-a programming language, only containing the very necessary features.
-
-\subsection{Different types of interpreters}
+\subsection{Different Types of Interpreters}
 In computing, there are several types of interpreters that behave differently.
@@
 % Append all soruce files
-\foreach \f in \files{\pagelabel{\f}\code{\f}{1}{9999}}
+% \foreach \f in \files{\pagelabel{\f}\code{\f}{1}{9999}}
 

### 2022-10-22 src/interpreter/parser.c
fix parser bug where -/+ don't behave correctly

@@ -105,6 +105,7 @@ char in_expression; // does not work for multi-thread
 stnode *expr(cache *cache, unsigned char rbp) {
 
     stnode *left = secondary(cache, rbp);
+   
     in_expression = 1;
 
     // If secondary is done, just return
@@ -362,6 +363,9 @@ stnode *secondary(cache *cache, unsigned char rbp){
                 ret->data.parent.left = allocate_typed(FILE_END);
                 return ret;
             }
+
+            // make sure to initialize right child
+            ret->data.parent.right = NULL;
 
             break;
 

### 2022-10-22 src/interpreter/processing/implementations.c

@@ -327,11 +327,9 @@ mementry *_bxor(opargs *args){
 mementry *_bor(opargs *args){
     CALCULATE( (number) ((long long)a | (long long) b) ) }
 
-mementry *_land(opargs *args){
-}
+mementry *_land(opargs *args){ CALCULATE(a && b) }
 
-mementry *_lor(opargs *args){
-}
+mementry *_lor(opargs *args){ CALCULATE(a || b) }
 
 mementry *_lambda(opargs *args){
 

### 2022-10-22 src/interpreter/error/error.c
typo

@@ -19,7 +19,7 @@ static const char *errformats[] = {
     "Token unmatched",
 
     "Missing argument(s) for given operator",
-    "The specified operator cannot be used on of given type(s)",
+    "The specified operator cannot be used on given type(s)",
     "The specified operand is not a function, therefore it cannot be called",
     "The specified operand must be mutable",
     "You can only assign the value from one field to another, if the source "

### 2022-10-22 src/interpreter/libraries/stdlib/stdlib.c
added std.len

@@ -9,7 +9,7 @@
 // ----- stdlib implementations -----
 
 // return type-string of given argument
-void _type(mementry *args, mementry *dst) {
+void _type(mementry *args, mementry *dst, errorinfo *info) {
     static const char* typeNames[] = {
         "number", "string", "undefined", "array", "tuple",
         "object", "function", "cfunction", "reference", 
@@ -22,7 +22,7 @@ void _type(mementry *args, mementry *dst) {
 }
 
 // print a value to the console
-void _print(mementry *args, mementry *dst) {
+void _print(mementry *args, mementry *dst, errorinfo *info) {
     switch(args->type) {
         case NUMBER:
             printf("%Lg", *((number *)args->value));
@@ -47,7 +47,7 @@ void _print(mementry *args, mementry *dst) {
 }
 
 // execute a system shell comand (sh on UNIX, cmd on Windows)
-void _exec(mementry *args, mementry *dst) {
+void _exec(mementry *args, mementry *dst, errorinfo *info) {
     dst->type = NUMBER;
     dst->value = malloc(sizeof(number));
 
@@ -124,6 +124,26 @@ void _repeat(mementry *args, mementry *dst, errorinfo *info) {
     free(params);
 }
 
+// get lenght of arrays, tuples, strings
+void _len(mementry *args, mementry *dst, errorinfo *info) {
+    number num = -1;
+    switch(args->type) {
+        case STRING:
+            num = strlen((char *)args->value);
+            break;
+
+        case TUPLE:
+        case ARRAY:
+            num = ((array *) args->value)->size;
+            break;
+    }
+
+    dst->value = malloc(sizeof(number));
+    dst->type = NUMBER;
+    *((number *)dst->value) = num;
+    return;
+}
+
 // -----
 
 void loadstd(hashtable *table){
@@ -136,10 +156,11 @@ void loadstd(hashtable *table){
 
         // --- Array of all functions and their names in the stdlib
 
+        { _repeat,"repeat" },
         { _print, "print" },
         { _type,  "type" },
         { _exec,  "exec" },
-        { _repeat,"repeat" },
+        { _len,   "len" },
         { _if,    "if" },
 
         // ---

### 2022-10-22 src/interpreter/processing/implementations.c
add string indexing

@@ -564,21 +564,53 @@ mementry *_call(opargs *args){
 
 mementry *_index(opargs *args){
     mementry *index = _recursiveprocess(args, DEREFERENCE);
-    mementry *arr = _recursiveprocess(args, DEREFERENCE);
+    mementry *val = _recursiveprocess(args, DEREFERENCE);
+  
+    if(index->type == NUMBER) {
+        size_t i = (size_t) *(number *)index->value;
 
-    if(arr->type != ARRAY && arr->type != TUPLE || index->type != NUMBER) {
-        throw(EI_WRONG_TYPE, args->info);
-        return NULL;
-    }
+        if(val->type == STRING) {
+            if(i < 0 || i >= strlen((char *)val->value)) {
+                throw(EI_INDEX_OOB, args->info);
+                return NULL;
+            }
+           
+            // create destination register
+            mementry *dst;
+            if(val->flags.value_synthetic)
+                dst = val;
+            else
+                (dst = malloc(sizeof(mementry)))->type = STRING;
 
-    size_t i = (size_t) *(number *)index->value;
-    if(i < 0 || i >= ((array *)arr->value)->size) {
-        throw(EI_INDEX_OOB, args->info);
-        return NULL;
+            // create new substring
+            char *substring = malloc(2);
+
+            // load substring
+            substring[0] = *((char *)val->value + i);
+            substring[1] = '\0';
+
+            // allocate previous string
+            if(dst == val){
+                if(val->flags.value_synthetic)
+                    free(val->value);
+            } else
+                _free_synth(dst);
+
+            // set and return substring
+            dst->value = substring;
+            return dst; 
+        } else {
+            if(i < 0 || i >= ((array *)val->value)->size) {
+                throw(EI_INDEX_OOB, args->info);
+                return NULL;
+            }
+            _free_synth(val);
+            return get((array *)val->value, i);
+        }
     }
-   
-    return get((array *)arr->value, i);
-}
+    throw(EI_WRONG_TYPE, args->info);
+    return NULL;
+ }
 
 mementry *_array(opargs *args){
     // array of references

### 2022-10-22 src/interpreter/memory/array.c
fix tokenization error & segfault

@@ -82,7 +82,7 @@ char set_element(array *arr, mementry *element, size_t index) {
 }
 
 mementry *get(array *arr, size_t index) {
-    if(index >= arr->size)
+    if(arr == NULL || index >= arr->size)
         return NULL;
     return *(arr->arr + index);
 }

### 2022-10-22 src/interpreter/tokenizer.c

@@ -113,7 +113,7 @@ void readtkn(cache *cache) {
                     if(cur == '\\'){
                         // read next character
                         cur = next(cache);
-                        if(!cur || cur == quotation)
+                        if(!cur)
                             break;
 
                         switch (cur) {

### 2022-10-22 src/interpreter/tokenizer.h

@@ -5,7 +5,7 @@
 #include "error/error.h"
 #include "memory/stack.h"
 
-#define MAX_CONTENT 256
+#define MAX_CONTENT 1024 
 
 /*
 * The Tokenizer takes a string and converts it to easy-to-understand tokens

### 2022-10-22 src/interpreter/libraries/libraries.c
add mandelbrot example

@@ -4,4 +4,5 @@ char callfunc(cfunction *func, mementry *params, mementry *dst, errorinfo *info)
     
     // call wfunction and return
     ((wfunction)(func->funptr))(params, dst, info);
+    return 1;
 }

### 2022-10-22 src/interpreter/libraries/stdlib/stdlib.c

@@ -1,8 +1,11 @@
+#include <math.h>
+
 #include "stdlib.h"
 #include "../../processing/implementations.h"
 #include "../../error/error.h"
 #include "../../bytecode.h"
 #include "../../memory/array.h"
+#include "../../memory/hashtable.h"
 
 #define ALLOC 1
 
@@ -34,9 +37,7 @@ void _print(mementry *args, mementry *dst, errorinfo *info) {
 
         // if anything else is passed, return 0
         default:
-            dst->type = NUMBER;
-            dst->value = malloc(sizeof(number));
-            *((number *)dst->value) = (number) 0;
+            dst->type = UNDEFINED;
             return;
     }
 
@@ -48,16 +49,15 @@ void _print(mementry *args, mementry *dst, errorinfo *info) {
 
 // execute a system shell comand (sh on UNIX, cmd on Windows)
 void _exec(mementry *args, mementry *dst, errorinfo *info) {
-    dst->type = NUMBER;
-    dst->value = malloc(sizeof(number));
-
     if(args->type != STRING){
-        *((number *)dst->value) = (number) 0;
+        dst->type = UNDEFINED;
         return;
     }
     
     system((char*) args->value);
     
+    dst->type = NUMBER;
+    dst->value = malloc(sizeof(number));
     *((number *)dst->value) = (number) 1;
 }
 
@@ -102,14 +102,12 @@ void _repeat(mementry *args, mementry *dst, errorinfo *info) {
     if(args->type == REFERENCE)
         args = args->value;
     if(args->type != FUNCTION) {
-        dst->type = NUMBER;
-        dst->value = malloc(sizeof(number));
-        *((number *)dst->value) = (number) 0;
+        dst->type = UNDEFINED;
         return; 
     }
 
     // prepare undefined mementry
-    mementry *params = malloc(sizeof(mementry));
+    mementry *params = _alloc_mementry();
     params->type = UNDEFINED;
 
     // call function
@@ -136,6 +134,9 @@ void _len(mementry *args, mementry *dst, errorinfo *info) {
         case ARRAY:
             num = ((array *) args->value)->size;
             break;
+        default:
+            dst->type = UNDEFINED;
+            return;
     }
 
     dst->value = malloc(sizeof(number));
@@ -144,6 +145,38 @@ void _len(mementry *args, mementry *dst, errorinfo *info) {
     return;
 }
 
[…]

### 2022-10-22 src/interpreter/memory/hashtable.c

@@ -15,6 +15,15 @@ static inline _H_HASH _hash(char* key) {
     return hash;
 }
 
+// return new mementry with default values
+mementry *_alloc_mementry() {
+    mementry *ret = malloc(sizeof(mementry));
+    ret->flags =
+        (struct s_mementry_flags) 
+        {.mutable = 0, .synthetic = 0, .value_synthetic = 0};
+    return ret;
+}
+
 hashtable *create_hashtable(size_t width, size_t cache_size) {
     hashtable *ret = (hashtable *) malloc(sizeof(hashtable) + width * sizeof(tableslice *));
 

### 2022-10-22 src/interpreter/memory/hashtable.h

@@ -52,6 +52,9 @@ typedef struct {
 
 } mementry;
 
+// function to allocate mementry
+mementry *_alloc_mementry();
+
 // internally used hash types
 typedef uint32_t _H_HASH;
 #define _H_MAX_VALUE UINT32_MAX

### 2022-10-22 src/interpreter/parser.c

@@ -10,6 +10,7 @@ static inline stnode *allocate_stnode() {
 static inline stnode *allocate_typed(nodetype type) {
     stnode *node = (stnode *)malloc(sizeof(stnode));
     node->type = type;
+    node->data.parent.right = NULL;
     return node;
 }
 

### 2022-10-22 src/interpreter/processing/implementations.c

@@ -49,7 +49,7 @@ mementry *_recursiveprocess(opargs *args, char flags) {
     // check if NULL
     if(memptr == NULL){
         // return undefined instead 
-        memptr = malloc(sizeof(mementry));
+        memptr = _alloc_mementry();
         memptr->type = UNDEFINED;
         memptr->value = NULL;
         memptr->flags = (struct s_mementry_flags) {.mutable = 0, .synthetic = 1, .value_synthetic = 0};
@@ -177,7 +177,7 @@ mementry *_bnot(opargs *args) {
     else if(ro->flags.synthetic)\
         dst = ro;\
     else {\
-        dst = malloc(sizeof(mementry));\
+        dst = _alloc_mementry();\
         dst->value = malloc(sizeof(number));\
         dst->type = NUMBER;\
         dst->flags = \
@@ -212,7 +212,7 @@ mementry *_add(opargs *args){
     mementry *lo = _recursiveprocess(args, DEREFERENCE); /* Load left into dst mementry */
     
     /* make sure only to add numbers or strings (concat) */
-    if(ro->type != NUMBER && ro->type != STRING || lo->type != ro->type) {
+    if((ro->type != NUMBER && ro->type != STRING) || lo->type != ro->type) {
         throw(EI_WRONG_TYPE, args->info);
         return NULL; 
     }
@@ -224,7 +224,7 @@ mementry *_add(opargs *args){
     else if(ro->flags.synthetic)
         dst = ro;
     else {
-        dst = malloc(sizeof(mementry));
+        dst = _alloc_mementry();
         dst->value = malloc(sizeof(number));
         dst->type = ro->type;
         dst->flags = 
@@ -275,7 +275,7 @@ mementry *_equ(opargs *args){
     else if(ro->flags.synthetic)
         dst = ro;
     else {
-        dst = malloc(sizeof(mementry));
+        dst = _alloc_mementry();
         dst->value = malloc(sizeof(number));
         dst->type = NUMBER;
         dst->flags = 
@@ -409,7 +409,7 @@ mementry *_list(opargs *args){
         set_element(arr, right, 1);
 
         // create new mementry for array
-        mementry *ret = malloc(sizeof(mementry));
+        mementry *ret = _alloc_mementry();
         ret->type = TUPLE;
         ret->value = arr;
         return ret;
@@ -440,7 +440,7 @@ mementry *_pos(opargs *args){
     if(o->flags.synthetic)
         dst = o;
[…]

### 2022-10-22 src/interpreter/bytecode.c
fix unitialized variables

@@ -1,4 +1,5 @@
 #include "bytecode.h"
+#include "memory/hashtable.h"
 #include "processing/implementations.h"
 
 #define INITIAL_DATA_SIZE 32
@@ -9,7 +10,7 @@
 
 static inline mementry *_register(bytecode *code, stackptr ptr, typing type) {
     mementry *ret = 
-        *(mementry **)push(code, ptr, sizeof(mementry *)) = malloc(sizeof(mementry));
+        *(mementry **)push(code, ptr, sizeof(mementry *)) = _alloc_mementry();
     ret->type = type;
     return ret;
 }

### 2022-10-22 src/interpreter/memory/array.c
make tuples,arrays make more sense

@@ -46,7 +46,7 @@ char set(array *arr, array *element, size_t index) {
     memcpy(arr->arr + index, element->arr, sizeof(mementry *) * element->size);
 
     // increase size
-    arr->size++;
+    arr->size += element->size;
     return 1;
 }
 

### 2022-10-22 src/interpreter/processing/implementations.c

@@ -385,8 +385,8 @@ mementry *_hardset(opargs *args) {
 }
 
 mementry *_list(opargs *args){
-    mementry *right = _recursiveprocess(args, 0);
-    mementry *left = _recursiveprocess(args, 0);
+    mementry *right = _recursiveprocess(args, DEREFERENCE);
+    mementry *left = _recursiveprocess(args, DEREFERENCE);
 
     if(left->type == TUPLE) {
         if(right->type == TUPLE) {
@@ -614,7 +614,7 @@ mementry *_index(opargs *args){
 
 mementry *_array(opargs *args){
     // array of references
-    mementry *content = _recursiveprocess(args, 0);
+    mementry *content = _recursiveprocess(args, DEREFERENCE);
 
     // if tuple, just make array
     if(content->type == TUPLE){

### 2022-10-22 src/interpreter/processing/implementations.c
fix error, use repeat in example

@@ -547,11 +547,6 @@ mementry *call(mementry *fun, mementry *params, errorinfo *info){
             return NULL;
     }
 
-
-    // free arguments
-    if(dst != fun)
-        _free_synth(fun);
-
     return dst;
 }
 

### 2022-10-23 paper.tex
add mandelbrot to title-page
add figures to paper

@@
 \usepackage{listings,xcolor,setspace,hyperref,dirtree}
-\usepackage{catchfile,pgffor}
+\usepackage{catchfile,pgffor,graphicx,float,caption}
 \usepackage[margin=2.5cm]{geometry} % Set page margin to 2.5cm
@@
 
-\addbibresource{bibliography.bib}
+\addbibresource{paper/bibliography.bib}
 
@@
 % Read contents of files.txt into \files
-\CatchFileDef{\files}{files.txt}{\endlinechar=-1}
+\CatchFileDef{\files}{paper/files.txt}{\endlinechar=-1}
 
@@
 \newcommand{\code}[3] {
-    \lstset{language=C,numbers=left,basicstyle=\scriptsize,
-        frame=single,title=#1,keywordstyle=\color{blue},stepnumber=5,
-        stringstyle=\color{green},commentstyle=\color{gray},
+    \lstset{basicstyle=\scriptsize,keywordstyle=\color{blue},stepnumber=5,
+        stringstyle=\color[rgb]{.3,.5,.1},commentstyle=\color{gray},
         morecomment=[l][\color{magenta}]{\#}}
-    \lstinputlisting[firstnumber=#2,firstline=#2,lastline=#3]{#1}
+    \lstinputlisting[language=C,numbers=left,
+        frame=single,title=#1,
+        firstnumber=#2,firstline=#2,lastline=#3]{#1}
 }
+
+% remove heading from lof
+\makeatletter
+\renewcommand\listoffigures{%
+        \@starttoc{lof}%
+}
+\makeatother
+
+% set caption width
+\captionsetup{width=.7\linewidth}
 
@@
 
+\newcommand{\paste}[1]{
+    \begin{tabular}{c}
+    \lstinputlisting[inputencoding=utf8,extendedchars=true,
+        basicstyle=\tiny\bfseries,language={}]{#1}
+    \end{tabular}
+}
+
+% the name of the programming language
+\newcommand{\name}{\emph{TestScript}}
 \newcommand{\pagelabel}[1]{\phantomsection\label{#1}}
@@
 
-    \vspace*{2.5cm}
+    \vspace*{1.5cm}
     \Huge
@@
 
+    \vspace{1.5cm}
+
+    % Include Mandelbrot on title page
+    \begin{figure}[H]
+        \centering
+        \paste{paper/mandelbrot.tex}
+        \caption{ASCII representation of the Mandelbrot Set, generated with \name{}}
+        \label{f_mandelbrot}
+    \end{figure}
+
     \Large
-
     \vspace{1.5cm}
@@
 
+\normalsize
+
 \abstract{
@@
 \tableofcontents\newpage
-
-% the name of the programming language
-\newcommand{\name}{\emph{TestScript }}
 
@@
 create such a programming language implementation, more specific an interpreter
-of the custom-made \name language.
+of the custom-made \name{} language.
 
@@
 The focus of this project is learning about the internal design of interpreters
-by creating a custom interpreter named \name interpreter, building upon
+by creating a custom interpreter named \name{} interpreter, building upon
 known strategies and algorithms used in existing programming languages.
@@
 people to share their code, all of which is not subject of this paper.
-The \name implementation will also not include many features seen in modern
+The \name{} implementation will also not include many features seen in modern
 languages, as those are often created in years of time spent planning and
@@
 This paper will implement a mixture form of an AST compiler and bytecode
-interpreter. This is not practical in terms of improving \name, but serves
-the purpose of combining different features of different interpreters into one
-and therefore extending the scope of the project.
+interpreter. The advantage is that an AST carries along lots of information
+not needed in later stages of interpretation which can be discarded. The AST
+is converted to an intermediary bytecode format which only contains
+necessary information about the program. Such bytecode can then be cached,
+e.g. kept in memory, meaning it only has to be converted once and can be
+used many times, also leading to a performance benefit.
 
@@
 
-\subsection{The Art of Abstraction}
+\subsection{The Art of Simplification}
 When dealing with problems in general, it is helpful to map out and think about
@@
 \subsection{A Project Overview}
+In the following paragraphs, the project in its entirety will be briefly
+presented, giving an overview over files and correlations.
 
 \subsubsection{File Structure}
+\begin{minipage}{\textwidth}
+
 Here is a file tree overview of the project which will be referenced
@@
 only source files are shown. No build files, executable files or files that
-are of no further importance are depicted.
+are of no further importance regarding \name are depicted.
 \begin{flushleft}
-    \scriptsize\singlespacing\setlength{\DTbaselineskip}{9pt}
+    \scriptsize\setlength{\DTbaselineskip}{5pt}
     \dirtree{%
@@
 
+\end{minipage}
+
 As visible in the file tree, most of the source-code files, ending in .c,
@@
 of the compilation process of the C language\cte{c_compilation}, which is the host language
-the \name interpreter is written in.
+the \name{} interpreter is written in.
 Each header file contains only information about functions, types or variables
@@
 
-\subsubsection{Interpreter Construction}
+\subsubsection{Flowchart}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.6\textwidth]{paper/Data Flow Diagram.png}
+    \caption{A \name{} Data Flow Diagram}
+    \label{f_diagram}
+\end{figure}
+
+The above figure is a representation of how different components of the entire
+\name{} implementation interact and work together. Each next stage performs
+operations on the output data of the proper. The src/interpreter/interpreter.c
+file, here depicted as 'Interpreter', chains all stages together.
+
+\subsection{Implementing \name{}}
 This section will explain the process of constructing and the inner workings of
-the \name interpreter. It is therefore the most complicated and detailed section
+the \name{} interpreter. It is therefore the most complicated and detailed section
 in this paper. Examples will be made with both pseudo code and actual source
 code from the project to further illustrate the decisions made during
-interpreter design. File paths
-used are relative to the project's root directory.
+interpreter design. The file paths used are relative to the project's root directory.
 
-The syntax of \name is not specified in a formal programming language grammar,
+The syntax of \name{} is not specified in a formal programming language grammar,
 but rather implemented from scratch in the files src/interpreter/tokenizer.c 
@@
 This section deals with the file src/interpreter/interpreter.c which guides
-input data through the various stages of the \name interpreter in its
-function 'interpret' which is called from the program's main function.
-This section does not go in detail on the following stages, see their own sections.
+input data through the various stages of the \name{} interpreter in its
+function 'interpret', which is called from the program's main function.
 
@@
 After that, variables with names are 'localized', meaning their names are
-replaced by addresses where their value is stored by src/interpreter/localizer.c.
-The advantage of performing this step so early is that mapping variable names
-to memory is a rather resource-intensive task with time complexity of maximally 
-O(log n), n being the amount of different variables.
+mapped to addresses in memory where their value is stored. Doing this before
+bytecode compilation has a performance benefit. 
+The mapping is done with a custom hashtable datastructure with average time
+complexity of O(log n), n being the amount of different variables.
 
-The syntax tree could now be interpreted directly, but is instead converted
-to an intermediary internal bytecode format by src/interpreter/bytecode.c.
-This step is done so that code that is ran multiple times does not convey
+The syntax tree could now be interpreted directly, but as said, is instead converted
+to an intermediate internal bytecode format by src/interpreter/bytecode.c.
+This step is done so that code that is run multiple times does not convey
 the problem of time overhead when traversing the syntax tree, but the current
@@
 
-The syntax tree in the implementation of \name consists of 'stnode' type
+The syntax tree in the implementation of \name{} consists of 'stnode' type
 pointers, as defined in src/interpreter/parser.h. Each such node depending
@@
 either 'Left to Right' or 'Right to Left'. An example of associativity is
-\emph{'20 / 10 / 2'}. Both in C as well as in \name, the division operator is
+\emph{'20 / 10 / 2'}. Both in C as well as in \name{}, the division operator is
 left-to-right associative, meaning the stated expression is the same as
@@
 
-Not to be confused with machine-native bytecode, the bytecode compiler in \name
+Not to be confused with machine-native bytecode, the bytecode compiler in \name{}
 generates byte-sequences in an internal format, that are then interpreted. A
@@
 
-To generate such a sequence of bytes, \name recursively reads through the
+To generate such a sequence of bytes, \name{} recursively reads through the
 binary tree created by the parser and pushes found instructions onto a stack,
@@
         \expr{\emph{'- 1 (/ (* 2 4) (+ 5 6))'}}
-    \item But in \name the operands are pushed onto the stack is reverse order.
+    \item But in \name{} the operands are pushed onto the stack is reverse order.
         This would look like the following:
@@
         This is an ideal abstraction of how data 
-        is represented internally in \name.
+        is represented internally in \name{}.
 \end{enumerate}
@@
 \subsubsection{The Runtime}
-This is the last stage of interpreting program code as implemented in \name.
+This is the last stage of interpreting program code as implemented in \name{}.
 The runtime steps through the intermediary bytecode created by the bytecode
@@
 The evaluation of expressions also works in a recursive manner. Here is an 
-example of \name's bytecode notation and how it would be evaluated at runtime:
+example of \name{}'s bytecode notation and how it would be evaluated at runtime:
 \begin{itemize}
-    \item Assume the input in \name's bytecode to resemble the following expression: 
+    \item Assume the input in \name{}'s bytecode to resemble the following expression: 
         \expr{\emph{'- 6 * 3 4'}}
@@
 
-That's it! With all the possible expressions available, entire programs
-can now be interpreted without a problem.
+That is it. With all the expressions available, entire programs
+can now be interpreted.
 
@@
 % TODO: add examples in /examples/
-\name is built upon the implementation of a recursive Pratt
+\name{} is built upon the implementation of a recursive Pratt
 parser and a basic tokenizer. Most actual features are reliant upon this
-foundation, such as operators and the standard library built into \name.
+foundation, such as operators and the standard library built into \name{}.
 
 \subsection{Syntax}
-The syntax of \name can be compared to the syntax of JavaScript, except
+The syntax of \name{} can be compared to the syntax of JavaScript, except
 there are no keywords at all, and everything is controlled via operators,
@@
 
-Following is a piece of example \name code, printing out the first ten numbers
+Following is a piece of example \name{} code, printing out the first ten numbers
 of the fibonacci sequence, as well as a disection of how that same code is interpreted.
@@
 \item Yet another variable, \emph{f}, is declared and assigned a function as its
-    value on line 6. In \name, normal functions do not have any arguments, 
+    value on line 6. In \name{}, normal functions do not have any arguments, 
     but they can interact with variables that were created outside their scope.
@@
     On lines 12 and 13, the function \emph{std.print} is referenced. This function is implemented
-    in plain C code, which internally is not handled the same as regular functions in \name,
+    in plain C code, which internally is not handled the same as regular functions in \name{},
     and is wrapper to C's function \emph{printf}. The \emph{std.print} function
-    takes any \name variable or value and prints it to standard output. On line 13, the "\\n"
-    is replaced by a newline-character, as \name can interpret Java-like escape-sequences.
+    takes any \name{} variable or value and prints it to standard output. On line 13, the "\\n"
+    is replaced by a newline-character, as \name{} can interpret Java-like escape-sequences.
 \item Another standard-library function is \emph{std.if}. It takes two or three arguments.
@@
 
-\subsection{Security}
+\subsection{Name Shadowing}
 Here is another example, which in contrast to the example above, focuses on a
-security concept implemented in \name.
+security concept implemented in \name{}.
 \code{examples/exec.nts}{1}{999}
@@
 Line number six holds a comment, indicated by the two slashes at the start of the line.
-Comments are ignored by the \name interpreter and their purpose is to contain notes
+Comments are ignored by the \name{} interpreter and their purpose is to contain notes
 and explanations inside the source code itself. Another type of commends are block-comments
@@
 That is why in most cases, the \emph{std.exec} function can be a risk waiting to be exploited
-by malicious \name programs. To solve this problem, there is the hashtag-operator, whose purpose
+by malicious \name{} programs. To solve this problem, there is the hashtag-operator, whose purpose
 it is to delete access to a variable through the specified name. It is used on line 13 of the
@@
 have access or knowledge about the variables that exist in the file that includes them.
-In \name, each imported file would instead have access to all variables and functions that
+In \name{}, each imported file would instead have access to all variables and functions that
 are made available to it, instead of having to include or import them seperately.
 This leads to another problem, but that is what the next section will be all about.
-
-\subsection{Name Shadowing}
-% TODO
 
@@
 
-\subsection{Limitation}
-\name is a programming language without a doubt. But it is not perfect in any
+\subsection{Limitations}
+\name{} is a programming language without a doubt. But it is not perfect in any
 way. There are limitations that make it practically unusable for actual
-professional projects. In this section, a few shortcomings of \name are listed
+professional projects. In this section, a few shortcomings of \name{} are listed
 and reasoned over.
@@
 has long been discarded and is of no further use.
-In \name, memory leaks are a major problem. Even though barely noticable for small
-programs it interprets, \name does not have proper setups in place to free up
+In \name{}, memory leaks are a major problem. Even though barely noticable for small
+programs it interprets, \name{} does not have proper setups in place to free up
 all used memory that is not used any longer.
@@
 use Garbage-Collection, which automatically tracks unused allocated memory and
-frees it up automaticadlly. While this is also the case with \name, there are
-many internal flaws that have nothing to do with variables used in a \name
+frees it up automaticadlly. While this is also the case with \name{}, there are
+many internal flaws that have nothing to do with variables used in a \name{}
 program, but rather internally.
@@
 A major reason for the memory leaks, but not an excuse, is the nature of the
-C programming language the \name interpreter was written in. The C language and
+C programming language the \name{} interpreter was written in. The C language and
 compiler do not keep track of states in which certain blocks of memory are
@@
 \subsubsection{Single-Threaded}
-The \name interpreter does not implement any multi-threaded, meaning parallel
-running, programs as most other languages do. Instead, any \name program can
+The \name{} interpreter does not implement any multi-threaded, meaning parallel
+running, programs as most other languages do. Instead, any \name{} program can
 only be interpreted and run sequentially. Since most modern computer processors
@@
 
-\subsubsection{Token input size}
-In the current iteration, each token's content's size is limited to 256 bytes,
-which makes it impossible to use e.g. variable names longer thatn 256 ASCII
-characters. This would probably be the easyest thing to fix of here-listed 
-problems of \name, but since 256 characters are long enough for 
-illustration purposes, fixing this is not worth spending time on.
-
 \subsubsection{Lack of a Development Environment}
-This is certainly the greatest problem of \name, aside the countless smaller
+This is certainly the greatest problem of \name{}, aside the countless smaller
 bothering aspects.
 
-Pretty much any good, modern programming language does not only consist
-of it's specification and a compiler or interpreter, but of many more
-tools surrounding it.
-Such tools often include a package-manager, with which people around the world
+Most modern programming languages do not only consist
+of a language specification and a compiler or interpreter, but of many more
+tools surrounding them.
+Such tools can include a package-manager, with which people around the world
 can share code. Examples would be npm\cte{npm} or cargo\cte{cargo}, 
@@
 that are great to have and are used in other programming languages, but not 
-could not be implemented into \name.
+could not be implemented into \name{}.
 
@@
 Foreign function interfaces are specially difficult to implement, or implement
-right. Let us assume a programming language interpreter \name written in C that
+right. Let us assume a programming language interpreter \name{} written in C that
 would like to call some functions that are part of yet another C program or
@@
 is by the way where the term foreign functions originates from. This is why the
-\name developer would have to provide that information by re-declaring the
-function inside \name, so that, with some trickery, data can be passed to and
+\name{} developer would have to provide that information by re-declaring the
+function inside \name{}, so that, with some trickery, data can be passed to and
 received from the function in the correct format, preventing a segmentation
@@
 differently under different conditions at runtime.
-The easiest and simplest way to go about implementing Reflection in \name is
+The easiest and simplest way to go about implementing Reflection in \name{} is
 certainly to allow the interpretation of strings at runtime. That way, strings
@@
 Hereby I, Florian Donnelly, declare to have written this paper, also including
-the entire source code of \name by myself using resources listed under the
+the entire source code of \name{} by myself using resources listed under the
 'references' section on page \pageref{bibliography}.
@@
 
+\section{List of Figures}\listoffigures
+\vspace{1cm}
+Figure \ref{f_mandelbrot} was generated using a \name{} program, the source code
+can be found in the file examples/mandelbrot.nts.
+Figure \ref{f_diagram} was created using an online diagram creation tool available
+on app.diagrams.net \cte{drawio}.
+
 % TODO: check correctness of all file paths!! e.g. src/...
@@
 \newpage\section{Appendix}
-The appendix contains the complete source code of \name.
+The appendix contains the complete source code of \name{}.
 

### 2022-10-23 paper.tex
add make target paper_raw for paper without images

@@
 \usepackage[margin=2.5cm]{geometry} % Set page margin to 2.5cm
+\usepackage{etoolbox}
 \usepackage{times} % Use Times New Roman font
@@
 
+% Specify the version
+\newcommand{\ver}[1]{
+    \expandafter\ifstrequal\expandafter{\jobname}{paper}
+    {#1}{\begin{center}-Removed, see full version-\end{center}}
+}
+
 % Read contents of files.txt into \files
@@
 \newcommand{\code}[3] {
+    \ver{
     \lstset{basicstyle=\scriptsize,keywordstyle=\color{blue},stepnumber=5,
         stringstyle=\color[rgb]{.3,.5,.1},commentstyle=\color{gray},
-        morecomment=[l][\color{magenta}]{\#}}
-    \lstinputlisting[language=C,numbers=left,
+        morecomment=[l][\color{magenta}]{\#},
+        postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space}}
+    \lstinputlisting[language=C,numbers=left,breaklines=true,
         frame=single,title=#1,
         firstnumber=#2,firstline=#2,lastline=#3]{#1}
+    }
 }
@@
 
-\subsubsection{File Structure}
+\subsubsection{File Structure}\label{FileStructure}
 \begin{minipage}{\textwidth}
@@
 are of no further importance regarding \name are depicted.
-\begin{flushleft}
+\ver{\begin{flushleft}
     \scriptsize\setlength{\DTbaselineskip}{5pt}
@@
     }
-\end{flushleft}
+\end{flushleft}}
 
 \end{minipage}
+\linebreak
 
@@
 variable values are stored in the respective .c files.
-% TODO:explain C compilation or link example
+% TODO:explain C compilation or link example, fix hash citation below
 
@@
 
+\ver{
 \begin{figure}[H]
@@
 \end{figure}
+}
 
@@
 
+The entire source code can be found in the Appendix, section \ref{Appendix}.
+
 % TODO: check correctness of all file paths!! e.g. src/...
 
-\newpage\section{Appendix}
+\newpage\section{Appendix}\label{Appendix}
 The appendix contains the complete source code of \name{}.
+See the file tree under 'File Structure' in section \ref{FileStructure}
+to get an overview.
 
 % Append all soruce files
-% \foreach \f in \files{\pagelabel{\f}\code{\f}{1}{9999}}
+\ver{
+\foreach \f in \files{\pagelabel{\f}\code{\f}{1}{9999}}
+}
 

### 2022-10-23 paper.tex
make paper file-tree clickable

@@
 \usepackage{times} % Use Times New Roman font
-\usepackage[style=alphabetic,backend=biber]{biblatex}
+\usepackage[style=numeric,backend=biber]{biblatex}
 
@@
 % Cite text
-\newcommand{\cte}[1] {
-    \cite{#1}
-}
+\DeclareCiteCommand{\cte}
+  {\usebibmacro{prenote}}
+  {\usebibmacro{citeindex}%
+   \usebibmacro{cite}}
+  {\multicitedelim}
+  {\usebibmacro{postnote}}
 
@@
         JIT-compilers come with only advantages. The most popular example of
-        a JIT-compiler would be Google's V8\cte{V8} JavaScript engine used
+        a JIT-compiler would be Google's V8 JavaScript engine used
         in NodeJS and any Chromium browsers.
@@
         the same across any machine, the only difference being the interpreter.
-        The most popular example would be the Java Virtual Machine\cte{jvm},
+        The most popular example would be the Java Virtual Machine,
         a bytecode interpreter for java bytecode, also used by many other languages
@@
         An example of a compiler written in itself is rustc, the Rust
-        compiler\cte{rustc}, which was bootstrapped by the OCaml language.
+        compiler, which was bootstrapped by the OCaml language.
 \end{itemize}
@@
 used many times, also leading to a performance benefit.
+[\cte{V8}, \cte{jvm}, \cte{rustc}]
 
@@
 are of no further importance regarding \name are depicted.
+\newcommand{\hr}[2]{\hyperref[#2]{#1}}
 \ver{\begin{flushleft}
-    \scriptsize\setlength{\DTbaselineskip}{5pt}
+    \scriptsize\setlength{\DTbaselineskip}{7pt}
     \dirtree{%
         .1 project root.
-        .2 .gitignore.
-        .2 README.md.
-        .2 Makefile.
-        .2 mapop.gperf.
+        .2 \hr{.gitignore}{.gitignore}.
+        .2 \hr{README.md}{README.md}.
+        .2 \hr{Makefile}{Makefile}.
+        .2 \hr{mapop.gperf}{mapop.gperf}.
         .2 pseudocode.
-            .3 src/interpreter/parser.c.
+            .3 \hr{parser.c}{pseudocode/parser.c}.
+        .2 examples.
+            .3 \hr{mandelbrot.nts}{examples/mandelbrot.nts}.
+            .3 \hr{fibonacci.nts}{examples/fibonacci.nts}.
+            .3 \hr{exec.nts}{examples/exec.nts}.
         .2 src.
-            .3 main.c.
+            .3 \hr{main.c}{src/main.c}.
             .3 interpreter.
-                .4 bytecode.c.
-                .4 bytecode.h.
+                .4 \hr{bytecode.c}{src/interpreter/bytecode.c}.
+                .4 \hr{bytecode.h}{src/interpreter/bytecode.h}.
                 .4 error.
-                    .5 error.c.
-                    .5 error.h.
-                .4 interpreter.c.
-                .4 interpreter.h.
+                    .5 \hr{error.c}{src/interpreter/error/error.c}.
+                    .5 \hr{error.h}{src/interpreter/error/error.h}.
+                .4 \hr{interpreter.c}{src/interpreter/interpreter.c}.
+                .4 \hr{interpreter.h}{src/interpreter/interpreter.h}.
                 .4 libraries.
-                    .5 libraries.c.
-                    .5 libraries.h.
+                    .5 \hr{libraries.c}{src/interpreter/libraries/libraries.c}.
+                    .5 \hr{libraries.h}{src/interpreter/libraries/libraries.h}.
                     .5 stdlib.
-                        .6 stdlib.c.
-                        .6 stdlib.h.
-                .4 localizer.c.
-                .4 localizer.h.
+                        .6 \hr{stdlib.c}{src/interpreter/libraries/stdlib/stdlib.c}.
+                        .6 \hr{stdlib.h}{src/interpreter/libraries/stdlib/stdlib.h}.
+                .4 \hr{localizer.c}{src/interpreter/localizer.c}.
+                .4 \hr{localizer.h}{src/interpreter/localizer.h}.
                 .4 mappings.
-                    .5 operations.c.
-                    .5 operations.h.
-                    .5 mapop.c.
-                    .5 mapop.h.
+                    .5 \hr{operations.c}{src/interpreter/mappings/operations.c}.
+                    .5 \hr{operations.h}{src/interpreter/mappings/operations.h}.
+                    .5 \hr{mapop.c}{src/interpreter/mappings/mapop.c}.
+                    .5 \hr{mapop.h}{src/interpreter/mappings/mapop.h}.
                 .4 memory.
-                    .5 hashtable.c.
-                    .5 hashtable.h.
-                    .5 array.c.
-                    .5 array.h.
-                    .5 stack.c.
-                    .5 stack.h.
-                .4 parser.c.
-                .4 parser.h.
+                    .5 \hr{hashtable.c}{src/interpreter/memory/hashtable.c}.
+                    .5 \hr{hashtable.h}{src/interpreter/memory/hashtable.h}.
+                    .5 \hr{array.c}{src/interpreter/memory/array.c}.
+                    .5 \hr{array.h}{src/interpreter/memory/array.h}.
+                    .5 \hr{stack.c}{src/interpreter/memory/stack.c}.
+                    .5 \hr{stack.h}{src/interpreter/memory/stack.h}.
+                .4 \hr{parser.c}{src/interpreter/parser.c}.
+                .4 \hr{parser.h}{src/interpreter/parser.h}.
                 .4 processing.
-                    .5 implementations.c.
-                    .5 implementations.h.
-                .4 runtime.c.
-                .4 runtime.h.
-                .4 tokenizer.c.
-                .4 tokenizer.h.
+                    .5 \hr{implementations.c}{src/interpreter/processing/implementations.c}.
+                    .5 \hr{implementations.h}{src/interpreter/processing/implementations.h}.
+                .4 \hr{runtime.c}{src/interpreter/runtime.c}.
+                .4 \hr{runtime.h}{src/interpreter/runtime.h}.
+                .4 \hr{tokenizer.c}{src/interpreter/tokenizer.c}.
+                .4 \hr{tokenizer.h}{src/interpreter/tokenizer.h}.
     }
@@
 have a corresponding header file ending in .h. This is due to the nature
-of the compilation process of the C language\cte{c_compilation}, which is the host language
+of the compilation process of the C language, which is the host language
 the \name{} interpreter is written in.
@@
 The file mapop.gperf is also a source file. It is used in conjunction with the GNU gperf
-tool. The tool creates a perfect hash function\cte{hash}, which is output to the file
+tool. The tool creates a perfect hash function, which is output to the file
 src/interpreter/mappings/mapop.c. The perfect hash function is used to translate
@@
 interpreter internal representation in constant time.
+[\cte{c_compilation}, \cte{hash}]
 
@@
 writing and compiling C code. One such example is the article 'The C Beginner's
-Handbook' provided by freeCodeCamp.org linked in the sources\cte{freeCodeCamp}
+Handbook' provided by freeCodeCamp.org linked in the sources
 on page \pageref{bibliography}.
@@
 from the said included header is called with the selected input.
+[\cte{freeCodeCamp}]
 
@@
 The parser, implemented in src/interpreter/parser.c, generates a syntax-tree
-from tokens. The algorithm used is inspired by the Pratt Parsing\cte{pratt}
+from tokens. The algorithm used is inspired by the Pratt Parsing
 algorithm first described by Vaughan Pratt in 1973, which is a kind of precedence
 parser based on recursive descent. There is a great simple explanation of the algorithm
-by Jonathan Apodaca on dev.io\cte{devio} linked in the references on page 
+by Jonathan Apodaca on dev.io linked in the references on page 
 \pageref{bibliography} which this paper's parser takes inspiration from.
+[\cte{pratt}, \cte{devio}]
 
@@
 The algorithm used is inspired by Edsger Dijkstra's 'Shunting yard 
-algorithm'\cte{shunting_yard}, traversing the tree post-order to create
+algorithm', traversing the tree post-order to create
 a prefix notation. 
+[\cte{shunting_yard}]
 
@@
 
-\subsection{Limitations}
+\subsection{Limitations}\label{Limitations}
 \name{} is a programming language without a doubt. But it is not perfect in any
@@
 \subsubsection{Memory Leaks}
-Memory leaks happen, when a program reserves memory to store some data, but
+Memory leaks happen when a program reserves memory to store data, but
 never frees up the space for use for other programs, even though the data
-has long been discarded and is of no further use.
+has long been forgotten about and will no longer be accessed.
 In \name{}, memory leaks are a major problem. Even though barely noticable for small
@@
 Such tools can include a package-manager, with which people around the world
-can share code. Examples would be npm\cte{npm} or cargo\cte{cargo}, 
+can share code. Examples would be npm or cargo, 
 the 'Node Package Manager' for JavaScript applications and and Rust's package 
 manager respectively.
+[\cte{npm}, \cte{cargo}]
 
@@
 \subsubsection{Foreign Function Interfaces}
-A foreign function interface\cte{FFI} often abbreviated as FFI,
+A foreign function interface often abbreviated as FFI,
 is a mechanism by which one programming language can call functions written 
@@
 Java enables users to call C, C++ and even assembly code with the Java Native
-Interface\cte{JNI}, JNI for short, whereas within Python C function can
-be run with the standard ctypes\cte{ctypes} library.
+Interface, JNI for short, whereas within Python C function can
+be run with the standard ctypes library.
 
@@
 received from the function in the correct format, preventing a segmentation
-fault. The libffi\cte{libffi} C-library, which is also used by Java and
+fault. The libffi C-library, which is also used by Java and
 Python for this purpose, can be used to load and call such foreign functions.
+[\cte{FFI}, \cte{JNI}, \cte{ctypes}, \cte{libffi}]
 
 \subsubsection{Reflection}
-Reflection\cte{reflection}, also called reflective programming, describes
+Reflection, also called reflective programming, describes
 the ability of a program to examine and modify its own structure and behavior 
@@
 holding program instructions can be created and modified and later executed.
+[\cte{reflection}]
+
+\subsection{Conclution}
+Creating an interpreter was a challenge. From the beginning on, the project
+had to be planned and commented strictly, not to get lost in the complex web
+of source code later on. The final product is a success, a working interpreter
+hosting a custom programming language that can run human-readable programs.
+Although there are several limitations and shortcomings as stated in section \ref{Limitations}
+of this paper, the details of the implementation brought about unexpected 
+problems and solutions to profit from. The goal of this paper, to learn about 
+the inner workings of an interpreter, has therefore been achieved.
 
 \section{Index}
-\section{Closing Words}
-\section{Declaration of Autonomy}
-Hereby I, Florian Donnelly, declare to have written this paper, also including
-the entire source code of \name{} by myself using resources listed under the
-'references' section on page \pageref{bibliography}.
+This section contains a bibliography which links to all online articles
+this project could profit from, and the list of figures.
+The entire source code of \name{} can be found in the Appendix, section \ref{Appendix}.
 
-\newpage\section{Bibliography}
+\newpage\subsection{Bibliography}
+\renewcommand*{\bibfont}{\normalsize}
 \pagelabel{bibliography}\printbibliography[heading=none]
 
-\section{List of Figures}\listoffigures
+\subsection{List of Figures}\listoffigures
 \vspace{1cm}
@@
 Figure \ref{f_diagram} was created using an online diagram creation tool available
-on app.diagrams.net \cte{drawio}.
+on app.diagrams.net.
+[\cte{drawio}]
 
-The entire source code can be found in the Appendix, section \ref{Appendix}.
-
+\section{Declaration of Autonomy}
+Hereby I, Florian Donnelly, declare to have written this paper, also including
+the entire source code of \name{} by myself using resources listed under the
+'Bibliography' section on page \pageref{bibliography}.
 % TODO: check correctness of all file paths!! e.g. src/...

### 2022-10-23 paper.tex
continue writing paper

@@
 
+% hyperref abbreviations
+\newcommand{\hr}[2]{\hyperref[#2]{#1}}
+\newcommand{\hrc}[1]{\hyperref[#1]{#1}}
+
 % Read contents of files.txt into \files
@@
 % Include C code from file #1 from lines #2 to #3
-\newcommand{\code}[3] {
+\newcommand{\sourcecode}[3] {
     \ver{
@@
     \lstinputlisting[language=C,numbers=left,breaklines=true,
-        frame=single,title=#1,
+        frame=single,title=#1,firstnumber=#2,firstline=#2,lastline=#3]{#1}
+    }
+}
+
+\newcommand{\code}[3] {
+    \ver{
+
+    \begin{minipage}{\linewidth}
+
+    \lstset{basicstyle=\scriptsize,keywordstyle=\color{blue},stepnumber=5,
+        stringstyle=\color[rgb]{.3,.5,.1},commentstyle=\color{gray},
+        morecomment=[l][\color{magenta}]{\#},
+        postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space}}
+    \lstinputlisting[language=C,numbers=left,breaklines=true,
+        frame=single,title=\hrc{#1},
         firstnumber=#2,firstline=#2,lastline=#3]{#1}
+
+    \end{minipage}
     }
@@
 
+One more goal is to learn and understand the C programming language, in which
+the interpreter of \name{} will be written in.
+
 \section{Theory}
@@
         transformed into an AST or parse tree by a parser. A simple example
-        can be found under 
-        \hyperref[simple interpreter]{'Implementation/The Art of Abstraction'}
-        on page \pageref{simple interpreter}. Compared to bytecode interpreters
+        can be found in section \ref{simple_interpreter}
+        on page \pageref{simple_interpreter}. Compared to bytecode interpreters
         there is a large time overhead when performing syntax related computation
@@
 popular programming languages:
-\begin{enumerate}\pagelabel{simple interpreter}
+\begin{enumerate}\pagelabel{simple_interpreter}
     \item To create meaningful groups of characters from the input string. This
@@
 are of no further importance regarding \name are depicted.
-\newcommand{\hr}[2]{\hyperref[#2]{#1}}
 \ver{\begin{flushleft}
@@
 
+% TODO: say sth about README.md and expand it
+
 The file mapop.gperf is also a source file. It is used in conjunction with the GNU gperf
@@
 interpreter internal representation in constant time.
-[\cte{c_compilation}, \cte{hash}]
+[\cte{c_compilation}, \cte{gperf}]
 
@@
 \name{} implementation interact and work together. Each next stage performs
-operations on the output data of the proper. The src/interpreter/interpreter.c
+operations on the output data of the proper. The \hrc{src/interpreter/interpreter.c}
 file, here depicted as 'Interpreter', chains all stages together.
@@
 \subsubsection{The Entry-Point}
-The entry point of the project lies in the src/main.c file.
+The entry point of the project lies in the \hrc{src/main.c} file.
 The file implements the C main function, which is the entry-point of any
-C application in general. Since the file is so small, here is its source code:
+C application. Since the file is so small, here is its source code:
 \code{src/main.c}{1}{999}
@@
 For the readers who do not know how to read source code, C in particular, here is a short
-explanation of the basic syntax of the C language based on the src/main.c file:
+explanation of the basic syntax of the C language taking the \hrc{src/main.c} file
+as an example:
 \begin{itemize}
@@
         The most common preprocessor instructions are the following.
-        On one hand there is \expr{\#include <header.h>}
+        On one hand there is \expr{\#include $<$header.h$>$}
         to include various definitions from a header file, i.e. to paste
@@
 
-    \item \emph{Functions.} Functions in C are defined the following way:
-        \expr{<return type> <function name>([arguments]) { <function body> }}
-        or with a concrete, working example:
-        \expr{int myFunction(int *arg) \{ printf("arg: \%p, arg); \}.}
-        In main.c there is one function. The function's name is 'main', its
-        return type is 'int', it has two arguments 'argc' and 'argv' and a
-        function body containing what the function does.
+    \item \emph{Functions.} Every C function has a return type, a
+        function name, arguments and a function body in that order.
+        Here is an example of a C function:
+        \code{pseudocode/function.c}{1}{999}
+        This would be an example of a main function, similar to the one
+        in \hrc{src/main.c}. It returns a value
+        of type 'int', has two arguments, 'argc' and 'argv', and a function body
+        which, in the case of the example above, prints the value of the 'argc' argument.
 
     \item \emph{Pointers.} Pointers are variables that store memory addresses
-        of data. In C-like syntax they are defined with the star (*) symbol.
-        An example from src/main.c is 'FILE *input;'. That means that input is not
-        actually a FILE, but rather the memory address to where a FILE is.
-
-        To dereference the pointer, i.e. to actually retrieve the value stored
-        at the memory address it contains, another star could be used as such:
+        of data. In C-like syntax they are defined with the star '*' symbol.
+        An example from \hrc{src/main.c} is 'FILE *input;'. That means that input is not
+        actually a FILE, but rather a memory address pointing to a spot in memory where a FILE is.
+        To dereference the pointer, i.e. to copy the value stored
+        at the memory address into a variable on the stack, another star is used like this:
         'FILE value = *input;'.
@@
 For those who are further interested in the C programming language, there are
-many internet articles or tutorials about reading and understanding or
+countless internet articles or tutorials about reading and understanding or
 writing and compiling C code. One such example is the article 'The C Beginner's
@@
 
-Here is a very quick summary of what happens in the src/main.c file: the src/interpreter/interpreter.h
-header file is included. In the main function, the program decides to either
-read a program from the standard input or from a file. Then, the interpret function
+Following is a very quick summary of the few 
+things happening in the \hrc{src/main.c} file.
+First, the \hrc{src/interpreter/interpreter.h} header file is included. 
+The main function decides whether to
+read a program from the standard input or from a file. 
+Then, the interpret function
 from the said included header is called with the selected input.
@@
 \subsubsection{The Interpreter}
-This section deals with the file src/interpreter/interpreter.c which guides
-input data through the various stages of the \name{} interpreter in its
-function 'interpret', which is called from the program's main function.
+This section deals with the file 
+\hrc{src/interpreter/interpreter.c} which guides
+input data through the various stages of the \name{} interpreter as 
+seen in figure \ref{f_diagram} on page \pageref{f_diagram}.
+Following is a breakdown of the 'interpret' function from said file.
 
 The 'interpret' function calls the parser to generate a syntax-tree from tokens retrieved
-from the tokenizer implemented in src/interpreter/parser.c and 
-src/interpreter/tokenizer.c respectively.
+from the tokenizer implemented in \hrc{src/interpreter/parser.c} and 
+\hrc{src/interpreter/tokenizer.c} respectively.
 
@@
 complexity of O(log n), n being the amount of different variables.
+The hashtable implementation used in \hrc{src/interpreter/memory/hashtable.c}
+uses binary search to locate elements variable-sized arrays in the hashtable's buckets.
+For collision detection, i.e. to circumvent problems arising when two keys have the
+same hash, separate chaining is used. In this approach, linked lists with
+key-value pairs are used for each array index and keys are compared directly
+to find exact matches.
+[\cte{separate_chaining}]
 
 The syntax tree could now be interpreted directly, but as said, is instead converted
-to an intermediate internal bytecode format by src/interpreter/bytecode.c.
+to an intermediate internal bytecode format by \hrc{src/interpreter/bytecode.c}.
 This step is done so that code that is run multiple times does not convey
@@
 The final step is to process or execute said bytecode. This step is performed
-by src/interpreter/runtime.c.
+by \hrc{src/interpreter/runtime.c}.
 
 \subsubsection{The Tokenizer}
-The tokenizer is implemented in src/interpreter/tokenizer.c. Its job is to
+The tokenizer is implemented in \hrc{src/interpreter/tokenizer.c}. Its job is to
 take source code in plain-text format as input and output so-called tokens.
@@
 The different types of tokens and other things are defined in the header file
-src/interpreter/tokenizer.h.
+\hrc{src/interpreter/tokenizer.h}.
 
@@
 \end{enumerate}
-FIELD type tokens are later replaced by REFERENCE tokens during the stage of localization
-and SYMBOL type tokens are mapped to EXPR by src/implementation/mappings/mapop.c
+FIELD type tokens are later replaced by REFERENCE tokens, effectively mapping
+variable names to addresses in memory.
+SYMBOL type tokens are mapped to operators such as plus or minux by \hrc{src/implementation/mappings/mapop.c}
 during parsing. NUMBER type tokens are replaced by an actual number type, which
-is C's 'long double' type as specified in src/interpreter/bytecode.h.
+is C's 'long double' type as specified in \hrc{src/interpreter/bytecode.h}.
 
@@
 \subsubsection{The Parser}
-The parser, implemented in src/interpreter/parser.c, generates a syntax-tree
+The parser, implemented in \hrc{src/interpreter/parser.c}, generates a syntax-tree
 from tokens. The algorithm used is inspired by the Pratt Parsing
 algorithm first described by Vaughan Pratt in 1973, which is a kind of precedence
-parser based on recursive descent. There is a great simple explanation of the algorithm
-by Jonathan Apodaca on dev.io linked in the references on page 
+parser based on recursive descent. There is an explanation of the algorithm
+written by Jonathan Apodaca on dev.io linked in the references on page 
 \pageref{bibliography} which this paper's parser takes inspiration from.
@@
 The syntax tree in the implementation of \name{} consists of 'stnode' type
-pointers, as defined in src/interpreter/parser.h. Each such node depending
-on its 'nodetype', declared in src/interpreter/tokenizer.h,
-is either a value-node storing a pointer to some data, e.g. a number or text, or a
+pointers defined in \hrc{src/interpreter/parser.h}. Each such node depending
+on its 'nodetype', declared in \hrc{src/interpreter/tokenizer.h},
+is either a value-node storing a pointer to data, e.g. a number or text, or a
 parent-node referencing two child nodes and an operator, e.g. two numbers as
@@
 
-The difficult part of implementing a parser is precedence. Precedence and
-associativity define the order of execution of statements. A simple example
-would be to perform multiplication before addition, e.g. '2+3*4' would
-result in 14 and is not the same as '(2+3)*4'. Therefore, each operator is
-assigned a precedence, or priority, as well as an associativity, which is
-either 'Left to Right' or 'Right to Left'. An example of associativity is
-\emph{'20 / 10 / 2'}. Both in C as well as in \name{}, the division operator is
+The difficult part of implementing a parser is precedence. Precedence together
+with associativity define the order of execution of statements. A simple example
+would be to perform multiplication before addition, e.g. \emph{2+3*4} would
+result in \emph{14} and is not the same as \emph{(2+3)*4}.
+Assosiativity is either 'Left to Right' or 'Right to Left'. An example of associativity is
+\emph{20/10/2}. Both in C as well as in \name{}, the division operator is
 left-to-right associative, meaning the stated expression is the same as
-\emph{'(20 / 10) / 2'} and not as \emph{'20 / (10 / 2)'}. Both precedence 
-and associativity are properties that are defined in the file mapop.gpref for
+\emph{(20/10)/2} and not as \emph{20/(10/2)}. Both precedence 
+and associativity are properties that are defined in the file \hrc{mapop.gpref} for
 each operator.
 
-The parser implementation in src/interpreter/parser.c may be difficult to
+The parser implementation in \hrc{src/interpreter/parser.c} may be difficult to
 read and understand, which is why the most important functions are explained here:
@@
 \subsubsection{The Bytecode Compiler}
-The bytecode compiler is part of the interpreter. Its job is to create a list
+The bytecode compiler is an internal part of the interpreter, located in the file
+\hrc{src/interpreter/bytecode.c}. Its job is to create a list
 of byte-sized instructions and data from the syntax-tree the parser created.
@@
 
-In src/interpreter/bytecode.c we can see, that first, the left operand and then
+In \hrc{src/interpreter/bytecode.c} we can see that first the left operand and then
 the right operand are added onto the stack of bytecode. Therefore, when
@@
     \item In common prefix notation, also called Polish notation, the
-        operators would now precede the operands as such:
+        operators would now precede the operands in this manner:
         \expr{\emph{'- 1 (/ (* 2 4) (+ 5 6))'}}
@@
 \end{enumerate}
+[\cte{polish_notation}]
 
@@
 All available expressions are implemented in 
-src/interpreter/processing/implementations.c.
+\hrc{src/interpreter/processing/implementations.c}.
 
@@
         operands must follow.
-    \item The processing function will call itsself
-        and notice the 6, a number, and therefore a value. The function simply returns the 6
-        to the subtraction expression. This 6 will be the right-hand operator
+    \item The processing function will call itself
+        and notice the 6, a number, and therefore a value. The function simply returns the six
+        to the subtraction expression. This six will be the right-hand operator
         of the subtraction.
@@
 
-That is it. With all the expressions available, entire programs
-can now be interpreted.
+That is it. With all expressions that are currently implemented, entire programs
+can be interpreted. Examples of such programs can be found in the \emph{examples} directory
+of the source code.
 
@@
 \name{} is built upon the implementation of a recursive Pratt
-parser and a basic tokenizer. Most actual features are reliant upon this
-foundation, such as operators and the standard library built into \name{}.
+parser and a basic tokenizer without which the entire \name{} implementation
+would be unusable.
 
 \subsection{Syntax}
-The syntax of \name{} can be compared to the syntax of JavaScript, except
-there are no keywords at all, and everything is controlled via operators,
+The syntax of \name{} is unlike most modern, common programming languages' syntax,
+but can be compared somone to that of JavaScript.
+In \name{}, there are no keywords at all and everything is controlled via operators,
 which are non-alphanumeric characters. Such operators include the plus-sign 
@@
 \subsection{Limitations}\label{Limitations}
-\name{} is a programming language without a doubt. But it is not perfect in any
+\name{} can be called a programming language without a doubt. But it is not perfect in any
 way. There are limitations that make it practically unusable for actual
@@
 Figure \ref{f_mandelbrot} was generated using a \name{} program, the source code
-can be found in the file examples/mandelbrot.nts.
+can be found in the file \hrc{examples/mandelbrot.nts}.
 Figure \ref{f_diagram} was created using an online diagram creation tool available
-on app.diagrams.net.
+on \url{https://app.diagrams.net}.
 [\cte{drawio}]
@@
 \ver{
-\foreach \f in \files{\pagelabel{\f}\code{\f}{1}{9999}}
+\foreach \f in \files{\pagelabel{\f}\sourcecode{\f}{1}{9999}}
 }

### 2022-10-23 pseudocode/function.c

@@ -1,0 +1,4 @@
+#include <stdio.h>
+int main(int argc, char *argv[]) {
+    printf("argc: %i\n", argc); 
+}

### 2022-10-25 src/interpreter/libraries/stdlib/stdlib.c
fix std.floor,std.ceil

@@ -162,7 +162,7 @@ void _floor(mementry *args, mementry *dst, errorinfo *info) {
         return;
     }
     dst->type = NUMBER;
-    number res = ceill(*(number *)args->value);
+    number res = floorl(*(number *)args->value);
     *((number *) (dst->value = malloc(sizeof(number)))) = res;
     return;
 }
@@ -173,7 +173,7 @@ void _ceil(mementry *args, mementry *dst, errorinfo *info) {
         return;
     }
     dst->type = NUMBER;
-    number res = floorl(*(number *)args->value);
+    number res = ceill(*(number *)args->value);
     *((number *) (dst->value = malloc(sizeof(number)))) = res;
     return;
 }

### 2022-10-25 paper.tex
continue writing paper

@@
 web front-end to store and collaborate on source code.
+% TODO: add gdb!
 
@@
 of byte-sized instructions and data from the syntax-tree the parser created.
-Having such an intermediary bytecode format, instead of directly interpreting
+Using such an intermediary bytecode format instead of directly interpreting
 the parse-tree can be beneficial. This is the case when the same piece of code
@@
 having to walk along tree-nodes recursively, only the list of instructions in
-the bytecode can be traversed, which is much faster.
+the bytecode can be traversed, which takes less time overall.
 
 Not to be confused with machine-native bytecode, the bytecode compiler in \name{}
-generates byte-sequences in an internal format, that are then interpreted. A
+generates byte-sequences in an internal format that are then interpreted. A
 JIT compiler would instead compile the AST down to machine-native bytecode, which
@@
 
-To generate such a sequence of bytes, \name{} recursively reads through the
-binary tree created by the parser and pushes found instructions onto a stack,
-which is in this case used similarly to a variable-sized list.
+To generate the bytecode, \name{} recursively reads through the
+binary tree created by the parser and pushes found instructions onto a stack.
 The algorithm used is inspired by Edsger Dijkstra's 'Shunting yard 
@@
     \item Assume the following mathematical expression in infix notation:
-        \expr{\emph{'1 - 2 * 4 / (5 + 6)'}}
+        \expr{\emph{1 - 2 * 4 / (5 + 6)}}
     \item In common prefix notation, also called Polish notation, the
         operators would now precede the operands in this manner:
-        \expr{\emph{'- 1 (/ (* 2 4) (+ 5 6))'}}
+        \expr{\emph{- 1 (/ (* 2 4) (+ 5 6))}}
     \item But in \name{} the operands are pushed onto the stack is reverse order.
         This would look like the following:
-        \expr{\emph{'- (/ (+ 6 5) (* 4 2)) 1'}}
+        \expr{\emph{- (/ (+ 6 5) (* 4 2)) 1}}
     \item The brackets are not needed, because for every operator, the amount
@@
         and the negative operator, which are both a minus sign, are distinguished
-        beforehand. This means the above statement would look like this:
-        \expr{\emph{'- / + 6 5 * 4 2 1'}}
-        This is an ideal abstraction of how data 
-        is represented internally in \name{}.
+        beforehand. The above statement would therefore look like this:
+        \expr{\emph{- / + 6 5 * 4 2 1}}
+        This is how expressions are represented in the bytecode of \name{}.
 \end{enumerate}
@@
 compiler and evaluates expressions by plugging in the provided values.
-All available expressions are implemented in 
+All available operators are implemented in 
 \hrc{src/interpreter/processing/implementations.c}.
@@
 The evaluation of expressions also works in a recursive manner. Here is an 
-example of \name{}'s bytecode notation and how it would be evaluated at runtime:
+example of \name{}'s bytecode notation and how it is evaluated at runtime:
 \begin{itemize}
     \item Assume the input in \name{}'s bytecode to resemble the following expression: 
-        \expr{\emph{'- 6 * 3 4'}}
-        In infix notation, this would be equivalent to 
-        \expr{\emph{'(4 * 3) - 6'}.}
+        \expr{\emph{- 6 * 3 4}}
+        In common infix notation, this would be equivalent to 
+        \expr{\emph{(4 * 3) - 6}.}
+        The question arises, how can an algorithm compute the result of this
+        calculation? The answer is recursion.
     \item The runtime will read through the bytecode left-to-right. First it will
@@
         operands must follow.
-    \item The processing function will call itself
-        and notice the 6, a number, and therefore a value. The function simply returns the six
-        to the subtraction expression. This six will be the right-hand operator
-        of the subtraction.
-    \item The subtraction will then call the processing function again to find
-        the left operand. As the processing function reads a multiplication operator,
-        it now recursively reads through that expression's right and then left
-        operator, which would be 3 and for respectively and compute and return
-        the result of the multiplication, which is 12.
-    \item Now the subtraction knows its right- and left-hand operators, 6 and 12,
-        and again, calculate and return the result, which is 6.
+    \item The function recursively calls itself in order to evaluate the
+        right operand and notice the 6, a value. The function simply returns the six
+        to the subtraction expression.
+    \item Again, the function calls itself to find the next operand.
+        As it now reaches a multiplication operator,
+        it recursively reads through that sub-expression's right and then left
+        operator, which are 3 and 4 respectively.
+        The result is computed and returned.
+    \item Now the subtraction knows its right- and left-hand operators, 6 and 12.
+        The subtraction is performed and the result, 6, is returned.
 \end{itemize}
 
-That is it. With all expressions that are currently implemented, entire programs
+That is it. With all the operators and standard functions that have been implemented, entire programs
 can be interpreted. Examples of such programs can be found in the \emph{examples} directory
@@
 \name{} is built upon the implementation of a recursive Pratt
-parser and a basic tokenizer without which the entire \name{} implementation
-would be unusable.
+parser and a basic tokenizer. It uses a hash table to map variable names to their
+memory addresses. There are no keywords, only a set of non-alphanumeric characters
+representing all possible operations. Extending upon that, there are some
+standard functions.
 
@@
 \subsection{Name Shadowing}
-Here is another example, which in contrast to the example above, focuses on a
-security concept implemented in \name{}.
+Here is another example, which in contrast to the example above, focuses on an
+original security concept implemented in \name{}.
 \code{examples/exec.nts}{1}{999}
-The code begins by creating the variable \emph{var}, which holds a string, and
+The code begins by creating the variable \emph{var}, which holds a string and
 \emph{f}, which calls the function \emph{std.exec} with var as argument.
@@
 \subsection{Limitations}\label{Limitations}
-\name{} can be called a programming language without a doubt. But it is not perfect in any
-way. There are limitations that make it practically unusable for actual
+Unfortunately, there are limitations that make \name{} practically unusable for actual
 professional projects. In this section, a few shortcomings of \name{} are listed
-and reasoned over.
+and discussed.
 
@@
 use Garbage-Collection, which automatically tracks unused allocated memory and
-frees it up automaticadlly. While this is also the case with \name{}, there are
+frees it up automatically. While this is also the case with \name{}, there are
 many internal flaws that have nothing to do with variables used in a \name{}
@@
 
-A major reason for the memory leaks, but not an excuse, is the nature of the
-C programming language the \name{} interpreter was written in. The C language and
-compiler do not keep track of states in which certain blocks of memory are
-forgotten about and warn the developer, instead, memory leaks happen.
-More modern C alternatives such as Rust have a ownership-system in place,
-in which any allocated memory has a function as an owner, and when the function
-is done processing, the memory is automatically freed. Such systems are a
-helpful tool for developers to create better programs with low mental overhead,
-and are what programming language design should be all about.
+One reason memory leaks is the nature of the
+C programming language, in which the \name{} interpreter was written in. The C language and
+compiler do not keep track of all possible states in which certain blocks of memory are
+forgotten about and do not warn the developer. Instead, memory leaks happen.
+More modern C alternatives such as Rust have an ownership-system in place,
+in which any allocated memory is owned by a function. When the function
+is done processing, the memory is automatically freed. Such a system
+is a helpful tool for developers to create better programs without having to get
+lost in the details.
+Inventing new ways to make the job of developers easier is what programming
+language design is all about.
 
 \subsubsection{Single-Threaded}
-The \name{} interpreter does not implement any multi-threaded, meaning parallel
-running, programs as most other languages do. Instead, any \name{} program can
-only be interpreted and run sequentially. Since most modern computer processors
+The \name{} interpreter does not implement any multi-threaded, i.e. in parallel
+running programs as most other languages do. Instead, any \name{} program can
+only run sequentially. Since most modern computer processors
 are equipped with a handful of processing cores and threads, not being able to
-make use of those is a major downside and in itsself a reason not to use this
-language.
+make use of this modern time's advantage is a major downside.
 
 \subsubsection{Lack of a Development Environment}
-This is certainly the greatest problem of \name{}, aside the countless smaller
-bothering aspects.
+This is certainly the greatest problem of \name{}.
 
-Most modern programming languages do not only consist
+Most modern programming languages not only consist
 of a language specification and a compiler or interpreter, but of many more
-tools surrounding them.
+tools that surround themn.
 Such tools can include a package-manager, with which people around the world
@@
 manager respectively.
+Another example of a helpful tool is a debugger, helping developers
+find the cause of errors, so-called bugs, in their programs.
+This includes the tool gdb, the GNU Project Debugger, that was used in this
+project to find errors in the \name{} implementation.
 [\cte{npm}, \cte{cargo}]
@@
 
-\subsection{Conclution}
+\subsection{Conclusion}
 Creating an interpreter was a challenge. From the beginning on, the project
@@
 hosting a custom programming language that can run human-readable programs.
+Most parts of the source code are annotated with comments, giving interested
+reader direct explanations of what is going on.
 Although there are several limitations and shortcomings as stated in section \ref{Limitations}
 of this paper, the details of the implementation brought about unexpected 
-problems and solutions to profit from. The goal of this paper, to learn about 
-the inner workings of an interpreter, has therefore been achieved.
+problems and solutions. The goal of this paper to learn about 
+the inner workings of an interpreter has been achieved, as well as the goal
+to get to know the C language.
 

### 2022-10-26 src/interpreter/bytecode.c
change err msg, fix bug with scopes and memfree

@@ -19,7 +19,7 @@ static inline mementry *_register(bytecode *code, stackptr ptr, typing type) {
 table_level max_level;
 
 // see protocol specifications in header file
-void _recursiveconsume(bytecode *dst, stackptr ptr, stnode *subtree) {
+void _recursiveconsume(bytecode *dst, stackptr ptr, stnode *subtree, char member_as_block) {
     switch(subtree->type){
 
         case NUMBER:
@@ -66,11 +66,11 @@ void _recursiveconsume(bytecode *dst, stackptr ptr, stnode *subtree) {
 
         case EXPR:
             // process left child first
-            _recursiveconsume(dst, ptr, subtree->data.parent.left);
+            _recursiveconsume(dst, ptr, subtree->data.parent.left, 0);
 
             // if there is one, process right child next
             if(subtree->data.parent.right != NULL)
-                _recursiveconsume(dst, ptr, subtree->data.parent.right);
+                _recursiveconsume(dst, ptr, subtree->data.parent.right, 0);
             // make sure to provide a right-hand argument if it's a function call
             else if(subtree->data.parent.op->opcode == OP_CALL)
                 _register(dst, ptr, UNDEFINED)->value = NULL;
@@ -81,6 +81,22 @@ void _recursiveconsume(bytecode *dst, stackptr ptr, stnode *subtree) {
             break;
 
         case MEMBER: {
+            stnode *tree;
+            if(member_as_block){
+                _register(dst, ptr, EXPR)->value = _block_end;
+                while(subtree != NULL && subtree->data.parent.left != NULL) {
+                    _recursiveconsume(dst, ptr, subtree->data.parent.left, 1);
+
+                    // Reassign values for the loop
+                    tree = subtree;
+                    subtree = subtree->data.parent.right;
+
+                    // Free the MEMBER node
+                    free(tree);
+                }
+                _register(dst, ptr, EXPR)->value = _block;
+                break;
+            }
             // Create a new stack for the underlying function
             bytecode *function = create_stack(64);
 
@@ -90,10 +106,9 @@ void _recursiveconsume(bytecode *dst, stackptr ptr, stnode *subtree) {
             max_level = 0;
 
             // Consume a member chain into the stack
-            stnode *tree;
             while(subtree != NULL && subtree->data.parent.left != NULL) {
                 // Consume the left subtree
-                _recursiveconsume(function, (stackptr) &offset, subtree->data.parent.left);
[…]

### 2022-10-26 src/interpreter/error/error.c

@@ -22,8 +22,7 @@ static const char *errformats[] = {
     "The specified operator cannot be used on given type(s)",
     "The specified operand is not a function, therefore it cannot be called",
     "The specified operand must be mutable",
-    "You can only assign the value from one field to another, if the source "
-        "field was created in a scope equal or lower than the destination field",
+    "Assignment could not be done because used variables will not exist anymore",
     "Index out of bounds",
 
     "Error during execution of if-statement",

### 2022-10-26 src/interpreter/localizer.c

@@ -33,12 +33,18 @@ void localize(stnode *subtree, hashtable *table, errorinfo *info) {
         case MEMBER:
 
             if(subtree->data.parent.left != NULL) {
+                // change ht level
+                ht_up(table);
+
                 // localize child first
                 if(subtree->data.parent.right != NULL)
                     _localize_member(subtree->data.parent.right, table, info);
 
                 // localize current expression
                 localize(subtree->data.parent.left, table, info);
+                
+                // reset ht level
+                ht_down(table);
             }
 
             break;

### 2022-10-26 src/interpreter/processing/implementations.c

@@ -34,7 +34,7 @@ char truth_of(mementry *entry) {
 // free a synthetic mementry and its synthetic value
 void _free_synth(mementry *entry) {
     if(entry->flags.synthetic) {
-        if(entry->flags.value_synthetic)
+        if(entry->flags.value_synthetic && entry->type != UNDEFINED)
             free(entry->value);
         free(entry);
     }
@@ -546,6 +546,11 @@ mementry *call(mementry *fun, mementry *params, errorinfo *info){
             _free_synth(fun);
             return NULL;
     }
+
+
+    // free arguments
+    if(dst != fun)
+        _free_synth(fun);
 
     return dst;
 }

### 2022-10-28 src/interpreter/libraries/stdlib/stdlib.c
fix allocation bug, make montecarlo work

@@ -71,7 +71,7 @@ void _exec(mementry *args, mementry *dst, errorinfo *info) {
 void _if(mementry *args, mementry *dst, errorinfo *info) {
     if(args->type == TUPLE){
         array *arr = ((array *)args->value);
-        mementry *fun = NULL;
+        mementry *ret = NULL;
         // can be (condition, if) or (condition, if, else)
         if(arr->size == 2 || arr->size == 3) {
             mementry *cond = arr->arr[0];
@@ -81,21 +81,21 @@ void _if(mementry *args, mementry *dst, errorinfo *info) {
             }
             if(truth_of(cond)) {
                 // perform if case
-                fun = arr->arr[1];
+                ret = arr->arr[1];
             } else if(arr->size == 3) {
                 // perform else case 
-                fun = arr->arr[2];
+                ret = arr->arr[2];
             }
-            if(fun == NULL) {
+            if(ret == NULL) {
                 dst->type = UNDEFINED; 
             } else {
-                dst->type = fun->type;
-                dst->level = fun->level;
-                dst->value = fun->value;
+                dst->type = ret->type;
+                dst->level = ret->level;
+                dst->value = ret->value;
 
                 // free fun if synthetic, but not value
-                if(fun->flags.synthetic)
-                    free(fun);
+                if(ret->flags.synthetic)
+                    free(ret);
             }
             return;
         }

### 2022-10-28 src/interpreter/memory/hashtable.c

@@ -18,6 +18,7 @@ static inline _H_HASH _hash(char* key) {
 // return new mementry with default values
 mementry *_alloc_mementry() {
     mementry *ret = malloc(sizeof(mementry));
+    ret->value = NULL;
     ret->flags =
         (struct s_mementry_flags) 
         {.mutable = 0, .synthetic = 0, .value_synthetic = 0};

### 2022-10-28 src/interpreter/processing/implementations.c

@@ -530,8 +530,6 @@ mementry *call(mementry *fun, mementry *params, errorinfo *info){
             else {
                 // type and value will be set by the called function
                 dst = _alloc_mementry();
-                dst->flags = (struct s_mementry_flags) 
-                    {.mutable=0, .synthetic=1, .value_synthetic=1};
             }
 
             // call function from shared object

### 2022-10-29 paper.tex
continue writing paper

@@
 But computers are not just the metals and semiconductors they are built from,
-but would be unusable without any programs running on them.
-Creating such programs, no matter if games, web browsers or even operating systems,
+but would be unusable without any software running on them.
+Creating programs, no matter if games, web browsers or even operating systems,
 involve the use of programming languages.
@@
 to writing machine-native code, i.e. not using a programming language.
+By making use of just the term \emph{language} this paper will refer to programming
+languages specifically. Sources are mentioned fater cohesive paragraphs in
+square brackets, an index can be found in the \emph{Bibliography} on page
+\pageref{bibliography}.
 
@@
 help with built-in git integration and auto-completion features.
-\emph{Git} is the project management and tracking software used to back up and share
+The tools \emph{Valgrind} and \emph{gdb}, the GNU Project Debugger were of great help
+regarding the finding of errors and memory leaks, see section \ref{memleaks} for details.
+\emph{Git} is the project management and version tracking software used to back up and share
 the project files on \emph{GitHub.com}, a Microsoft hosted git repository and
 web front-end to store and collaborate on source code.
-% TODO: add gdb!
 
@@
     \item To create meaningful groups of characters from the input string. This
-        process is called 'tokenization' and is done with the component called
-        tokenizer or lexer. As an example, the string 'aa + bb' could result
-        in three tokens, namely 'aa', '+' and 'bb'. In this case, space characters
-        are ignored. 
+        process is known as tokenization and is done with the component called
+        tokenizer or lexer. As an example, the string \emph{ab + cd} could result
+        in three tokens, namely \emph{ab}, \emph{+} and \emph{cd}. In this case, space characters
+        are simply ignored and could be left out, unlike in the English language where spaces are relevant. 
         This is only an example and the grouping of characters could be done 
@@
         Preprocessing is what the compiler does before actually compiling the
-        code.
+        code. %TODO: add src (Preprocessor)
 
@@
 
-    \item \emph{Functions.} Every C function has a return type, a
+    \item \emph{Functions.} In programming, functions or methods are
+        a set of code that accomplish a certain task. They take in data,
+        process it and return a result. Calling a function means to evaluate it for a given input.
+        In C, every function has a return type, a
         function name, arguments and a function body in that order.
@@
 
-    \item \emph{Pointers.} Pointers are variables that store memory addresses
-        of data. In C-like syntax they are defined with the star '*' symbol.
-        An example from \hrc{src/main.c} is 'FILE *input;'. That means that input is not
+    \item \emph{Pointers.} Pointers are variables that store memory addresses. 
+        In C syntax they can be defined with the star '*' symbol.
+        An example from \hrc{src/main.c} is \expr{FILE *input;.} That means that \emph{input} is not
         actually a FILE, but rather a memory address pointing to a spot in memory where a FILE is.
         To dereference the pointer, i.e. to copy the value stored
-        at the memory address into a variable on the stack, another star is used like this:
-        'FILE value = *input;'.
+        at the memory address into another variable, another star is used like this:
+        \expr{FILE value = *input;}.
+        \emph{value} now contains a copy of the actual FILE that \emph{input} was pointing to.
 \end{itemize}
@@
 on page \pageref{bibliography}.
+[\cte{freeCodeCamp}]
 
@@
 First, the \hrc{src/interpreter/interpreter.h} header file is included. 
-The main function decides whether to
-read a program from the standard input or from a file. 
-Then, the interpret function
+The main function then decides whether to
+read a program from a file or from standard input. 
+At last, the interpret function
 from the said included header is called with the selected input.
-[\cte{freeCodeCamp}]
 
@@
 seen in figure \ref{f_diagram} on page \pageref{f_diagram}.
-Following is a breakdown of the 'interpret' function from said file.
+Following is a breakdown of the \emph{interpret} function from said file
+as well as an overview of some of the stages that follow.
 
-The 'interpret' function calls the parser to generate a syntax-tree from tokens retrieved
+The \emph{interpret} function calls the parser to generate a syntax-tree from tokens retrieved
 from the tokenizer implemented in \hrc{src/interpreter/parser.c} and 
@@
 
-After that, variables with names are 'localized', meaning their names are
+After that, the names of variables are
 mapped to addresses in memory where their value is stored. Doing this before
 bytecode compilation has a performance benefit. 
-The mapping is done with a custom hashtable datastructure with average time
-complexity of O(log n), n being the amount of different variables.
-The hashtable implementation used in \hrc{src/interpreter/memory/hashtable.c}
+The mapping is acheived with a custom hashtable datastructure with average time
+complexity of O(log n), n being the total amount of variables a \name{} program is using.
+The hashtable implementation used in \hrc{src/interpreter/memory/hashtable.c}                   % TODO: maybe its own paragraph! interpolation search vs binary search
 uses binary search to locate elements variable-sized arrays in the hashtable's buckets.
-For collision detection, i.e. to circumvent problems arising when two keys have the
-same hash, separate chaining is used. In this approach, linked lists with
-key-value pairs are used for each array index and keys are compared directly
+This is because for such few variables that are used in a standard \name{} program,
+an interpolation search algorithm is actually slower.
+For collision detection, i.e. to circumvent problems arising when two variable names have the
+same hash, a technique called separate chaining is used. In this approach, linked lists with
+key-value pairs are used for each array index of the buckets and variable names are compared directly
 to find exact matches.
@@
 
-The syntax tree could now be interpreted directly, but as said, is instead converted
+The syntax tree from the parser could now be interpreted directly, but as mentioned 
+before, is instead converted
 to an intermediate internal bytecode format by \hrc{src/interpreter/bytecode.c}.
-This step is done so that code that is run multiple times does not convey
-the problem of time overhead when traversing the syntax tree, but the current
-implementation does not take advantage of this.
+This step is done so that code that is run multiple times does not have
+the problem of time overhead when traversing the syntax tree.
 
 The final step is to process or execute said bytecode. This step is performed
-by \hrc{src/interpreter/runtime.c}.
+in \hrc{src/interpreter/runtime.c}.
 
@@
 Here is an example of tokenization:
-An input of 'std.test==23.4' would result in the following tokens according
+An input of \emph{a.b==23.4} would result in the following tokens according
 to the tokenizer implementation of this paper:
 \begin{enumerate}
-    \item Token of type FIELD with content 'std.text'.
-    \item Token of type SYMBOL with content '=='.
-    \item Token of type NUMBER with content '23.4'.
+    \item Token of type FIELD with content \emph{a.b}.
+    \item Token of type SYMBOL with content \emph{==}.
+    \item Token of type NUMBER with content \emph{23.4}.
 \end{enumerate}
@@
 
-The tokenizer used in this paper is a simple finite-state machine, reading
-character by character and determining the token type as it goes.
-
-These tokens are then directly consumed by the parser as described in the next
+While tokenizers can be programmed using regular expressions, 
+this paper uses a more simple finite-state machine tokenizer.
+It reads character by character, determining the token type as it goes.
+It stops when there is a token thad does no longer match the current token type.
+Tokens are each requested and integrated into a syntax-tree by the parser as described in the next
 section.
@@
 \subsubsection{The Parser}
-The parser, implemented in \hrc{src/interpreter/parser.c}, generates a syntax-tree
-from tokens. The algorithm used is inspired by the Pratt Parsing
-algorithm first described by Vaughan Pratt in 1973, which is a kind of precedence
+The parser, implemented in \hrc{src/interpreter/parser.c}, generates a syntax tree,
+also called parse tree, from tokens. Since this paper's implementation 
+does not parse code from a
+concise context free grammar, the syntax tree the parser creates is very
+similar to an otherwise known abstract syntax tree, AST for short.
+ASTs describe source code conceptually and do not contain all syntactical
+elements required to parse code.
+In this paper the terms parse tree, syntax tree and AST will be used
+synonymously.
+
+The algorithm used is inspired by the 'Pratt Parsing
+algorithm' first described by Vaughan Pratt in 1973, which is a kind of precedence
 parser based on recursive descent. There is an explanation of the algorithm
-written by Jonathan Apodaca on dev.io linked in the references on page 
+written by Jonathan Apodaca on \emph{dev.io} linked in the references on page 
 \pageref{bibliography} which this paper's parser takes inspiration from.
@@
 
-The syntax tree in the implementation of \name{} consists of 'stnode' type
-pointers defined in \hrc{src/interpreter/parser.h}. Each such node depending
-on its 'nodetype', declared in \hrc{src/interpreter/tokenizer.h},
-is either a value-node storing a pointer to data, e.g. a number or text, or a
-parent-node referencing two child nodes and an operator, e.g. two numbers as
-children and a plus-operator to signify an addition.
+The syntax tree in the implementation of \name{} is a binary tree consisting of nodes, internally
+called \emph{stnode}. The datatype is defined in the header file \hrc{src/interpreter/parser.h}.
+Each node can either be an internal node having one or two nodes as children, or a leaf node
+which stands for a value, e.g. a number or a text. Each internal node, also known as parent node, signifies
+an operation done on its children. This can be an addition of two numbers for example.
 
@@
 each operator.
+In the end the parse tree contains all values and 
+operations as well as information about the order of execution.
 
@@
 \begin{itemize}
-    \item \emph{advance.} This function reads a new token from the tokenizer.
-    \item \emph{secondary.} The 'secondary' function deals with parsing single,
-        simple values that make up the start of an expression, such as numbers,
-        fields, strings and it handles brackets, such as indexing with square
-        brackets, functions or code blocks with curly braces and sub-expressions
-        in standard brackets.
-    \item \emph{expr.} The 'expr' function calls 'secondary' which acts
+    \item \emph{advance.} This function reads the next token from the tokenizer.
+    \item \emph{secondary.} This function reads tokens that will become the leaf nodes
+        of the parse tree.
+        It deals with simple values that make up the start of an expression, such as numbers,
+        fields, strings and it handles brackets. 
+        In addition, it can handle prefix operators, such as the unary plus or unary minus.
+    \item \emph{expr.} The \emph{expr} function calls \emph{secondary} whose result acts
         as the basis for a new expression. It then repeatedly checks
-        if an operator follows the secondary token and creates a tree of
-        EXPR type stnodes, each having children which are recursively parsed
-        and information about the operator.
+        if an operator follows the secondary token and creates the syntax-tree
+        with help of recursive mechanisms. The nature of this function also 
+        removees the need for semi-colons,
+        which are essential in languages like Java or C.
     \item \emph{parse.} The parse function is a small wrapper which only
-        calls the 'expr' function, initiating the parsing process.
+        calls the \emph{expr} function a first time, initiating the parsing process.
 \end{itemize}
@@
 the parse-tree can be beneficial. This is the case when the same piece of code
-has to be run several times. Instead of generating a time and memory overhead 
-having to walk along tree-nodes recursively, only the list of instructions in
-the bytecode can be traversed, which takes less time overall.
+has to be run several times. Compared to
+having to walk along tree nodes recursively, reading the list of instructions in
+the bytecode takes less time overall.
 
@@
 generates byte-sequences in an internal format that are then interpreted. A
-JIT compiler would instead compile the AST down to machine-native bytecode, which
-could run directly on the system's hardware and would not have to be interpreted.
+Just-In-Time compiler, JIT compiler for short, would instead compile the AST 
+down to machine-native bytecode, which
+runs directly on the system's hardware and does not have to be interpreted.
 
@@
 a prefix notation. 
-[\cte{shunting_yard}]
-
 A prefix notation is a way to write down mathematical expressions, except
@@
 where operands surround the operator.
+[\cte{shunting_yard}]
 
@@
         \expr{\emph{- 1 (/ (* 2 4) (+ 5 6))}}
-    \item But in \name{} the operands are pushed onto the stack is reverse order.
+    \item But in \name{} the operands are pushed onto the stack is reverse order,
+        as popping from the stack again reverses the order back to the original.
         This would look like the following:
@@
     \item The brackets are not needed, because for every operator, the amount
-        of operands is known. This is the case as e.g. the subtraction operator
-        and the negative operator, which are both a minus sign, are distinguished
-        beforehand. The above statement would therefore look like this:
+        of operands is known. This is also the case for the unary minus and unary plus,
+        because they are distinguished from their infix counterparts beforehand.
+        The above statement would therefore look like this:
         \expr{\emph{- / + 6 5 * 4 2 1}}
@@
 
-The evaluation of expressions also works in a recursive manner. Here is an 
+The evaluation of expressions works in a recursive manner. Here is an 
 example of \name{}'s bytecode notation and how it is evaluated at runtime:
@@
         calculation? The answer is recursion.
-    \item The runtime will read through the bytecode left-to-right. First it will
-        notice the subtraction-operator and accordingly assume that two
+    \item The runtime will read through the bytecode from left to right. First it will
+        notice the subtraction operator and accordingly assume that two
         operands must follow.
     \item The function recursively calls itself in order to evaluate the
-        right operand and notice the 6, a value. The function simply returns the six
-        to the subtraction expression.
+        right operand and notice the 6, a value and a leaf node in the parse tree. 
+        The function simply returns the 6 to the subtraction expression.
     \item Again, the function calls itself to find the next operand.
@@
 representing all possible operations. Extending upon that, there are some
-standard functions.
+standard functions that enlargen the capabilities of the language.
 
@@
 The syntax of \name{} is unlike most modern, common programming languages' syntax,
-but can be compared somone to that of JavaScript.
+but comes closest to that of JavaScript.
 In \name{}, there are no keywords at all and everything is controlled via operators,
 which are non-alphanumeric characters. Such operators include the plus-sign 
-for addition, or the equals-sign for variable assignment.
+for addition, the equals-sign for variable assignment or a set of brackets to signal
+a function call.
 
@@
 \begin{itemize}
-\item The source code begins by declaring two variables \emph{a} and \emph{b}, which
+\item The source code begins with the definition two variables \emph{a} and \emph{b}, which
     will be utilized to keep track of the data used to calculate fibonacci numbers.
-    The memory allocation is done by creating a tuple of \emph{a} and \emph{b}, as suggested by 
-    the comma seperating the two. Since both variables \emph{a} and \emph{b} are not yet
-    known to the interpreter, they are automatically created and assigned a value
-    which has no content and the \emph{undefined} typing.
-\item The two variables \emph{a} and \emph{b} are then reassigned new values on line
-    2 of the program. In this case it is numbers with values zero and one respectively.
-    In this example, the two expressions are seperated by a space, but they can
-    also be separated by newline-characters or semicolons.
-\item Another variable \emph{count} is declared and defined as number in one step on line 4.
+    They are each assigned a number, zero and one respectively.
+    Since the variable names \emph{a} and \emph{b} are not yet
+    known to the interpreter, they are automatically allocated in the current scope
+    and assigned a default value. That default value is then overwritten with the
+    assignment operator.
+\item Another variable \emph{count} is declared and defined as the number ten on line 4.
     This variable will be used to limit the amount of numbers the program outputs.
-\item Yet another variable, \emph{f}, is declared and assigned a function as its
-    value on line 6. In \name{}, normal functions do not have any arguments, 
-    but they can interact with variables that were created outside their scope.
-\item Lines 7 to 10 hold mathematical expressions. At first, a is assigned its current value
-    plus the current value of b. Then in line 8, b is assigned the now new value of a minus the
-    current value of b, which corresponds to the value of a before it was just reassigned.
-    Lastly the variable \emph{count} is decremented by one.
-\item Now come the interesting parts of this program.
-    On lines 12 and 13, the function \emph{std.print} is referenced. This function is implemented
-    in plain C code, which internally is not handled the same as regular functions in \name{},
-    and is wrapper to C's function \emph{printf}. The \emph{std.print} function
-    takes any \name{} variable or value and prints it to standard output. On line 13, the "\\n"
-    is replaced by a newline-character, as \name{} can interpret Java-like escape-sequences.
-\item Another standard-library function is \emph{std.if}. It takes two or three arguments.
-    If the first argument is true, then the second, else the the third argument or 
-    \emph{undefined} is returned. A value is true, only if it is a number not equal to zero
-    or a non-empty string. In the case of this example, \emph{std.if} returns \emph{f} if 
-    count is greater than zero, and an empty function if count reaches zero. The two
-    brackets following the call of the \emph{std.if} function signal yet another call to the
-    returned value.
-\item The final thing left to do is to get the ball roling by calling the function \emph{f} for
-    the first time. As stated above, it will now print numbers and recursively call itsself,
-    until the variable \emph{count} reaches a value of zero.
+\item Now the function \emph{std.repeat} is called. This function is a
+    standard library function, meaning it is included in the \name{} interpreter and
+    written in C. Another function is passed as argument to the function call.
+    This function will be repeated.
+\item On line 7 and 8, some calculations are done.
+    Remember, the equals sign stands for an assignment, not an
+    equation. First, \emph{a} is assigned the sum of \emph{a} and \emph{b}. Then 
+    \emph{b} is set to be the new \emph{a} minus \emph{b}, 
+    which is in turn the same as \emph{a} before the reassignment on line 7.
+\item On lines 10 and 11, the standard library function \emph{std.print} is used
+    to print out the value of a and a line break. The backslash in the text
+    that is print out signifies a so-called escape-sequence. The \emph{n} stands
+    for a newline, but other letters such as a \emph{t} for a tabulator are possible.
+\item Lastly, line 13 contains another calculation in which the variable \emph{count}
+    is decremented by one. Since this is the last instruction in the inner function,
+    it is simoultaneously its return value, e.g. the value of count is returned
+    to the \emph{std.repeat} function. The \emph{std.repeat} function will
+    repeat the inner function until the returned value is zero or undefined.
+    This means the inner function is executed exacly ten times, because
+    the initial value of count has to be decremented by one ten times until
+    it equals zero.
 \end{itemize}
@@
 \emph{f}, which calls the function \emph{std.exec} with var as argument.
-Line number six holds a comment, indicated by the two slashes at the start of the line.
+Line number seven holds a comment, indicated by the two slashes at the start of the line.
 Comments are ignored by the \name{} interpreter and their purpose is to contain notes
@@
 
+% TODO: link more examples
+
 \section{Discussion}
+The following paragraphs will go over some positive results and some caveats of the
+final product.
+
+The name \name{} was chosen, because it implised this programming language
+to be a learning project as well as a scripting language, which employ
+high-level abstractions of underlying operations and interpret one line
+at a time. The file name extension \emph{.nts} is used in the examples. This
+file extension is not required, but was chosen as it stands for \emph{Not TypeScript}, because
+the language neither enforces strict typing, nor is it Microsoft's \emph{TypeScript}.
 
@@
 \subsubsection{Numbers}
-% TODO
+Another goal that turned out not to be practical to acheive was to use just as much
+memory for each variable as needed and not more. This means every variable would
+have to be assigned a datatype, containing information about the size and possible values.
+Taking C as a reference, the datatype \emph{char} is exactly one byte in size, and
+can therefore be in exactly 2^8 or 256 different states. But in C, information about
+the datatype can be discarded at compile time, whereas in an interpreter like \name{} such
+information needs to persist throughout the program's lifetime.
 
-\subsubsection{Memory Leaks}
+\subsubsection{Memory Leaks}\label{memleaks}
 Memory leaks happen when a program reserves memory to store data, but
@@
 has long been forgotten about and will no longer be accessed.
-In \name{}, memory leaks are a major problem. Even though barely noticable for small
+In \name{}, memory leaks are a major problem. Even though barely noticable in small
 programs it interprets, \name{} does not have proper setups in place to free up
@@
 
-Other programming languages, notably object-oriented languages such as Java,
-use Garbage-Collection, which automatically tracks unused allocated memory and
-frees it up automatically. While this is also the case with \name{}, there are
-many internal flaws that have nothing to do with variables used in a \name{}
-program, but rather internally.
+Manx rogramming languages, such as Java and Python,
+make use of Garbage-Collection, a strategy to track unused allocated memory and
+free it up automatically. While \name{} also tries to make use of this technique,
+there are many flaws introduced while developing the interpreter that hinder 
+Garbage-Collection from working correctly.
 
-One reason memory leaks is the nature of the
+One reason for memory leaks is the nature of the
 C programming language, in which the \name{} interpreter was written in. The C language and
 compiler do not keep track of all possible states in which certain blocks of memory are
-forgotten about and do not warn the developer. Instead, memory leaks happen.
-More modern C alternatives such as Rust have an ownership-system in place,
+forgotten about and do not warn the developer. Instead, memory leaks happen and a program
+uses much more system resources than required.
+Modern alternatives for C such as Rust deal with the problem in a more elegant way. 
+Rust has an ownership system in place,
 in which any allocated memory is owned by a function. When the function
@@
 
+A tool that is helpful to find and fix memory leaks is Valgrind. It runs
+the executable program against an input and finds the origins of unfreed data.
+
 \subsubsection{Single-Threaded}
-The \name{} interpreter does not implement any multi-threaded, i.e. in parallel
-running programs as most other languages do. Instead, any \name{} program can
-only run sequentially. Since most modern computer processors
-are equipped with a handful of processing cores and threads, not being able to
-make use of this modern time's advantage is a major downside.
+The \name{} interpreter does not implement any multi-threading or multiprocessing.
+These techniques make use of the multiple CPU cores available in modern computers
+to run programs in parallel.
+Any \name{} program can therefore
+only contain instructions that run sequentially.
+Since most languages used today allow increasing a
+program's throughput using multi-threading, not having it at hand in \name{}
+is a major limitation.
 
 \subsubsection{Lack of a Development Environment}
-This is certainly the greatest problem of \name{}.
+This is probably the greatest problem of \name{} and any new prorgramming
+languages in general.
 
-Most modern programming languages not only consist
+Most widely used programming languages not only consist
 of a language specification and a compiler or interpreter, but of many more
@@
 Such tools can include a package-manager, with which people around the world
-can share code. Examples would be npm or cargo, 
+can share code. Examples would be \emph{npm} or \emph{cargo}, 
 the 'Node Package Manager' for JavaScript applications and and Rust's package 
@@
 
+On the other hand, having to deal with countless package-managers for each
+language can be annoying for developers. A solution is to build upon and
+extend already existing languages and their tools with e.g. a new syntax.
+This is exactly what Kotlin does. Kotlin can produce Java Virtual Machine
+compatible bytecode, can be transpiled into
+JavaScript, or ECMAScript 5.1 to be more specific,
+and it can be compiled to machine-native programs via LLVM. 
+[\cte{kotlin_compile}]
+
+% TODO: write about LLVM and recommend it lol6
+
 \subsection{Interesting Ideas untouched}
@@
 out. This section of the paper gives insight on interesting ideas or concepts
-that are great to have and are used in other programming languages, but not 
+that are interesting and are used in other programming languages, but 
 could not be implemented into \name{}.
@@
 \subsubsection{Foreign Function Interfaces}
-A foreign function interface often abbreviated as FFI,
+A foreign function interface, often abbreviated as FFI,
 is a mechanism by which one programming language can call functions written 
@@
 of this paper, the details of the implementation brought about unexpected 
-problems and solutions. The goal of this paper to learn about 
-the inner workings of an interpreter has been achieved, as well as the goal
-to get to know the C language.
+problems and solutions. The goal of this project to learn about 
+the inner workings of an interpreter and
+to get to know the C language has been acheived.
 
@@
 the entire source code of \name{} by myself using resources listed under the
-'Bibliography' section on page \pageref{bibliography}.
+\emph{Bibliography} section on page \pageref{bibliography}.
 % TODO: check correctness of all file paths!! e.g. src/...
@@
 The appendix contains the complete source code of \name{}.
-See the file tree under 'File Structure' in section \ref{FileStructure}
+See the file tree under \emph{File Structure} in section \ref{FileStructure}
 to get an overview.

### 2022-10-29 src/interpreter/interpreter.c
fixing memory leaks

@@ -63,8 +63,11 @@ void interpret(FILE *stream, char* filename) {
         #endif
 
         // break if null or FILE_END
-        if(root == NULL || root->type == FILE_END)
+        if(root == NULL || root->type == FILE_END) {
+            if(root != NULL)
+                free(root);
             break;
+        }
 
         // resolve fields
         localize(root, memory, &cac.info);
@@ -91,6 +94,7 @@ void interpret(FILE *stream, char* filename) {
     #endif
 
     // free token content
+    free_stack(cac.bracketstack);
     free(tkn.content);
 }
 

### 2022-10-29 src/interpreter/libraries/stdlib/stdlib.c

@@ -27,6 +27,7 @@ void _type(mementry *args, mementry *dst, errorinfo *info) {
 void _random(mementry *args, mementry *dst, errorinfo *info) {
     dst->value = malloc(sizeof(number));
     *((number *)dst->value) = (number) rand() / (number) RAND_MAX;
+    dst->flags.val_tmp = 1;
     dst->type = NUMBER;
     return;
 }
@@ -92,10 +93,7 @@ void _if(mementry *args, mementry *dst, errorinfo *info) {
                 dst->type = ret->type;
                 dst->level = ret->level;
                 dst->value = ret->value;
-
-                // free fun if synthetic, but not value
-                if(ret->flags.synthetic)
-                    free(ret);
+                dst->flags.val_tmp = 0;
             }
             return;
         }
@@ -124,9 +122,6 @@ void _repeat(mementry *args, mementry *dst, errorinfo *info) {
         dst->type = tmp->type;
         dst->value = tmp->value;
     } while(truth_of(dst));
-    
-    // free params
-    free(params);
 }
 
 // get lenght of arrays, tuples, strings

### 2022-10-29 src/interpreter/memory/hashtable.c

@@ -21,7 +21,7 @@ mementry *_alloc_mementry() {
     ret->value = NULL;
     ret->flags =
         (struct s_mementry_flags) 
-        {.mutable = 0, .synthetic = 0, .value_synthetic = 0};
+        {.from_ht= 0,.val_tmp=0};
     return ret;
 }
 
@@ -366,7 +366,7 @@ static inline tableentry *_find(hashtable *table, char *key) {
     new_value->entry->value = NULL;
     new_value->entry->level = table->level;
     new_value->entry->flags = 
-        (struct s_mementry_flags) {.mutable=1, .synthetic=0, .value_synthetic=0};
+        (struct s_mementry_flags) {.from_ht=1, .val_tmp=0};
 
     // register entry in cache
     table->cache[hash % table->cache_size] = new_value;
@@ -397,7 +397,8 @@ void walk_table(hashtable *table, void (*callback)(tableentry *)) {
 }
 
 mementry *find(hashtable *table, char *key) {
-    char *next = key = strdup(key);
+    char *tst;
+    char *next = tst = key = strdup(key);
     char done = 0;
 
     mementry *ret = NULL;
@@ -417,7 +418,7 @@ mementry *find(hashtable *table, char *key) {
         ret = _find(table, key)->entry;
 
         if(done || ret == NULL)
-            return ret;
+            break;
 
         // just allocate it as as an object
         if(ret->type == UNDEFINED) {
@@ -426,7 +427,7 @@ mementry *find(hashtable *table, char *key) {
         }
 
         if(ret->type != OBJECT)
-            return ret;
+            break;
 
         // make the result the new parent object
         table = (hashtable *)ret->value;
@@ -434,6 +435,8 @@ mementry *find(hashtable *table, char *key) {
         // make sure to search for the child's key
         key = ++next;
     }
+    free(tst);
+    return ret;
 }
 
 void free_mementry(mementry *entry) {

### 2022-10-29 src/interpreter/memory/hashtable.h

@@ -39,15 +39,10 @@ typedef struct {
 
     // some flags
     struct s_mementry_flags {
-        // set if this mementry is a mutable field (mutable entry in hashtable)
-        unsigned char mutable: 1;
-
-        // > used by implementations.c < to see if
-        // a mementry only exists during expression-evaluation
-        unsigned char synthetic: 1; 
-        // used to check if the value this mementry is pointing to is synthetic
-        // this will only be set, if synthetic is set as well
-        unsigned char value_synthetic: 1;
+        // check if allocated by hashtable
+        unsigned char from_ht: 1;
+        // check if value is temporary, can be discarded
+        unsigned char val_tmp: 1;
     } flags;
 
 } mementry;

### 2022-10-29 src/interpreter/processing/implementations.c

@@ -30,13 +30,22 @@ char truth_of(mementry *entry) {
     return 1;
 }
 
-
 // free a synthetic mementry and its synthetic value
 void _free_synth(mementry *entry) {
-    if(entry->flags.synthetic) {
-        if(entry->flags.value_synthetic && entry->type != UNDEFINED)
-            free(entry->value);
-        free(entry);
+    if(!entry->flags.from_ht) {
+        if(entry->flags.val_tmp){
+            switch(entry->type){
+                case UNDEFINED:
+                    break;
+                case ARRAY:
+                case TUPLE:
+                    free_array((array *) entry->value);
+                    break;
+                default:
+                    free(entry->value);
+                    break; 
+            } 
+        }
     }
 }
 
@@ -52,7 +61,6 @@ mementry *_recursiveprocess(opargs *args, char flags) {
         memptr = _alloc_mementry();
         memptr->type = UNDEFINED;
         memptr->value = NULL;
-        memptr->flags = (struct s_mementry_flags) {.mutable = 0, .synthetic = 1, .value_synthetic = 0};
         return memptr;
     }
 
@@ -72,7 +80,7 @@ mementry *_recursiveprocess(opargs *args, char flags) {
         // only return the value, if it's a mutable field
         if(memptr->type == REFERENCE) {
             memptr = memptr->value;
-            if(memptr->flags.mutable)
+            if(memptr->flags.from_ht)
                 return memptr;
         }
         return NULL;
@@ -172,31 +180,29 @@ mementry *_bnot(opargs *args) {
     \
     /* determine the dst entry (or allocate a new one if ro and lo are not synthetics) */\
     mementry *dst = NULL;\
-    if(lo->flags.synthetic)\
+    if(lo->flags.val_tmp)\
         dst = lo;\
-    else if(ro->flags.synthetic)\
+    else if(ro->flags.val_tmp)\
         dst = ro;\
     else {\
         dst = _alloc_mementry();\
         dst->value = malloc(sizeof(number));\
         dst->type = NUMBER;\
-        dst->flags = \
-        (struct s_mementry_flags) {.mutable=0, .synthetic=1, .value_synthetic=1};\
+        dst->flags.val_tmp = 1; \
     }\
     \
-    number *valdst = dst->flags.value_synthetic ? dst->value : malloc(sizeof(number));\
+    number *valdst = dst->flags.val_tmp ? dst->value : malloc(sizeof(number));\
+    dst->value = valdst; \
[…]

### 2022-10-29 paper.tex
spellchecking

@@
     \expandafter\ifstrequal\expandafter{\jobname}{paper}
-    {#1}{\begin{center}-Removed, see full version-\end{center}}
+    {#1}{}
 }
@@
     % Include Mandelbrot on title page
+    \ver{
     \begin{figure}[H]
@@
     \end{figure}
+    }
 
@@
 By making use of just the term \emph{language} this paper will refer to programming
-languages specifically. Sources are mentioned fater cohesive paragraphs in
+languages specifically. Sources are mentioned after cohesive paragraphs in
 square brackets, an index can be found in the \emph{Bibliography} on page
@@
 Since the year 1973, when the first programming language was invented, the
-research in programming languages has fastly advanced. But the goal has stayed
+research in programming languages has vastly advanced. But the goal has stayed
 the same, to make programming as intuitive and easy, yet fast and efficient as
@@
 The tools \emph{Valgrind} and \emph{gdb}, the GNU Project Debugger were of great help
-regarding the finding of errors and memory leaks, see section \ref{memleaks} for details.
+when trying to find of errors and memory leaks, see section \ref{memleaks} for details.
 \emph{Git} is the project management and version tracking software used to back up and share
@@
 The problem to solve in this paper is concisely to make a computer understand
-human-written and -readable text, called source code, and follow the instructions
+human written and -readable text, called source code, and follow the instructions
 it proposes.
@@
         in \hrc{src/main.c}. It returns a value
-        of type 'int', has two arguments, 'argc' and 'argv', and a function body
-        which, in the case of the example above, prints the value of the 'argc' argument.
+        of type \emph{int}, has two arguments, \emph{argc} and \emph{argv}, and a function body
+        which, in the case of the example above, prints out the value of the \emph{argc} argument.
 
@@
 read a program from a file or from standard input. 
-At last, the interpret function
+At last, the function \emph{interpret}
 from the said included header is called with the selected input.
@@
 
-The \emph{interpret} function calls the parser to generate a syntax-tree from tokens retrieved
+The function \emph{interpret} calls the parser to generate a syntax-tree from tokens retrieved
 from the tokenizer implemented in \hrc{src/interpreter/parser.c} and 
@@
 bytecode compilation has a performance benefit. 
-The mapping is acheived with a custom hashtable datastructure with average time
+The mapping is achieved with a custom hash table data structure with average time
 complexity of O(log n), n being the total amount of variables a \name{} program is using.
-The hashtable implementation used in \hrc{src/interpreter/memory/hashtable.c}                   % TODO: maybe its own paragraph! interpolation search vs binary search
+The hash table implementation used in \hrc{src/interpreter/memory/hashtable.c}                   % TODO: maybe its own paragraph! interpolation search vs binary search
 uses binary search to locate elements variable-sized arrays in the hashtable's buckets.
@@
 \begin{enumerate}
-    \item Token of type FIELD with content \emph{a.b}.
-    \item Token of type SYMBOL with content \emph{==}.
-    \item Token of type NUMBER with content \emph{23.4}.
+    \item A token of type FIELD with content \emph{a.b}.
+    \item A second token of type SYMBOL with content \emph{==}.
+    \item A third token of type NUMBER with content \emph{23.4}.
 \end{enumerate}
@@
 While tokenizers can be programmed using regular expressions, 
-this paper uses a more simple finite-state machine tokenizer.
+this paper uses a simpler finite-state machine tokenizer.
 It reads character by character, determining the token type as it goes.
-It stops when there is a token thad does no longer match the current token type.
+It stops when there is a token that does no longer match the current token type.
 Tokens are each requested and integrated into a syntax-tree by the parser as described in the next
@@
 result in \emph{14} and is not the same as \emph{(2+3)*4}.
-Assosiativity is either 'Left to Right' or 'Right to Left'. An example of associativity is
-\emph{20/10/2}. Both in C as well as in \name{}, the division operator is
+Associativity is either 'Left to Right' or 'Right to Left'. An example of associativity is
+\emph{20/10/2}. Both in C and in \name{}, the division operator is
 left-to-right associative, meaning the stated expression is the same as
@@
         with help of recursive mechanisms. The nature of this function also 
-        removees the need for semi-colons,
+        removes the need for semicolons,
         which are essential in languages like Java or C.
@@
 
-Following is a piece of pseudocode to better illustrate the inner workings of the
+Following is a piece of pseudo code to better illustrate the inner workings of the
 parser which is based on a Pratt-parser. 
@@
 
-\subsubsection{The Runtime}
+\subsubsection{Program Evaluation}
 This is the last stage of interpreting program code as implemented in \name{}.
-The runtime steps through the intermediary bytecode created by the bytecode
+The interpreter steps through the intermediary bytecode created by the bytecode
 compiler and evaluates expressions by plugging in the provided values.
@@
 The evaluation of expressions works in a recursive manner. Here is an 
-example of \name{}'s bytecode notation and how it is evaluated at runtime:
+example of \name{}'s bytecode notation and how it is evaluated:
 \begin{itemize}
@@
         calculation? The answer is recursion.
-    \item The runtime will read through the bytecode from left to right. First it will
+    \item The interpreter will read through the bytecode from left to right. First it will
         notice the subtraction operator and accordingly assume that two
@@
 That is it. With all the operators and standard functions that have been implemented, entire programs
-can be interpreted. Examples of such programs can be found in the \emph{examples} directory
-of the source code.
+can be interpreted. Examples of such programs can be found in the directory
+\emph{examples} of the source code.
 
@@
 representing all possible operations. Extending upon that, there are some
-standard functions that enlargen the capabilities of the language.
+standard functions that enlarge the capabilities of the language.
 
@@
 Following is a piece of example \name{} code, printing out the first ten numbers
-of the fibonacci sequence, as well as a disection of how that same code is interpreted.
+of the Fibonacci sequence, as well as a dissection of how that same code is interpreted.
 \code{examples/fibonacci.nts}{1}{999}
@@
 \item The source code begins with the definition two variables \emph{a} and \emph{b}, which
-    will be utilized to keep track of the data used to calculate fibonacci numbers.
+    will be utilized to keep track of the data used to calculate Fibonacci numbers.
     They are each assigned a number, zero and one respectively.
@@
     to print out the value of a and a line break. The backslash in the text
-    that is print out signifies a so-called escape-sequence. The \emph{n} stands
-    for a newline, but other letters such as a \emph{t} for a tabulator are possible.
+    that is print out signifies a so-called escape sequence and will be replaced with
+    piece of text that can not be entered directly. The \emph{n} stands
+    for a line break, but other letters such as a \emph{t} for a tabulator are possible.
 \item Lastly, line 13 contains another calculation in which the variable \emph{count}
     is decremented by one. Since this is the last instruction in the inner function,
-    it is simoultaneously its return value, e.g. the value of count is returned
+    it is simultaneously its return value, e.g. the value of count is returned
     to the \emph{std.repeat} function. The \emph{std.repeat} function will
     repeat the inner function until the returned value is zero or undefined.
-    This means the inner function is executed exacly ten times, because
+    This means the inner function is executed exactly ten times, because
     the initial value of count has to be decremented by one ten times until
@@
 On line seven, the \emph{std.exec} function is called for the first time. It runs the
-command it is given as argument in the user's shell, which is most commonly \emph{cmd} on
+command it is given as argument in the user's shell, which is most commonly \emph{cmd.exe} on
 Microsoft Windows and \emph{bash} on most Linux distributions. In this example, \emph{std.exec}
@@
 
-This security concept enforces the use of a top-down programming model. First, safe-to-use aliases
+This security concept enforces the use of a top-down programming model. First, safe to use aliases
 to dangerous functions have to be created. Then, access to all those exploitable functions is
 revoked. This way, running unknown code can only access data and functions that are being provided.
-This contrasts the models of languages like Python. In Python, each seperate source file can
+This contrasts the models of languages like Python. In Python, each separate source file can
 make use of the \emph{import} keyword to include other files. Files imported this way do not
@@
 In \name{}, each imported file would instead have access to all variables and functions that
-are made available to it, instead of having to include or import them seperately.
+are made available to it, instead of having to include or import them separately.
 This leads to another problem, but that is what the next section will be all about.
@@
 
-The name \name{} was chosen, because it implised this programming language
+The name \name{} was chosen, because it implied this programming language
 to be a learning project as well as a scripting language, which employ
@@
 \subsubsection{Numbers}
-Another goal that turned out not to be practical to acheive was to use just as much
+Another goal that turned out not to be practical to achieve was to use just as much
 memory for each variable as needed and not more. This means every variable would
@@
 Taking C as a reference, the datatype \emph{char} is exactly one byte in size, and
-can therefore be in exactly 2^8 or 256 different states. But in C, information about
+can therefore be in exactly 2\^8 or 256 different states. But in C, information about
 the datatype can be discarded at compile time, whereas in an interpreter like \name{} such
@@
 has long been forgotten about and will no longer be accessed.
-In \name{}, memory leaks are a major problem. Even though barely noticable in small
+In \name{}, memory leaks are a major problem. Even though barely noticeable in small
 programs it interprets, \name{} does not have proper setups in place to free up
@@
 
-Manx rogramming languages, such as Java and Python,
+Many programming languages, such as Java and Python,
 make use of Garbage-Collection, a strategy to track unused allocated memory and
@@
 
-A tool that is helpful to find and fix memory leaks is Valgrind. It runs
-the executable program against an input and finds the origins of unfreed data.
+A tool that is helpful to find and fix memory leaks is \emph{Valgrind}. It runs
+the executable program against an input and finds the origins of not freed data.
 
@@
 of a language specification and a compiler or interpreter, but of many more
-tools that surround themn.
+tools that surround them.
 Such tools can include a package-manager, with which people around the world
@@
 This is exactly what Kotlin does. Kotlin can produce Java Virtual Machine
-compatible bytecode, can be transpiled into
+compatible bytecode, can be transcompiled, e.g. translated, into
 JavaScript, or ECMAScript 5.1 to be more specific,
@@
 re-written. Enabling C function calls often allows a certain language to
-directly interact with system libraries or the operating system itsself.
+directly interact with system libraries or the operating system itself.
 
@@
 Interface, JNI for short, whereas within Python C function can
-be run with the standard ctypes library.
+be run with the standard library \emph{ctypes}.
 
@@
 of objects.
-The advantage of reflection over not having it is that it allows for non-static
-programs and program flows, effectively making smaller executables and more
-resource efficient programs, as those programs can adjust themselves to run
-differently under different conditions at runtime.
 The easiest and simplest way to go about implementing Reflection in \name{} is
@@
 
-\newpage\subsection{Bibliography}
+\subsection{Bibliography}
 \renewcommand*{\bibfont}{\normalsize}
@@
 
-\section{Declaration of Autonomy}
-Hereby I, Florian Donnelly, declare to have written this paper, also including
-the entire source code of \name{} by myself using resources listed under the
-\emph{Bibliography} section on page \pageref{bibliography}.
+\section{Declaration of Authenticity}
+Hereby I, Florian Donnelly, declare to have written this paper independently, 
+also including the entire source code of \name{} by myself using resources 
+listed under the \emph{Bibliography} section on page \pageref{bibliography}.
 % TODO: check correctness of all file paths!! e.g. src/...

### 2022-10-29 paper.tex
continue writing paper

@@
 
+\definecolor{mygreen}{rgb}{0.3, 0.5, 0.2}
+\definecolor{myteal}{rgb}{0.1, 0.2, 0.3}
+\definecolor{myblue}{rgb}{0.2, 0.0, 0.6}
+\definecolor{mypink}{rgb}{0.8, 0.3, 0.6}
+\definecolor{myorange}{rgb}{0.7, 0.4, 0.1}
+
+% Use fancier hyperrefs
+\hypersetup{
+    colorlinks=true,
+    pdfborder={3 3 .2},
+    citecolor=mygreen,
+    linkcolor=myteal,
+    urlcolor=myblue,
+}
+
 % hyperref abbreviations
+\makeatletter
+\DeclareRobustCommand{\filename}[1]{%
+    \begingroup
+    \def\textendash{-}%
+    \filename@parse{#1}%
+    \edef\filename@base{\detokenize\expandafter{\filename@base}}%
+    \texttt{\filename@base.\filename@ext}%
+    \endgroup
+}
+\makeatother
 \newcommand{\hr}[2]{\hyperref[#2]{#1}}
-\newcommand{\hrc}[1]{\hyperref[#1]{#1}}
+\newcommand{\hrc}[1]{\hyperref[#1]{\filename{#1}}}
 
@@
 % Include C code from file #1 from lines #2 to #3
+\lstset{basicstyle=\scriptsize,keywordstyle=\color{myorange},stepnumber=5,
+        frame=none,keepspaces=false,
+        rulecolor=\color{black},tabsize=2,
+        stringstyle=\color{mygreen},commentstyle=\color{mypink},
+        morekeywords={NULL},
+        morecomment=[l][\color{magenta}]{\#},
+        postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space}}
+
 \newcommand{\sourcecode}[3] {
     \ver{
-    \lstset{basicstyle=\scriptsize,keywordstyle=\color{blue},stepnumber=5,
-        stringstyle=\color[rgb]{.3,.5,.1},commentstyle=\color{gray},
-        morecomment=[l][\color{magenta}]{\#},
-        postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space}}
         \lstinputlisting[language=C,numbers=left,breaklines=true,
@@
     \begin{minipage}{\linewidth}
-
-    \lstset{basicstyle=\scriptsize,keywordstyle=\color{blue},stepnumber=5,
-        stringstyle=\color[rgb]{.3,.5,.1},commentstyle=\color{gray},
-        morecomment=[l][\color{magenta}]{\#},
-        postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space}}
     \lstinputlisting[language=C,numbers=left,breaklines=true,
-        frame=single,title=\hrc{#1},
+        frame=single,title=\hr{#1}{#1},
         firstnumber=#2,firstline=#2,lastline=#3]{#1}
@@
 
+{
+    \hypersetup{hidelinks}
     \tableofcontents\newpage
+}
 
@@
 \emph{GNU+Linux}, the operating system used to program on.
-The code editor used was \emph{Microsoft Visual Studio Code}, providing great
-help with built-in git integration and auto-completion features.
-The tools \emph{Valgrind} and \emph{gdb}, the GNU Project Debugger were of great help
+The code editor used was \emph{Microsoft Visual Studio Code}.
+It has built-in git integration and auto-completion features.
+The tools \emph{Valgrind} and \emph{gdb}, the GNU Project Debugger, were of help
 when trying to find of errors and memory leaks, see section \ref{memleaks} for details.
@@
         .1 project root.
-        .2 \hr{.gitignore}{.gitignore}.
-        .2 \hr{README.md}{README.md}.
-        .2 \hr{Makefile}{Makefile}.
-        .2 \hr{mapop.gperf}{mapop.gperf}.
+        .2 \hrc{.gitignore}.
+        .2 \hrc{README.md}.
+        .2 \hrc{Makefile}.
+        .2 \hrc{mapop.gperf}.
         .2 pseudocode.
-            .3 \hr{parser.c}{pseudocode/parser.c}.
+            .3 \hrc{pseudocode/parser.c}.
         .2 examples.
-            .3 \hr{mandelbrot.nts}{examples/mandelbrot.nts}.
-            .3 \hr{fibonacci.nts}{examples/fibonacci.nts}.
-            .3 \hr{exec.nts}{examples/exec.nts}.
+            .3 \hrc{examples/mandelbrot.nts}.
+            .3 \hrc{examples/fibonacci.nts}.
+            .3 \hrc{examples/exec.nts}.
         .2 src.
-            .3 \hr{main.c}{src/main.c}.
+            .3 \hrc{src/main.c}.
             .3 interpreter.
-                .4 \hr{bytecode.c}{src/interpreter/bytecode.c}.
-                .4 \hr{bytecode.h}{src/interpreter/bytecode.h}.
+                .4 \hrc{src/interpreter/bytecode.c}.
+                .4 \hrc{src/interpreter/bytecode.h}.
                 .4 error.
-                    .5 \hr{error.c}{src/interpreter/error/error.c}.
-                    .5 \hr{error.h}{src/interpreter/error/error.h}.
-                .4 \hr{interpreter.c}{src/interpreter/interpreter.c}.
-                .4 \hr{interpreter.h}{src/interpreter/interpreter.h}.
+                    .5 \hrc{src/interpreter/error/error.c}.
+                    .5 \hrc{src/interpreter/error/error.h}.
+                .4 \hrc{src/interpreter/interpreter.c}.
+                .4 \hrc{src/interpreter/interpreter.h}.
                 .4 libraries.
-                    .5 \hr{libraries.c}{src/interpreter/libraries/libraries.c}.
-                    .5 \hr{libraries.h}{src/interpreter/libraries/libraries.h}.
+                    .5 \hrc{src/interpreter/libraries/libraries.c}.
+                    .5 \hrc{src/interpreter/libraries/libraries.h}.
                     .5 stdlib.
-                        .6 \hr{stdlib.c}{src/interpreter/libraries/stdlib/stdlib.c}.
-                        .6 \hr{stdlib.h}{src/interpreter/libraries/stdlib/stdlib.h}.
-                .4 \hr{localizer.c}{src/interpreter/localizer.c}.
-                .4 \hr{localizer.h}{src/interpreter/localizer.h}.
+                        .6 \hrc{src/interpreter/libraries/stdlib/stdlib.c}.
+                        .6 \hrc{src/interpreter/libraries/stdlib/stdlib.h}.
+                .4 \hrc{src/interpreter/localizer.c}.
+                .4 \hrc{src/interpreter/localizer.h}.
                 .4 mappings.
-                    .5 \hr{operations.c}{src/interpreter/mappings/operations.c}.
-                    .5 \hr{operations.h}{src/interpreter/mappings/operations.h}.
-                    .5 \hr{mapop.c}{src/interpreter/mappings/mapop.c}.
-                    .5 \hr{mapop.h}{src/interpreter/mappings/mapop.h}.
+                    .5 \hrc{src/interpreter/mappings/operations.c}.
+                    .5 \hrc{src/interpreter/mappings/operations.h}.
+                    .5 \hrc{src/interpreter/mappings/mapop.c}.
+                    .5 \hrc{src/interpreter/mappings/mapop.h}.
                 .4 memory.
-                    .5 \hr{hashtable.c}{src/interpreter/memory/hashtable.c}.
-                    .5 \hr{hashtable.h}{src/interpreter/memory/hashtable.h}.
-                    .5 \hr{array.c}{src/interpreter/memory/array.c}.
-                    .5 \hr{array.h}{src/interpreter/memory/array.h}.
-                    .5 \hr{stack.c}{src/interpreter/memory/stack.c}.
-                    .5 \hr{stack.h}{src/interpreter/memory/stack.h}.
-                .4 \hr{parser.c}{src/interpreter/parser.c}.
-                .4 \hr{parser.h}{src/interpreter/parser.h}.
+                    .5 \hrc{src/interpreter/memory/hashtable.c}.
+                    .5 \hrc{src/interpreter/memory/hashtable.h}.
+                    .5 \hrc{src/interpreter/memory/array.c}.
+                    .5 \hrc{src/interpreter/memory/array.h}.
+                    .5 \hrc{src/interpreter/memory/stack.c}.
+                    .5 \hrc{src/interpreter/memory/stack.h}.
+                .4 \hrc{src/interpreter/parser.c}.
+                .4 \hrc{src/interpreter/parser.h}.
                 .4 processing.
-                    .5 \hr{implementations.c}{src/interpreter/processing/implementations.c}.
-                    .5 \hr{implementations.h}{src/interpreter/processing/implementations.h}.
-                .4 \hr{runtime.c}{src/interpreter/runtime.c}.
-                .4 \hr{runtime.h}{src/interpreter/runtime.h}.
-                .4 \hr{tokenizer.c}{src/interpreter/tokenizer.c}.
-                .4 \hr{tokenizer.h}{src/interpreter/tokenizer.h}.
+                    .5 \hrc{src/interpreter/processing/implementations.c}.
+                    .5 \hrc{src/interpreter/processing/implementations.h}.
+                .4 \hrc{src/interpreter/runtime.c}.
+                .4 \hrc{src/interpreter/runtime.h}.
+                .4 \hrc{src/interpreter/tokenizer.c}.
+                .4 \hrc{src/interpreter/tokenizer.h}.
     }
@@
 in this paper. Examples will be made with both pseudo code and actual source
-code from the project to further illustrate the decisions made during
-interpreter design. The file paths used are relative to the project's root directory.
+code from the project to show the decisions made during
+interpreter design. Only file names are used, as those are unique. To view the
+entire file path, look up the file names in the file tree on page \pageref{FileStructure}.
 
@@
 At last, the function \emph{interpret}
-from the said included header is called with the selected input.
+from the included header is called with the selected input.
 
@@
 \subsubsection{Lack of a Development Environment}
-This is probably the greatest problem of \name{} and any new prorgramming
+This is probably the largest problem of \name{} and new programming
 languages in general.
@@
 
-\subsection{List of Figures}\listoffigures
+\subsection{List of Figures}
+{
+    \hypersetup{hidelinks}
+    \listoffigures
+}
 \vspace{1cm}
@@
 Hereby I, Florian Donnelly, declare to have written this paper independently, 
-also including the entire source code of \name{} by myself using resources 
+also including the entire source code of \name{}, using resources 
 listed under the \emph{Bibliography} section on page \pageref{bibliography}.

### 2022-10-29 src/interpreter/mappings/mapop.c

@@ -1,5 +1,5 @@
 /* ANSI-C code produced by gperf version 3.1 */
-/* Command-line: gperf --output-file=./src//interpreter/mappings/mapop.c ./mapop.gperf  */
+/* Command-line: gperf --output-file=./src/interpreter/mappings/mapop.c ./mapop.gperf  */
 /* Computed positions: -k'1-2' */
 
 #if !((' ' == 32) && ('!' == 33) && ('"' == 34) && ('#' == 35) \

### 2022-10-29 src/interpreter/bytecode.c
try fixing memory leaks

@@ -11,6 +11,7 @@
 static inline mementry *_register(bytecode *code, stackptr ptr, typing type) {
     mementry *ret = 
         *(mementry **)push(code, ptr, sizeof(mementry *)) = _alloc_mementry();
+    ret->flags.persistent = 1;
     ret->type = type;
     return ret;
 }

### 2022-10-29 src/interpreter/interpreter.c

@@ -78,6 +78,7 @@ void interpret(FILE *stream, char* filename) {
         // process bytecode, also passing debug information from parser
         process(code, &cac.info);
 
+        free_stack(code);
     };
 
     #if DEBUG

### 2022-10-29 src/interpreter/libraries/stdlib/stdlib.c

@@ -122,6 +122,8 @@ void _repeat(mementry *args, mementry *dst, errorinfo *info) {
         dst->type = tmp->type;
         dst->value = tmp->value;
     } while(truth_of(dst));
+
+    free(params);
 }
 
 // get lenght of arrays, tuples, strings

### 2022-10-29 src/interpreter/localizer.c

@@ -34,7 +34,7 @@ void localize(stnode *subtree, hashtable *table, errorinfo *info) {
 
             if(subtree->data.parent.left != NULL) {
                 // change ht level
-                ht_up(table);
+                // ht_up(table);
 
                 // localize child first
                 if(subtree->data.parent.right != NULL)
@@ -44,7 +44,7 @@ void localize(stnode *subtree, hashtable *table, errorinfo *info) {
                 localize(subtree->data.parent.left, table, info);
                 
                 // reset ht level
-                ht_down(table);
+                // ht_down(table);
             }
 
             break;

### 2022-10-29 src/interpreter/memory/hashtable.c

@@ -21,7 +21,7 @@ mementry *_alloc_mementry() {
     ret->value = NULL;
     ret->flags =
         (struct s_mementry_flags) 
-        {.from_ht= 0,.val_tmp=0};
+        {.from_ht= 0,.val_tmp=0,.persistent=0};
     return ret;
 }
 

### 2022-10-29 src/interpreter/memory/hashtable.h

@@ -43,6 +43,8 @@ typedef struct {
         unsigned char from_ht: 1;
         // check if value is temporary, can be discarded
         unsigned char val_tmp: 1;
+        // if it is part of the bytecode 
+        unsigned char persistent: 1;
     } flags;
 
 } mementry;

### 2022-10-29 src/interpreter/processing/implementations.c

@@ -31,8 +31,9 @@ char truth_of(mementry *entry) {
 }
 
 // free a synthetic mementry and its synthetic value
-void _free_synth(mementry *entry) {
-    if(!entry->flags.from_ht) {
+void _free_synth(mementry *entry, char free_entry) {
+    if(!entry->flags.from_ht && !entry->flags.persistent) {
+
         if(entry->flags.val_tmp){
             switch(entry->type){
                 case UNDEFINED:
@@ -40,6 +41,9 @@ void _free_synth(mementry *entry) {
                 case ARRAY:
                 case TUPLE:
                     free_array((array *) entry->value);
+                    break;
+                case FUNCTION:
+                    free_stack((bytecode *)entry->value);
                     break;
                 default:
                     free(entry->value);
@@ -198,11 +202,10 @@ mementry *_bnot(opargs *args) {
     *valdst = calculation;\
     \
     /* free right operand if it's not the entry to be returned */\
-    if(dst != lo && ro != lo)\
-        _free_synth(lo);\
-    if(dst != ro){\
-        _free_synth(ro);\
-    }\
+    if(dst != lo && ro != lo && !lo->flags.from_ht)\
+        _free_synth(lo, lo != dst);\
+    if(dst != ro)\
+        _free_synth(ro, ro != dst);\
     \
     return dst;
 
@@ -248,10 +251,9 @@ mementry *_add(opargs *args){
        
     /* free right operand if it's not the entry to be returned */
     if(dst != lo && ro != lo)
-        _free_synth(lo);
-    if(dst != ro){
-        _free_synth(ro);
-    }
+        _free_synth(lo, lo != dst);
+    if(dst != ro)
+        _free_synth(ro, lo != dst);
     
     return dst;
 
@@ -303,12 +305,9 @@ mementry *_equ(opargs *args){
        
     /* free right operand if it's not the entry to be returned */
     if(dst != lo && !lo->flags.from_ht)
-        free(lo);
-    if(dst != ro){
-        _free_synth(ro);
-        if(!ro->flags.from_ht)
-            free(ro);
-    }
+        _free_synth(lo, lo != dst);
+    if(dst != ro)
+        _free_synth(ro, lo != dst);
     
     return dst;
 }
@@ -374,10 +373,6 @@ mementry *_ass(opargs *args){
     // reset flags
     dst->flags.val_tmp = 0;
 
-    // free src if synthetic
-    if(!src->flags.from_ht)
[…]

### 2022-10-29 src/interpreter/processing/implementations.h

@@ -8,7 +8,7 @@
 // a reference from hashtable is returned (and element is allocated)
 //  else, a copy is returned
 mementry *_recursiveprocess(opargs *args, char return_reference);
-void _free_synth(mementry *entry);
+void _free_synth(mementry *entry, char free_entry);
 char truth_of(mementry *entry);
 
 mementry *call(mementry *fun, mementry *params, errorinfo *info);

### 2022-10-29 src/interpreter/runtime.c

@@ -87,4 +87,6 @@ void process(bytecode *code, errorinfo *info) {
     _indent = 0;
     _print_value(dst);
     #endif
+
+    _free_synth(dst, 1);
 }

### 2022-10-29 src/interpreter/tokenizer.c
fix tokenizer symbol issue

@@ -221,15 +221,14 @@ void readtkn(cache *cache) {
             default:
                 // set type
                 dst->type = SYMBOL;
-                char last, check = 0;
 
                 // read all following punctuation characters
                 while (ispunct(cur) && cur != '\'' && cur != '"') {
-                    if(check && (last == '@' || cur == '@'))
+                    // check if cur is a bracket
+                    if(strrchr("(){}[]", cur))
                         break;
-                    *con++ = last = cur;
+                    *con++ = cur;
                     cur = next(cache);
-                    check = 1;
                 };
 
                 // push back last read character if not space

### 2022-10-29 src/interpreter/bytecode.c
try fixing memory leaks

@@ -16,6 +16,19 @@ static inline mementry *_register(bytecode *code, stackptr ptr, typing type) {
     return ret;
 }
 
+void free_bytecode(bytecode *code, char flags) {
+    // free the mementry elements
+    mementry *memptr;
+    size_t offset = code->elements;
+    while(offset != 0) {
+        memptr = *(mementry **)pop(code, &offset, sizeof(mementry *));
+        _free_synth(memptr, flags);
+    }
+
+    // free the stack itself
+    free_stack(code);
+}
+
 // TODO: does not work for multi-thread
 table_level max_level;
 

### 2022-10-29 src/interpreter/bytecode.h

@@ -50,3 +50,6 @@ typedef stack bytecode;
 
 // Consumes a syntax tree (freeing it) and a hashtable and returns the corresponding bytecode 
 bytecode *consume(stnode *root);
+
+// free bytecode stack and all elements
+void free_bytecode(bytecode *code, char flags);

### 2022-10-29 src/interpreter/interpreter.c

@@ -78,7 +78,8 @@ void interpret(FILE *stream, char* filename) {
         // process bytecode, also passing debug information from parser
         process(code, &cac.info);
 
-        free_stack(code);
+        free_bytecode(code, IGNORE_PERSISTENCE);
+
     };
 
     #if DEBUG
@@ -96,6 +97,7 @@ void interpret(FILE *stream, char* filename) {
 
     // free token content
     free_stack(cac.bracketstack);
+    free_hashtable(memory);
     free(tkn.content);
 }
 

### 2022-10-29 src/interpreter/memory/hashtable.c

@@ -1,6 +1,7 @@
 #include <string.h>
 
 #include "hashtable.h"
+#include "array.h"
 #include "../bytecode.h"
 
 #define DEFAULT_ENTRY_SIZE 8
@@ -24,6 +25,32 @@ mementry *_alloc_mementry() {
         {.from_ht= 0,.val_tmp=0,.persistent=0};
     return ret;
 }
+
+// free a synthetic mementry and its synthetic value
+void _free_synth(mementry *entry, char free_entry) {
+    if((!entry->flags.from_ht || free_entry & IGNORE_HASHTABLE) 
+            && (!entry->flags.persistent || free_entry & IGNORE_PERSISTENCE)) {
+        if(entry->flags.val_tmp){
+            switch(entry->type){
+                case UNDEFINED:
+                    break;
+                case ARRAY:
+                case TUPLE:
+                    free_array((array *) entry->value);
+                    break;
+                case FUNCTION:
+                    free_bytecode((bytecode *)entry->value, IGNORE_PERSISTENCE | IGNORE_HASHTABLE);
+                    break;
+                default:
+                    free(entry->value);
+                    break; 
+            } 
+    if(free_entry)
+        free(entry);
+        }
+    }
+}
+
 
 hashtable *create_hashtable(size_t width, size_t cache_size) {
     hashtable *ret = (hashtable *) malloc(sizeof(hashtable) + width * sizeof(tableslice *));
@@ -60,7 +87,7 @@ static inline void _free_tableentry(tableentry *entry) {
     // TODO: free the mementries / data values
     // free(entry->entry->value);
     
-    free(entry->entry);
+    _free_synth(entry->entry, IGNORE_PERSISTENCE | IGNORE_HASHTABLE | 1);
     free(entry);
 }
 

### 2022-10-29 src/interpreter/memory/hashtable.h

@@ -51,6 +51,9 @@ typedef struct {
 
 // function to allocate mementry
 mementry *_alloc_mementry();
+#define IGNORE_PERSISTENCE 2
+#define IGNORE_HASHTABLE 4
+void _free_synth(mementry *entry, char free_entry);
 
 // internally used hash types
 typedef uint32_t _H_HASH;

### 2022-10-29 src/interpreter/processing/implementations.c

@@ -30,29 +30,6 @@ char truth_of(mementry *entry) {
     return 1;
 }
 
-// free a synthetic mementry and its synthetic value
-void _free_synth(mementry *entry, char free_entry) {
-    if(!entry->flags.from_ht && !entry->flags.persistent) {
-
-        if(entry->flags.val_tmp){
-            switch(entry->type){
-                case UNDEFINED:
-                    break;
-                case ARRAY:
-                case TUPLE:
-                    free_array((array *) entry->value);
-                    break;
-                case FUNCTION:
-                    free_stack((bytecode *)entry->value);
-                    break;
-                default:
-                    free(entry->value);
-                    break; 
-            } 
-        }
-    }
-}
-
 mementry *_recursiveprocess(opargs *args, char flags) {
     // get the next mementry from the bytecode
     mementry *memptr = *(mementry **)pop(args->code, &args->offset, sizeof(mementry *));
@@ -84,9 +61,11 @@ mementry *_recursiveprocess(opargs *args, char flags) {
         // only return the value, if it's a mutable field
         if(memptr->type == REFERENCE) {
             memptr = memptr->value;
-            if(memptr->flags.from_ht)
+            if(memptr->flags.from_ht || memptr->flags.persistent)
                 return memptr;
         }
+        _free_synth(memptr, 1);
+        // only return null in this case
         return NULL;
     }
 
@@ -137,10 +116,12 @@ mementry *_decr(opargs *args) {
 }
 
 mementry *_lnot(opargs *args) {
-    mementry *dst = _recursiveprocess(args, MUTABLE); // Load only operand into mementry
-    if(dst == NULL){
-        throw(EI_REQ_MUTABLE, args->info);
-        return NULL;
+    mementry *src = _recursiveprocess(args, DEREFERENCE); // Load only operand into mementry
+    mementry *dst;
+    if(src->flags.from_ht || src->flags.persistent){
+        dst = _alloc_mementry();
+        dst->flags.val_tmp = 1;
+        dst->value = malloc(sizeof(number));
     }
 
     if(dst->type != NUMBER){
@@ -153,10 +134,12 @@ mementry *_lnot(opargs *args) {
 }
 
 mementry *_bnot(opargs *args) {
-    mementry *dst = _recursiveprocess(args, MUTABLE); // Load only operand into mementry
-    if(dst == NULL){
[…]

### 2022-10-29 src/interpreter/processing/implementations.h

@@ -8,7 +8,6 @@
 // a reference from hashtable is returned (and element is allocated)
 //  else, a copy is returned
 mementry *_recursiveprocess(opargs *args, char return_reference);
-void _free_synth(mementry *entry, char free_entry);
 char truth_of(mementry *entry);
 
 mementry *call(mementry *fun, mementry *params, errorinfo *info);

### 2022-10-29 src/interpreter/localizer.c
fix localizer.c hashtable level for members

@@ -34,7 +34,7 @@ void localize(stnode *subtree, hashtable *table, errorinfo *info) {
 
             if(subtree->data.parent.left != NULL) {
                 // change ht level
-                // ht_up(table);
+                ht_up(table);
 
                 // localize child first
                 if(subtree->data.parent.right != NULL)
@@ -44,7 +44,7 @@ void localize(stnode *subtree, hashtable *table, errorinfo *info) {
                 localize(subtree->data.parent.left, table, info);
                 
                 // reset ht level
-                // ht_down(table);
+                ht_down(table);
             }
 
             break;

### 2022-10-30 src/interpreter/mappings/mapop.c
improve stdin; remove unused operators

@@ -1,6 +1,6 @@
 /* ANSI-C code produced by gperf version 3.1 */
 /* Command-line: gperf --output-file=./src/interpreter/mappings/mapop.c ./mapop.gperf  */
-/* Computed positions: -k'1-2' */
+/* Computed positions: -k'1' */
 
 #if !((' ' == 32) && ('!' == 33) && ('"' == 34) && ('#' == 35) \
       && ('%' == 37) && ('&' == 38) && ('\'' == 39) && ('(' == 40) \
@@ -45,12 +45,12 @@
 #line 23 "./mapop.gperf"
 struct s_op;
 
-#define TOTAL_KEYWORDS 26
+#define TOTAL_KEYWORDS 23
 #define MIN_WORD_LENGTH 1
 #define MAX_WORD_LENGTH 2
 #define MIN_HASH_VALUE 1
-#define MAX_HASH_VALUE 62
-/* maximum key range = 62, duplicates = 0 */
+#define MAX_HASH_VALUE 32
+/* maximum key range = 32, duplicates = 0 */
 
 #ifdef __GNUC__
 __inline
@@ -64,45 +64,34 @@ hash (register const char *str, register size_t len)
 {
   static unsigned char asso_values[] =
     {
-      63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
-      63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
-      63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
-      63, 63, 63,  8, 63, 63, 63, 14, 30, 63,
-      63, 63, 31, 25, 26,  5, 63,  6, 63, 63,
-      63, 63, 63, 63, 63, 63, 63, 63, 28, 23,
-       3,  0, 15, 63, 18, 63, 63, 63, 63, 63,
-      63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
-      63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
-      63, 63, 63, 63, 13, 63, 63, 63, 63, 63,
-      63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
-      63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
-      63, 63, 63, 63, 20, 63, 10, 63, 63, 63,
-      63, 63, 63, 63, 63, 63, 63, 63, 63, 63,
[…]

### 2022-10-30 src/interpreter/mappings/mapop.h

@@ -15,12 +15,11 @@ typedef enum {
     OP_DIV, OP_MOD, OP_ADD, OP_SUB, OP_LET, 
     OP_LEQ, OP_GRT, OP_GEQ, OP_EQU, OP_NEQU, 
     OP_BAND, OP_BXOR, OP_BOR, OP_LAND, OP_LOR, 
-    OP_LAMBDA, OP_ASS, OP_HARDSET,
-    OP_MULASS, OP_DIVASS, OP_MODASS, 
-    OP_LIST, OP_END, OP_ARGS,
+    OP_ASS, OP_MULASS, OP_DIVASS, OP_MODASS, 
+    OP_LIST, OP_ARGS,
 
     // Special prefix operators
-    OP_PREINCR, OP_PREDECR, OP_POS, OP_NEG,
+    OP_POS, OP_NEG,
 
     // Internal operators
     OP_CALL, OP_INDEX, OP_ARRAY,

### 2022-10-30 src/interpreter/mappings/operations.c

@@ -30,14 +30,9 @@ operation implementationof(opcode code) {
             [OP_BOR]       = _bor,
             [OP_LAND]      = _land,
             [OP_LOR]       = _lor,
-            [OP_LAMBDA]    = _lambda,
             [OP_ASS]       = _ass,
-            [OP_HARDSET]   = _hardset,
             [OP_LIST]      = _list,
-            [OP_END]       = _end,
 
-            [OP_PREINCR]   = _preincr,
-            [OP_PREDECR]   = _predecr,
             [OP_POS]       = _pos,
             [OP_NEG]       = _neg,
 

### 2022-10-30 src/interpreter/parser.c

@@ -117,11 +117,12 @@ stnode *expr(cache *cache, unsigned char rbp) {
     repeat:
 
     if(cache->cur->type == SYMBOL) {
-        operator *op;
+        operator *op = 0;
         stnode *new;
 
         while(
             cache->cur->type == SYMBOL &&
+            cache->cur->content[0] != ';' &&
             (op = mapop(cache->cur->content)) != 0 &&
             op->precedence > rbp
         ) {
@@ -136,15 +137,15 @@ stnode *expr(cache *cache, unsigned char rbp) {
                 new->data.parent.right = expr(cache, op->precedence - op->associativity);
             } else {
                 new->data.parent.right = NULL;
-
-                // return on semicolon
-                if(op->opcode == OP_END)
-                    return left;
             }
 
             left = new;
         }
         if(op == 0) {
+            // return on semicolon (without advance!)
+            if(cache->cur->content[0] == ';')
+                return left;
+
             // free already allocated nodes
             free_stnode(left);
             
@@ -218,6 +219,11 @@ stnode *expr(cache *cache, unsigned char rbp) {
 stnode *secondary(cache *cache, unsigned char rbp){
     // declare node to return
     stnode *ret = NULL;
+
+    // clear left-over semicolon
+    if(cache->cur->type == SYMBOL && cache->cur->content[0] == ';')
+        advance(cache);
+
 
     switch(cache->cur->type) {
         case BRACKET:
@@ -349,6 +355,7 @@ stnode *secondary(cache *cache, unsigned char rbp){
             }
 
             if(ret->data.parent.op->position != PREFIX) {
+                printf("got operator %s which is not prefix\n", cache->cur->content);
                 throw(EP_EXPECTED_PREFIX_OPERATOR, &cache->info);
                 
                 ret->type = FILE_END;

### 2022-10-30 src/interpreter/processing/implementations.c

@@ -314,10 +314,6 @@ mementry *_land(opargs *args){ CALCULATE(a && b) }
 
 mementry *_lor(opargs *args){ CALCULATE(a || b) }
 
-mementry *_lambda(opargs *args){
-
-}
-
 mementry *_ass(opargs *args){
     mementry *src = _recursiveprocess(args, 0); // Load right operand into src
     mementry *dst = _recursiveprocess(args, 0); // Load left operand into dst
@@ -362,10 +358,6 @@ mementry *_ass(opargs *args){
     return dst;
 }
 
-mementry *_hardset(opargs *args) {
-    // not implemented 
-}
-
 mementry *_list(opargs *args){
     mementry *right = _recursiveprocess(args, DEREFERENCE);
     mementry *left = _recursiveprocess(args, DEREFERENCE);
@@ -403,16 +395,6 @@ mementry *_list(opargs *args){
         ret->value = arr;
         return ret;
     }
-}
-
-mementry *_end(opargs *args){
-}
-
-
-mementry *_preincr(opargs *args){
-}
-
-mementry *_predecr(opargs *args){
 }
 
 mementry *_pos(opargs *args){

### 2022-10-30 src/interpreter/processing/implementations.h

@@ -33,14 +33,9 @@ mementry *_bxor(opargs *args);
 mementry *_bor(opargs *args);
 mementry *_land(opargs *args);
 mementry *_lor(opargs *args);
-mementry *_lambda(opargs *args);
 mementry *_ass(opargs *args);
-mementry *_hardset(opargs *args);
 mementry *_list(opargs *args);
-mementry *_end(opargs *args);
 
-mementry *_preincr(opargs *args);
-mementry *_predecr(opargs *args);
 mementry *_pos(opargs *args);
 mementry *_neg(opargs *args);
 

### 2022-10-30 src/interpreter/tokenizer.c

@@ -223,11 +223,17 @@ void readtkn(cache *cache) {
                 dst->type = SYMBOL;
 
                 // read all following punctuation characters
-                while (ispunct(cur) && cur != '\'' && cur != '"') {
-                    // check if cur is a bracket
-                    if(strrchr("(){}[]", cur))
-                        break;
+                while (ispunct(cur) 
+                        && cur != '\'' && cur != '"' 
+                        && !strrchr("(){}[]", cur)) {
                     *con++ = cur;
+
+                    // break on semicolon
+                    if(cur == ';'){
+                        cur = 0;
+                        break;
+                    }
+
                     cur = next(cache);
                 };
 

### 2022-10-30 paper.tex
continue writing paper

@@
 \CatchFileDef{\files}{paper/files.txt}{\endlinechar=-1}
+% Get current git hash
+\CatchFileDef{\githash}{paper/githash.txt}{\endlinechar=-1}
 
@@
 \abstract{
-    This is an abstract...
+    This paper is about creating a programming language and an accompanying
+    interperter. The goal is to learn about algorithms and techniques already
+    popular in the industry, while still creating something unique.
+
+    The source code of the interpreter can be found on GitHub at
+    \url{https://github.com/LeFlorians/TestScript}.
+
+    The source code as of the latest commit before hand-in date of the paper
+    is available at
+    \url{https://github.com/LeFlorians/TestScript/tree/\githash{}}.
 }
@@
 Those serve as a bridge between the human mind and its imagination, and
-the capabilities of computer hardware and are a huge time-saver when compared
+the capabilities of computer hardware, and are a huge time-saver when compared
 to writing machine-native code, i.e. not using a programming language.
+
 By making use of just the term \emph{language} this paper will refer to programming
@@
 has to be understood by the computer hardware in order to run. It can either
-be translated to machine-native bytecode with the help of a compiler, or
+be translated to machine-native code with the help of a compiler, or
 read line-by-line and interpreted by another program known as interpreter.
@@
 In computing, there are several types of interpreters that behave differently.
-Here are some examples and their use-cases:
+Here are a few examples and their use-cases:
 \begin{itemize}
@@
 supports more languages than solely C. Substantiating the development process is 
-\emph{GNU Make}, a general-purpose build system. It handles tasks like building
-the final executable program step-by-step from the project source code and directory clean-up.
+\emph{GNU Make}, a general-purpose build system. It simplifies repetitive tasks,
+removing the need to write and remember long build commands.
 The C ecosystem is very easy to use on and fundamentally supported by 
@@
         .2 pseudocode.
-            .3 \hrc{pseudocode/parser.c}.
+            .3 \hrc{pseudocode/pratt_parser.c}.
         .2 examples.
@@
 
-% TODO: say sth about README.md and expand it
-
-The file mapop.gperf is also a source file. It is used in conjunction with the GNU gperf
+% TODO: say sth about README.md and expand it, makefile
+The file named \hrc{Makefile} carries information for the program \emph{GNU make}.
+It contains built targets, which are specifications on how to build and compile
+the C source code down to the final executable interpreter. 
+These build targets contain shell commands, which would otherwise have to be
+typed out in their entirety every time the project should be built. 
+[\cte{makefile_tutorial}]
+
+A short guide on how to build the \name{} interpreter can be found in 
+the file \hrc{README.md}. The file name is spelled in capital letters by convention.
+It tells users about
+which programs have to be preinstalled to compile the project and which \emph{GNU make}
+commands are used to build or clean up the project.
+It also contains some information about all the available standard library
+functions that are included.
+
+The file \hrc{mapop.gperf} is also a source file. It is used in conjunction with the GNU gperf
 tool. The tool creates a perfect hash function, which is output to the file
-src/interpreter/mappings/mapop.c. The perfect hash function is used to translate
+\hrc{src/interpreter/mappings/mapop.c}. The perfect hash function is used to translate
 single- or multi-character symbols, such as '!=' and '+', to their respective
@@
 the problem of time overhead when traversing the syntax tree.
-
 The final step is to process or execute said bytecode. This step is performed
@@
 % TODO: validate code
-\code{pseudocode/parser.c}{1}{999}
+\code{pseudocode/pratt_parser.c}{1}{999}
 
@@
 
-Not to be confused with machine-native bytecode, the bytecode compiler in \name{}
+Not to be confused with machine-native code, the bytecode compiler in \name{}
 generates byte-sequences in an internal format that are then interpreted. A
 Just-In-Time compiler, JIT compiler for short, would instead compile the AST 
-down to machine-native bytecode, which
+down to machine-native code, which
 runs directly on the system's hardware and does not have to be interpreted.

### 2022-10-30 paper.tex
continue writing paper

@@
 \CatchFileDef{\files}{paper/files.txt}{\endlinechar=-1}
-% Get current git hash
 \CatchFileDef{\githash}{paper/githash.txt}{\endlinechar=-1}
@@
         firstnumber=#2,firstline=#2,lastline=#3]{#1}
-
     \end{minipage}
+
     }
@@
 
-    The source code of the interpreter can be found on GitHub at
+    \vfill
+    \centerline{\begin{minipage}{\linewidth}
+
+    The source code of the interpreter can also be found on GitHub:
+
     \url{https://github.com/LeFlorians/TestScript}.
 
-    The source code as of the latest commit before hand-in date of the paper
-    is available at
-    \url{https://github.com/LeFlorians/TestScript/tree/\githash{}}.
+    The source code as of the latest commit before the submission of the paper
+    is available at:
+
+    \url{https://github.com/LeFlorians/TestScript/tree/\githash}.
+
+    \end{minipage}}
+
 }
@@
 {
+    \newpage
     \hypersetup{hidelinks}
@@
             .3 \hrc{pseudocode/pratt_parser.c}.
+            .3 \hrc{pseudocode/function.c}.
         .2 examples.
             .3 \hrc{examples/mandelbrot.nts}.
+            .3 \hrc{examples/montecarlo.nts}.
             .3 \hrc{examples/fibonacci.nts}.
@@
 % TODO: validate code
-\code{pseudocode/pratt_parser.c}{1}{999}
+\code{pseudocode/pratt-parser.c}{1}{999}
 
@@
 \ver{
-\foreach \f in \files{\pagelabel{\f}\sourcecode{\f}{1}{9999}}
+\foreach \fname in \files{\pagelabel{\fname}\code{\fname}{1}{9999}}
 }

### 2022-10-30 paper.tex
continue writing paper

@@
 \usepackage{etoolbox}
-\usepackage{times} % Use Times New Roman font
+\usepackage{mathptmx} % Use Times New Roman font
 \usepackage[style=numeric,backend=biber]{biblatex}
@@
 
-\author{Florian Donnelly}
+\author{Florian M. Donnelly}
 \title{The Creation of a Programming Language}
@@
 \definecolor{mygreen}{rgb}{0.3, 0.5, 0.2}
-\definecolor{myteal}{rgb}{0.1, 0.2, 0.3}
+\definecolor{myteal}{rgb}{0.1, 0.3, 0.5}
 \definecolor{myblue}{rgb}{0.2, 0.0, 0.6}
 \definecolor{mypink}{rgb}{0.8, 0.3, 0.6}
-\definecolor{myorange}{rgb}{0.7, 0.4, 0.1}
+\definecolor{myorange}{rgb}{0.4, 0.2, 0.0}
 
@@
     colorlinks=true,
-    pdfborder={3 3 .2},
     citecolor=mygreen,
@@
     urlcolor=myblue,
+}
+
+% Define language for listings
+\lstdefinelanguage{tslang}{
+  keywords={std},
+  morecomment=[l]{//},
+  morecomment=[s]{/*}{*/},
+  morestring=[b]',
+  morestring=[b]",
+  ndkeywords={},
+  sensitive=true
 }
@@
 
-\newcommand{\sourcecode}[3] {
+\newcommand{\sourcecode}[4] {
+    \pagelabel{#1}
+    \lstinputlisting[language=#4,numbers=left,breaklines=true,
+    frame=single,title=#1,firstnumber=#2,firstline=#2,lastline=#3]{#1}
+}
+
+\newcommand{\code}[4] {
     \ver{
-        \lstinputlisting[language=C,numbers=left,breaklines=true,
-        frame=single,title=#1,firstnumber=#2,firstline=#2,lastline=#3]{#1}
-    }
-}
-
-\newcommand{\code}[3] {
-    \ver{
-
-    \begin{minipage}{\linewidth}
-    \lstinputlisting[language=C,numbers=left,breaklines=true,
+
+    \centerline{\begin{minipage}{\linewidth}
+    \lstinputlisting[language=#4,numbers=left,breaklines=true,
         frame=single,title=\hr{#1}{#1},
         firstnumber=#2,firstline=#2,lastline=#3]{#1}
-    \end{minipage}
+    \end{minipage}}
 
@@
     \vspace{1.5cm}
-    Matura Paper by\\ \textbf{Florian Donnelly}
+    Matura Paper by\\ \textbf{Florian M. Donnelly}
 
@@
 \abstract{
-    This paper is about creating a programming language and an accompanying
-    interperter. The goal is to learn about algorithms and techniques already
-    popular in the industry, while still creating something unique.
+    This project is about creating a programming language and an accompanying
+    interpreter. The goal is to learn about algorithms and techniques already
+    popular in the industry while still creating something unique.
+    The paper will explain the tools and procedures used to implement
+    the different entities that make up an interpreter. 
+    The interpreter is written in plain C in the gnu18 dialect with
+    the only library used being libc, the C standard library.
 
     \vfill
+
+    By making use of just the term \emph{language} this paper will refer to programming
+    languages specifically. Sources are mentioned after informationally coherent paragraphs in
+    square brackets, an index can be found in the Bibliography on page
+    \pageref{bibliography}.
+
+    \vspace{2cm}
+
     \centerline{\begin{minipage}{\linewidth}
@@
 {
+    \small
     \newpage
@@
 
+\normalsize
+\onehalfspacing
 \section{Introduction}
@@
 to writing machine-native code, i.e. not using a programming language.
-
-By making use of just the term \emph{language} this paper will refer to programming
-languages specifically. Sources are mentioned after cohesive paragraphs in
-square brackets, an index can be found in the \emph{Bibliography} on page
-\pageref{bibliography}.
 
@@
 \end{itemize}
+[Source \cte{interpreters}]
+
 This paper will implement a mixture form of an AST compiler and bytecode
@@
 used many times, also leading to a performance benefit.
-[\cte{V8}, \cte{jvm}, \cte{rustc}]
+[Sources \cte{V8}, \cte{jvm}, \cte{rustc}]
 
@@
 \emph{GNU Make}, a general-purpose build system. It simplifies repetitive tasks,
-removing the need to write and remember long build commands.
+removing the need to write and remember long build commands. Building a project
+means to convert its source code into an executable program.
 The C ecosystem is very easy to use on and fundamentally supported by 
@@
 several times in the following sections of this paper. In the depicted file tree
-only source files are shown. No build files, executable files or files that
+only source- or configuration files are shown. No output files, executable files or files that
 are of no further importance regarding \name are depicted.
@@
         .2 \hrc{README.md}.
-        .2 \hrc{Makefile}.
+        .2 \hr{Makefile}{Makefile}.
         .2 \hrc{mapop.gperf}.
         .2 pseudocode.
-            .3 \hrc{pseudocode/pratt_parser.c}.
+            .3 \hrc{pseudocode/pratt-parser.c}.
             .3 \hrc{pseudocode/function.c}.
@@
 % TODO: say sth about README.md and expand it, makefile
-The file named \hrc{Makefile} carries information for the program \emph{GNU make}.
-It contains built targets, which are specifications on how to build and compile
-the C source code down to the final executable interpreter. 
+The file named \hr{Makefile}{Makefile} carries information for the program \emph{GNU make}.
+It contains built targets, which are specifications on how to compile
+certain parts of the project. When combined, these targets can build the
+entire project, taking the source code and compiling it to machine-native code.
 These build targets contain shell commands, which would otherwise have to be
 typed out in their entirety every time the project should be built. 
-[\cte{makefile_tutorial}]
+[Source \cte{makefile_tutorial}]
 
@@
 It also contains some information about all the available standard library
-functions that are included.
+functions that are included. The standard library is a set of function
+included with the interpreter that extend its functionality over the
+available operators.
 
@@
 interpreter internal representation in constant time.
-[\cte{c_compilation}, \cte{gperf}]
+[Sources \cte{c_compilation}, \cte{gperf}, \cte{perfect_hash}, \cte{hash_functions}]
 
@@
 the \name{} interpreter. It is therefore the most complicated and detailed section
-in this paper. Examples will be made with both pseudo code and actual source
+in this paper. Examples will be made with both pseudo code, i.e. code for
+illustration purposes, and actual, working source
 code from the project to show the decisions made during
@@
 C application. Since the file is so small, here is its source code:
-\code{src/main.c}{1}{999}
+\code{src/main.c}{1}{999}{C}
 
@@
         Here is an example of a C function:
-        \code{pseudocode/function.c}{1}{999}
+        \code{pseudocode/function.c}{1}{999}{C}
         This would be an example of a main function, similar to the one
@@
 on page \pageref{bibliography}.
-[\cte{freeCodeCamp}]
+[Source \cte{freeCodeCamp}]
 
@@
 to find exact matches.
-[\cte{separate_chaining}]
+[Source \cte{separate_chaining}]
 
@@
 While tokenizers can be programmed using regular expressions, 
-this paper uses a simpler finite-state machine tokenizer.
+this paper uses a simpler deterministic finite-state machine tokenizer.
 It reads character by character, determining the token type as it goes.
@@
 section. 
+[Source \cte{dfsm}]
 
@@
 \pageref{bibliography} which this paper's parser takes inspiration from.
-[\cte{pratt}, \cte{devio}]
+[Sources \cte{pratt}, \cte{devio}, \cte{parsing_guide}]
 
@@
 
+\centerline{\begin{minipage}{\linewidth}
 Following is a piece of pseudo code to better illustrate the inner workings of the
-parser which is based on a Pratt-parser. 
+parser which is based on a Pratt-Parser. 
 This is only an example and not completely accurate to the actual parser used
@@
 % TODO: validate code
-\code{pseudocode/pratt-parser.c}{1}{999}
+\code{pseudocode/pratt-parser.c}{1}{999}{C}
+\end{minipage}}
 
@@
 where operands surround the operator.
-[\cte{shunting_yard}]
+[Source \cte{shunting_yard}]
 
@@
 \end{enumerate}
-[\cte{polish_notation}]
+[Source \cte{polish_notation}]
 
@@
 representing all possible operations. Extending upon that, there are some
-standard functions that enlarge the capabilities of the language.
-
-\subsection{Syntax}
+standard functions that are included in the interpreter and that enlarge the capabilities of the language.
+The following sections make up examples of the syntax and capabilities of the language and
+how programs are written in it.
+
+\subsection{The Fibonacci Sequence}
 The syntax of \name{} is unlike most modern, common programming languages' syntax,
@@
 of the Fibonacci sequence, as well as a dissection of how that same code is interpreted.
-\code{examples/fibonacci.nts}{1}{999}
+\code{examples/fibonacci.nts}{1}{999}{tslang}
 \begin{itemize}
@@
 Here is another example, which in contrast to the example above, focuses on an
-original security concept implemented in \name{}.
-\code{examples/exec.nts}{1}{999}
-The code begins by creating the variable \emph{var}, which holds a string and
-\emph{f}, which calls the function \emph{std.exec} with var as argument.
-Line number seven holds a comment, indicated by the two slashes at the start of the line.
-Comments are ignored by the \name{} interpreter and their purpose is to contain notes
-and explanations inside the source code itself. Another type of commends are block-comments
-which span across multiple lines, such as on lines 9 to 12.
-On line seven, the \emph{std.exec} function is called for the first time. It runs the
-command it is given as argument in the user's shell, which is most commonly \emph{cmd.exe} on
-Microsoft Windows and \emph{bash} on most Linux distributions. In this example, \emph{std.exec}
-launches the Firefox browser, but it is capable of all kinds of things that can otherwise be
-done in the shell, such as delete files or shut down the computer.
-
-That is why in most cases, the \emph{std.exec} function can be a risk waiting to be exploited
-by malicious \name{} programs. To solve this problem, there is the hashtag-operator, whose purpose
-it is to delete access to a variable through the specified name. It is used on line 13 of the
-example to remove access to the \emph{std.exec} function, which is why the call to that function
-on line 16, trying to shut down the computer, will result in an error. Unexpectedly though, the
-call on line 20 to the function \emph{f} which was created as an as to \emph{std.exec} will work.
-
-This security concept enforces the use of a top-down programming model. First, safe to use aliases
-to dangerous functions have to be created. Then, access to all those exploitable functions is
-revoked. This way, running unknown code can only access data and functions that are being provided.
-This contrasts the models of languages like Python. In Python, each separate source file can
-make use of the \emph{import} keyword to include other files. Files imported this way do not
-have access or knowledge about the variables that exist in the file that includes them.
-In \name{}, each imported file would instead have access to all variables and functions that
-are made available to it, instead of having to include or import them separately.
-This leads to another problem, but that is what the next section will be all about.
-
-% TODO: link more examples
+original concept implemented in \name{}.
+\code{examples/exec.nts}{1}{999}{tslang}
+The code begins by creating the variable \emph{f}.
+The function it holds is a wrapper around the standard library function \emph{std.exec}.
+It executes a command in the user's shell. A shell is a program that
+lets users interact with the operating system, most often through a text-based interface.
+Common shells include \emph{cmd.exe}
+on Microsoft Windows or \emph{GNU Bash} on most Linux distributions.
+The command that will be executed in the shell is \emph{'echo Hello There!'},
+which will simply output the text 'Hello There!'.
+As the comment on line 5 explains, the function call on line 6 of the example
+program opens a YouTube video in Firefox. Because the start command is not
+available in \emph{GNU Bash} this instruction will most likely not work
+on GNU+Linux.
+Because \emph{std.exec} could be used for malicious purposes, such as
+shutting down the computer or deleting important files, access to that
+function can be revoked via a hash symbol as on line 12. The instruction
+on line 15 that is supposed to turn off the machine will therefore not work
+anymore. 
+The interesting thing is that the function call to \emph{f} in line 18
+will still work. This is because a hash symbol does not simply delete
+a variable, but rather make it unusable for future references. To sum up,
+the function \emph{f} is still capable of using \emph{std.exec}, because
+at the time \emph{f} was defined, \emph{std.exec} was still accessible.
+
+This security concept forces developers to think about access restriction. First, safe to use aliases
+to dangerous functions have to be created. Next, access to all those exploitable functions is
+revoked. This way, any code can only access variables and functions that are still available.
 
@@
 
-\subsection{Distribution}
-% TODO
-
 \subsection{Limitations}\label{Limitations}
@@
 Taking C as a reference, the datatype \emph{char} is exactly one byte in size, and
-can therefore be in exactly 2\^8 or 256 different states. But in C, information about
+can therefore be in exactly $2^8$ or 256 different states. But in C, information about
 the datatype can be discarded at compile time, whereas in an interpreter like \name{} such
@@
 project to find errors in the \name{} implementation.
-[\cte{npm}, \cte{cargo}]
+[Sources \cte{npm}, \cte{cargo}]
 
@@
 and it can be compiled to machine-native programs via LLVM. 
-[\cte{kotlin_compile}]
-
-% TODO: write about LLVM and recommend it lol6
+[Source \cte{kotlin_compile}]
 
@@
 out. This section of the paper gives insight on interesting ideas or concepts
-that are interesting and are used in other programming languages, but 
+that are used in other programming languages, but 
 could not be implemented into \name{}.
@@
 Python for this purpose, can be used to load and call such foreign functions.
-[\cte{FFI}, \cte{JNI}, \cte{ctypes}, \cte{libffi}]
+[Sources \cte{FFI}, \cte{JNI}, \cte{ctypes}, \cte{libffi}]
 
@@
 holding program instructions can be created and modified and later executed.
-[\cte{reflection}]
+[Source \cte{reflection}]
 
@@
 the inner workings of an interpreter and
-to get to know the C language has been acheived.
+to get to know the C language has been achieved.
 
@@
 on \url{https://app.diagrams.net}.
-[\cte{drawio}]
+[Source \cte{drawio}]
 
 \section{Declaration of Authenticity}
-Hereby I, Florian Donnelly, declare to have written this paper independently, 
+Hereby I, Florian M. Donnelly, declare to have written this paper independently, 
 also including the entire source code of \name{}, using resources 
@@
 See the file tree under \emph{File Structure} in section \ref{FileStructure}
-to get an overview.
+to get an overview. In the digital PDF version of this paper, all file names
+in the text and the file tree are clickable and will open the file in the appendix.
 
@@
 \ver{
-\foreach \fname in \files{\pagelabel{\fname}\code{\fname}{1}{9999}}
+    \foreach \fname in \files{\sourcecode{\fname}{1}{9999}{C}}
+    \sourcecode{Makefile}{1}{9999}{make}
+    \sourcecode{README.md}{1}{9999}{TeX}
+    \sourcecode{.gitignore}{1}{9999}{tslang}
+    \sourcecode{mapop.gperf}{1}{9999}{tslang}
+    \sourcecode{examples/mandelbrot.nts}{1}{9999}{tslang}
+    \sourcecode{examples/montecarlo.nts}{1}{9999}{tslang}
+    \sourcecode{examples/fibonacci.nts}{1}{9999}{tslang}
+    \sourcecode{examples/exec.nts}{1}{9999}{tslang}
 }

### 2022-10-30 paper.tex
fix spelling in paper

@@
     the different entities that make up an interpreter. 
-    The interpreter is written in plain C in the gnu18 dialect with
+    The interpreter is written in plain C in the gnu18 dialect, with
     the only library used being libc, the C standard library.
@@
 be translated to machine-native code with the help of a compiler, or
-read line-by-line and interpreted by another program known as interpreter.
+read line-by-line and interpreted by another program known as an interpreter.
 Such compilers or interpreters are programs on their own that are classified
 as programming language implementation. The focus of this paper will be to
-create such a programming language implementation, more specific an interpreter
-of the custom-made \name{} language.
+create such a programming language implementation,or more specific, an interpreter
+of the custom-designed \name{} language.
 
@@
         language, in which runs another interpreted for and written in said
-        language. Using a host language to initiate such system is called
+        language. Using a host language to initiate such a system is called
         bootstrapping.
@@
         Self-interpreters are closely related to self-hosting compilers. These
-        are in turn compilers which are written in the language they compile.
+        are compilers written in the language they compile.
         An example of a compiler written in itself is rustc, the Rust
@@
 This paper will implement a mixture form of an AST compiler and bytecode
-interpreter. The advantage is that an AST carries along lots of information
-not needed in later stages of interpretation which can be discarded. The AST
-is converted to an intermediary bytecode format which only contains
+interpreter. This has benefits, as an AST carries along lots of information
+that is not needed in later stages of interpretation and can be discarded.
+The AST is converted to an intermediary bytecode format which only contains
 necessary information about the program. Such bytecode can then be cached,
@@
 actual implementation of each of the steps can be heavily customized.
-Here is the problem, broken down into three steps as commonly done in other
-popular programming languages:
+Here is a possible approach on implementing an interpreter, broken
+down into three steps:
 \begin{enumerate}\pagelabel{simple_interpreter}
@@
 the \name{} interpreter. It is therefore the most complicated and detailed section
-in this paper. Examples will be made with both pseudo code, i.e. code for
+in this paper. Examples will be made with both pseudocode, i.e. code for
 illustration purposes, and actual, working source
@@
         The most common preprocessor instructions are the following.
-        On one hand there is \expr{\#include $<$header.h$>$}
+        On one hand, there is \expr{\#include $<$header.h$>$}
         to include various definitions from a header file, i.e. to paste
@@
     \item \emph{Pointers.} Pointers are variables that store memory addresses. 
-        In C syntax they can be defined with the star '*' symbol.
+        In C syntax they can be defined with the star (*) symbol.
         An example from \hrc{src/main.c} is \expr{FILE *input;.} That means that \emph{input} is not
@@
 The hash table implementation used in \hrc{src/interpreter/memory/hashtable.c}                   % TODO: maybe its own paragraph! interpolation search vs binary search
-uses binary search to locate elements variable-sized arrays in the hashtable's buckets.
-This is because for such few variables that are used in a standard \name{} program,
-an interpolation search algorithm is actually slower.
+uses binary search to locate elements variable-sized arrays in the hash table's buckets.
+This is because for such few variables as are used in a standard \name{} program,
+an interpolation search algorithm actually performs worse.
 For collision detection, i.e. to circumvent problems arising when two variable names have the
@@
 variable names to addresses in memory.
-SYMBOL type tokens are mapped to operators such as plus or minux by \hrc{src/implementation/mappings/mapop.c}
+SYMBOL type tokens are mapped to operators such as plus or minus by \hrc{src/implementation/mappings/mapop.c}
 during parsing. NUMBER type tokens are replaced by an actual number type, which
@@
 which stands for a value, e.g. a number or a text. Each internal node, also known as parent node, signifies
-an operation done on its children. This can be an addition of two numbers for example.
+an operation done on its children. This can be for example the addition of two numbers.
 
@@
 each operator.
-In the end the parse tree contains all values and 
+In the end, the parse tree contains all values and 
 operations as well as information about the order of execution.
@@
         It deals with simple values that make up the start of an expression, such as numbers,
-        fields, strings and it handles brackets. 
+        fields, strings, and it handles brackets. 
         In addition, it can handle prefix operators, such as the unary plus or unary minus.
-    \item \emph{expr.} The \emph{expr} function calls \emph{secondary} whose result acts
-        as the basis for a new expression. It then repeatedly checks
+    \item \emph{expr.} The \emph{expr} function calls \emph{secondary}, the result of which
+        acts as the basis for a new expression. It then repeatedly checks
         if an operator follows the secondary token and creates the syntax-tree
@@
 \centerline{\begin{minipage}{\linewidth}
-Following is a piece of pseudo code to better illustrate the inner workings of the
+Following is a piece of pseudocode to better illustrate the inner workings of the
 parser which is based on a Pratt-Parser. 
@@
         calculation? The answer is recursion.
-    \item The interpreter will read through the bytecode from left to right. First it will
+    \item The interpreter will read through the bytecode from left to right. First, it will
         notice the subtraction operator and accordingly assume that two
         operands must follow.
-    \item The function recursively calls itself in order to evaluate the
+    \item The function then recursively calls itself in order to evaluate the
         right operand and notice the 6, a value and a leaf node in the parse tree. 
@@
     \emph{b} is set to be the new \emph{a} minus \emph{b}, 
-    which is in turn the same as \emph{a} before the reassignment on line 7.
+    which is again the same as \emph{a} before the reassignment on line 7.
 \item On lines 10 and 11, the standard library function \emph{std.print} is used
@@
     that is print out signifies a so-called escape sequence and will be replaced with
-    piece of text that can not be entered directly. The \emph{n} stands
+    another piece of text that can not be entered directly. The \emph{n} stands
     for a line break, but other letters such as a \emph{t} for a tabulator are possible.
@@
 can share code. Examples would be \emph{npm} or \emph{cargo}, 
-the 'Node Package Manager' for JavaScript applications and and Rust's package 
+the 'Node Package Manager' for JavaScript applications and Rust's package 
 manager respectively.
@@
 language can be annoying for developers. A solution is to build upon and
-extend already existing languages and their tools with e.g. a new syntax.
+extend already existing languages and their tools with for example a new syntax.
 This is exactly what Kotlin does. Kotlin can produce Java Virtual Machine
-compatible bytecode, can be transcompiled, e.g. translated, into
+compatible bytecode, can be transcompiled, i.e. translated, into
 JavaScript, or ECMAScript 5.1 to be more specific,
@@
 would like to call some functions that are part of yet another C program or
-library. The problem lies in the fact that at runtime, C does not know
+library. The problem lies in the fact that at runtime, C does not know the
 structure of the functions to be called, say return- or argument-types, which
@@
 Most parts of the source code are annotated with comments, giving interested
-reader direct explanations of what is going on.
+readers direct explanations of what is going on.
 Although there are several limitations and shortcomings as stated in section \ref{Limitations}

### 2022-10-30 paper.tex
final touches

@@
 \author{Florian M. Donnelly}
-\title{The Creation of a Programming Language}
+\title{The Development of a Programming Language}
 
@@
 \lstset{basicstyle=\scriptsize,keywordstyle=\color{myorange},stepnumber=5,
-        frame=none,keepspaces=false,
+        frame=none,keepspaces=false,showstringspaces=false,
         rulecolor=\color{black},tabsize=2,
@@
     \lstinputlisting[language=#4,numbers=left,breaklines=true,
-    frame=single,title=#1,firstnumber=#2,firstline=#2,lastline=#3]{#1}
+    frame=single,title=Contents of the file #1:,firstnumber=#2,firstline=#2,lastline=#3]{#1}
 }
@@
 
-    \vspace*{1.5cm}
+    \vspace*{0.5cm}
     \Huge
-    \textbf{The Creation of a Programming Language}
+    \textbf{The Development of a Programming Language}
 
@@
     \Large
-    \vspace{1.5cm}
-    Matura Paper by\\ \textbf{Florian M. Donnelly}
+    \vfill
+    Matura Paper written by\\ \textbf{Florian M. Donnelly}
 
     \vspace{0.5cm}
-    With Supervision of\\ \textbf{Adrian Lüthi}
-
-    \vfill
+    Under Supervision of\\ \textbf{Adrian Lüthi}
+
+    \vspace{0.5cm}
     Gymnasium Burgdorf\\ October 31, 2022
@@
 
-\abstract{
-    This project is about creating a programming language and an accompanying
+\begin{abstract}
+    This project is about developing a programming language and an accompanying
     interpreter. The goal is to learn about algorithms and techniques already
@@
     The paper will explain the tools and procedures used to implement
-    the different entities that make up an interpreter. 
+    the different entities that make up the interpreter. 
     The interpreter is written in plain C in the gnu18 dialect, with
     the only library used being libc, the C standard library.
+
+    The project including its structure and source will be presented and
+    then be explained in more detail. A copy of the
+    entire source code is contained in the appendix of this paper, starting on
+    page \pageref{Appendix}, as well as on GitHub.com through the URL at the
+    bottom of this page.
 
@@
 
-}
+\end{abstract}
 
 {
+    \ver{
         \small
@@
         \tableofcontents\newpage
+    }
 }
@@
 Creating programs, no matter if games, web browsers or even operating systems,
-involve the use of programming languages.
+involves the use of programming languages.
 Those serve as a bridge between the human mind and its imagination, and
@@
 Such compilers or interpreters are programs on their own that are classified
-as programming language implementation. The focus of this paper will be to
-create such a programming language implementation,or more specific, an interpreter
+as programming language implementations. Subject of this paper will be to
+create such a programming language implementation, or more specific, an interpreter
 of the custom-designed \name{} language.
@@
 The focus of this project is learning about the internal design of interpreters
-by creating a custom interpreter named \name{} interpreter, building upon
-known strategies and algorithms used in existing programming languages.
-The goal of this paper is to further understand the inner workings of
-interpreters, stepping through human-readable source code and computing
-desired results.
+by creating a custom one that will be called the \name{} interpreter, building upon
+strategies and algorithms used in already existent programming languages.
 
@@
 the implementation. This includes dependency management tools to
-easily create projects extending upon existing ones, or a web platforms for
-people to share their code, all of which is not subject of this paper.
-The \name{} implementation will also not include many features seen in modern
-languages, as those are often created in years of time spent planning and
+easily create projects extending upon existing ones, or web platforms for
+people to share their code, all of which are not subject of this paper.
+The \name{} implementation will also not include many features seen in 
+languages popular to date, as those are often created in years of time spent planning and
 programming.
@@
         The most popular example would be the Java Virtual Machine,
-        a bytecode interpreter for java bytecode, also used by many other languages
+        a bytecode interpreter for JVM bytecode, also used by many other languages
         such as Kotlin, Scala or Groovy.
@@
         language they themselves are written in. This requires a program
-        written in another language running an interpreter for the wanted
-        language, in which runs another interpreted for and written in said
-        language. Using a host language to initiate such a system is called
+        written in a host language running an interpreter for the wanted
+        language, in which runs another interpreter, for and written in the
+        wanted language. Using a host language to initiate such a system is called
         bootstrapping.
@@
 e.g. kept in memory, meaning it only has to be converted once and can be
-used many times, also leading to a performance benefit.
+used many times, also leading to an improvement in performance.
 [Sources \cte{V8}, \cte{jvm}, \cte{rustc}]
@@
 means to convert its source code into an executable program.
-The C ecosystem is very easy to use on and fundamentally supported by 
+The ecosystem of C language related tools is fundamentally supported by 
 \emph{GNU+Linux}, the operating system used to program on.
@@
 The tools \emph{Valgrind} and \emph{gdb}, the GNU Project Debugger, were of help
-when trying to find of errors and memory leaks, see section \ref{memleaks} for details.
+when trying to find errors and memory leaks, see section \ref{memleaks} for details.
 \emph{Git} is the project management and version tracking software used to back up and share
@@
 
-\subsection{The Art of Simplification}
+\subsection{Simplyfing the Problem}
 When dealing with problems in general, it is helpful to map out and think about
@@
         in any thinkable, logical way.
-    \item To order the tokens into parse-tree, also called syntax-tree, as
+    \item To order the tokens into a parse-tree, also called syntax-tree, as
         specified by rules. These rules are called a grammar.
         Creating an order is important, as some calculations depend on others, e.g. multiplication
-        is always done before addition. An IT data structure that can accommodate
-        such requirements is the binary tree. A binary tree consists of tree-nodes
-        that each have a maximum of two other tree nodes as their children, and
-        a tree node whose child they are, the parent node.
+        is always done before addition. As the name suggests, a data structure
+        called tree, often times a binary tree, is used.
     \item To interpret the parse tree, traversing through it and following the
@@
 
-\subsection{A Project Overview}
+\newpage\subsection{A Project Overview}
 In the following paragraphs, the project in its entirety will be briefly
 presented, giving an overview over files and correlations.
+
+\subsubsection{Flowchart}
+
+\ver{
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\textwidth]{paper/Data Flow Diagram.png}
+    \caption{A \name{} Data Flow Diagram}
+    \label{f_diagram}
+\end{figure}
+}
+
+The above figure is a representation of how different components of the entire
+\name{} implementation interact and work together. Each next stage performs
+operations on the output data of the proper. The \hrc{src/interpreter/interpreter.c}
+file, here depicted as 'Interpreter', chains all stages together.
 
@@
 only source- or configuration files are shown. No output files, executable files or files that
-are of no further importance regarding \name are depicted.
+are of no further importance regarding \name{} are depicted.
 \ver{\begin{flushleft}
@@
 variable values are stored in the respective .c files.
-% TODO:explain C compilation or link example, fix hash citation below
-
-% TODO: say sth about README.md and expand it, makefile
+
 The file named \hr{Makefile}{Makefile} carries information for the program \emph{GNU make}.
@@
 It also contains some information about all the available standard library
-functions that are included. The standard library is a set of function
-included with the interpreter that extend its functionality over the
-available operators.
-
-The file \hrc{mapop.gperf} is also a source file. It is used in conjunction with the GNU gperf
+functions that are included. The standard library is a set of functions
+that are part of the interpreter and that extend its functionality.
+
+The file \hrc{mapop.gperf} is also a source file. It is used in conjunction with the \emph{GNU gperf}
 tool. The tool creates a perfect hash function, which is output to the file
 \hrc{src/interpreter/mappings/mapop.c}. The perfect hash function is used to translate
-single- or multi-character symbols, such as '!=' and '+', to their respective
+single- or multi-character symbols, such as \emph{!=} and \emph{+}, to their respective
 interpreter internal representation in constant time.
@@
 
-\subsubsection{Flowchart}
-
-\ver{
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=0.6\textwidth]{paper/Data Flow Diagram.png}
-    \caption{A \name{} Data Flow Diagram}
-    \label{f_diagram}
-\end{figure}
-}
-
-The above figure is a representation of how different components of the entire
-\name{} implementation interact and work together. Each next stage performs
-operations on the output data of the proper. The \hrc{src/interpreter/interpreter.c}
-file, here depicted as 'Interpreter', chains all stages together.
-
 \subsection{Implementing \name{}}
-This section will explain the process of constructing and the inner workings of
+This section will explain the process of constructing, and the inner workings of
 the \name{} interpreter. It is therefore the most complicated and detailed section
@@
 code from the project to show the decisions made during
-interpreter design. Only file names are used, as those are unique. To view the
-entire file path, look up the file names in the file tree on page \pageref{FileStructure}.
+interpreter design. Only the file names are used, as those are unique. To view the
+entire file paths, look up the file names in the file tree on page \pageref{FileStructure}.
 
 The syntax of \name{} is not specified in a formal programming language grammar,
-but rather implemented from scratch in the files src/interpreter/tokenizer.c 
-and src/interpreter/parser.c.
-
+but rather implemented from scratch in the files \hrc{src/interpreter/tokenizer.c}
+and \hrc{src/interpreter/parser.c}.
+
+\centerline{\begin{minipage}[H]{\linewidth}
 \subsubsection{The Entry-Point}
@@
 
+\end{minipage}}
+
 For the readers who do not know how to read source code, C in particular, here is a short
@@
 \begin{itemize}
-    \item \emph{Preprocessor Instructions.} Lines beginning with hashtags are
-        C preprocessor instructions, they are dealt with by the C preprocessor.
+    \item \emph{Preprocessor Instructions.} Lines beginning with hash symbols (\#) are
+        C preprocessor instructions and are dealt with by the C preprocessor.
         Preprocessing is what the compiler does before actually compiling the
-        code. %TODO: add src (Preprocessor)
-
-        The most common preprocessor instructions are the following.
+        code.
+
+        The most common preprocessor instructions are the following two.
         On one hand, there is \expr{\#include $<$header.h$>$}
         to include various definitions from a header file, i.e. to paste
-        its content at where the 'include' statement lies.
+        its content at where the \emph{include} statement lies.
         On the other hand exists \expr{\#define A B} used to
-        define A as an alias for B and replace every following occurrence of A
-        with B.
-
-    \item \emph{Functions.} In programming, functions or methods are
+        define \emph{A} as an alias for \emph{B} and replace every following occurrence of \emph{A}
+        with \emph{B}.
+
+    \item \emph{Functions.} In programming, functions, or methods, are
         a set of code that accomplish a certain task. They take in data,
@@
         function name, arguments and a function body in that order.
+
+        \centerline{\begin{minipage}[H]{\linewidth}
         Here is an example of a C function:
         \code{pseudocode/function.c}{1}{999}{C}
-        This would be an example of a main function, similar to the one
+        \end{minipage}}
+
+        This is a main function, similar to the one
         in \hrc{src/main.c}. It returns a value
-        of type \emph{int}, has two arguments, \emph{argc} and \emph{argv}, and a function body
-        which, in the case of the example above, prints out the value of the \emph{argc} argument.
+        of type \emph{int}, has two arguments, \emph{argc} and \emph{argv}, and a function body in
+        which the value of the \emph{argc} argument is printed out.
 
@@
         To dereference the pointer, i.e. to copy the value stored
-        at the memory address into another variable, another star is used like this:
-        \expr{FILE value = *input;}.
+        at the memory address into another variable, one more star is used like this:
+        \expr{FILE value = *input;.}
         \emph{value} now contains a copy of the actual FILE that \emph{input} was pointing to.
 \end{itemize}
+[Source \cte{cpreprocessor}]
 
 For those who are further interested in the C programming language, there are
-countless internet articles or tutorials about reading and understanding or
-writing and compiling C code. One such example is the article 'The C Beginner's
-Handbook' provided by freeCodeCamp.org linked in the sources
+countless internet articles or tutorials about reading and understanding, or
+writing and compiling C code. One such example is the article \emph{The C Beginner's
+Handbook} provided by freeCodeCamp.org linked in the sources
 on page \pageref{bibliography}.
@@
 
-Following is a very quick summary of the few 
+The next few sentences are a summary of all
 things happening in the \hrc{src/main.c} file.
@@
 At last, the function \emph{interpret}
-from the included header is called with the selected input.
+from the included header is called with the selected input and
+interpretation can begin.
 
@@
 Following is a breakdown of the \emph{interpret} function from said file
-as well as an overview of some of the stages that follow.
+as well as an overview of some of the sections that follow.
 
 The function \emph{interpret} calls the parser to generate a syntax-tree from tokens retrieved
-from the tokenizer implemented in \hrc{src/interpreter/parser.c} and 
+from the tokenizer, implemented in \hrc{src/interpreter/parser.c} and 
 \hrc{src/interpreter/tokenizer.c} respectively.
@@
 After that, the names of variables are
-mapped to addresses in memory where their value is stored. Doing this before
-bytecode compilation has a performance benefit. 
+mapped to addresses in memory where their value is stored. 
+Doing this before
+bytecode compilation has a performance benefit as the memory addresses can be
+cached and reused. 
+The file walking the AST and replacing
+each variable name with a memory address is called \hrc{src/interpreter/localizer.c}. The name
+localizer is not an industry standard and was invented for this project.
+
 The mapping is achieved with a custom hash table data structure with average time
 complexity of O(log n), n being the total amount of variables a \name{} program is using.
-The hash table implementation used in \hrc{src/interpreter/memory/hashtable.c}                   % TODO: maybe its own paragraph! interpolation search vs binary search
-uses binary search to locate elements variable-sized arrays in the hash table's buckets.
+The hash table implementation used in \hrc{src/interpreter/memory/hashtable.c}
+uses a binary search algorithm to locate elements in variable-sized arrays in the hash table's buckets.
 This is because for such few variables as are used in a standard \name{} program,
@@
 to an intermediate internal bytecode format by \hrc{src/interpreter/bytecode.c}.
-This step is done so that code that is run multiple times does not have
-the problem of time overhead when traversing the syntax tree.
+This step is done so that code that is run multiple times does not cause 
+a large time overhead when traversing the syntax tree.
 The final step is to process or execute said bytecode. This step is performed
@@
 These tokens represent groups of characters that belong together. In this paper's
-implementation, each token has a strict type and a text content.
-The different types of tokens and other things are defined in the header file
+implementation, each token has a strict type and a text content. In an IT context,
+a text or group of characters is often reffered to as string.
+The different types of tokens are defined in the header file
 \hrc{src/interpreter/tokenizer.h}.
 
-Here is an example of tokenization:
-An input of \emph{a.b==23.4} would result in the following tokens according
+Following is an example of tokenization.
+An input of \emph{a.b==23.4} would result in the the tokens depicted below according
 to the tokenizer implementation of this paper:
@@
 variable names to addresses in memory.
-SYMBOL type tokens are mapped to operators such as plus or minus by \hrc{src/implementation/mappings/mapop.c}
-during parsing. NUMBER type tokens are replaced by an actual number type, which
-is C's 'long double' type as specified in \hrc{src/interpreter/bytecode.h}.
+SYMBOL type tokens are mapped to operators such as plus or minus by \hrc{src/interpreter/mappings/mapop.c}
+during parsing. NUMBER type tokens are replaced by an actual number, for which
+C's \emph{long double} data type is used as specified in \hrc{src/interpreter/bytecode.h}.
 
 While tokenizers can be programmed using regular expressions, 
-this paper uses a simpler deterministic finite-state machine tokenizer.
+this paper uses a simpler \emph{deterministic finite-state machine tokenizer}.
 It reads character by character, determining the token type as it goes.
-It stops when there is a token that does no longer match the current token type.
+It stops when there is a character that does no longer conform to the rules of the current token type.
 Tokens are each requested and integrated into a syntax-tree by the parser as described in the next
@@
 also called parse tree, from tokens. Since this paper's implementation 
-does not parse code from a
-concise context free grammar, the syntax tree the parser creates is very
+does not parse code given a
+formal context free grammar, the syntax tree the parser creates is very
 similar to an otherwise known abstract syntax tree, AST for short.
@@
 
-The algorithm used is inspired by the 'Pratt Parsing
-algorithm' first described by Vaughan Pratt in 1973, which is a kind of precedence
+The algorithm used is inspired by \emph{the Pratt Parsing
+algorithm} first described by Vaughan Pratt in 1973, which is a kind of precedence
 parser based on recursive descent. There is an explanation of the algorithm
@@
 result in \emph{14} and is not the same as \emph{(2+3)*4}.
-Associativity is either 'Left to Right' or 'Right to Left'. An example of associativity is
+Associativity is either \emph{Left to Right} or \emph{Right to Left}. An example of associativity is
 \emph{20/10/2}. Both in C and in \name{}, the division operator is
@@
 \emph{(20/10)/2} and not as \emph{20/(10/2)}. Both precedence 
-and associativity are properties that are defined in the file \hrc{mapop.gpref} for
+and associativity are properties that are defined in the file \hrc{mapop.gperf} for
 each operator.
@@
         fields, strings, and it handles brackets. 
-        In addition, it can handle prefix operators, such as the unary plus or unary minus.
+        In addition, it can handle prefix operators, such as the unary plus or the unary minus.
     \item \emph{expr.} The \emph{expr} function calls \emph{secondary}, the result of which
@@
 Following is a piece of pseudocode to better illustrate the inner workings of the
-parser which is based on a Pratt-Parser. 
+parser which is based on a \emph{Pratt Parser}.
 This is only an example and not completely accurate to the actual parser used
-by this project.
-% TODO: validate code
+in this project.
+
 \code{pseudocode/pratt-parser.c}{1}{999}{C}
@@
 \subsubsection{The Bytecode Compiler}
-The bytecode compiler is an internal part of the interpreter, located in the file
+The bytecode compiler is an internal part of the interpreter and is located in the file
 \hrc{src/interpreter/bytecode.c}. Its job is to create a list
@@
 binary tree created by the parser and pushes found instructions onto a stack.
-The algorithm used is inspired by Edsger Dijkstra's 'Shunting yard 
-algorithm', traversing the tree post-order to create
+The algorithm used is inspired by Edsger Dijkstra's \emph{Shunting yard 
+algorithm}, traversing the tree post-order to create
 a prefix notation. 
 A prefix notation is a way to write down mathematical expressions, except
-operators precede their operands, in contrast to the more common infix notation,
+operators precede their operands in contrast to the more common infix notation,
 where operands surround the operator.
-[Source \cte{shunting_yard}]
 
@@
         \expr{\emph{- (/ (+ 6 5) (* 4 2)) 1}}
-    \item The brackets are not needed, because for every operator, the amount
+    \item The brackets are not needed, because for every operator the amount
         of operands is known. This is also the case for the unary minus and unary plus,
@@
 \end{enumerate}
-[Source \cte{polish_notation}]
-
-\subsubsection{Program Evaluation}
+[Sources \cte{polish_notation}, \cte{shunting_yard}]
+
+\subsubsection{Calculating Results}
 This is the last stage of interpreting program code as implemented in \name{}.
@@
 \begin{itemize}
-    \item Assume the input in \name{}'s bytecode to resemble the following expression: 
+    \item Assume the following input in \name{}'s bytecode format: 
         \expr{\emph{- 6 * 3 4}}
@@
 
-% TODO: add examples in /examples/
-\name{} is built upon the implementation of a recursive Pratt
-parser and a basic tokenizer. It uses a hash table to map variable names to their
+\name{} is built upon the implementation of a recursive Pratt Parser
+and a basic tokenizer. It uses a hash table to map variable names to their
 memory addresses. There are no keywords, only a set of non-alphanumeric characters
@@
 standard functions that are included in the interpreter and that enlarge the capabilities of the language.
-The following sections make up examples of the syntax and capabilities of the language and
+The following subsections display examples of the syntax and of the language and
 how programs are written in it.
@@
     Since the variable names \emph{a} and \emph{b} are not yet
-    known to the interpreter, they are automatically allocated in the current scope
-    and assigned a default value. That default value is then overwritten with the
+    known to the interpreter, memory is automatically allocated for them
+    and initiated to a default value. That default value is then overwritten with the
     assignment operator.
@@
     standard library function, meaning it is included in the \name{} interpreter and
-    written in C. Another function is passed as argument to the function call.
-    This function will be repeated.
+    written in C. A so-called \emph{anonymous function} is passed as argument to the function call.
+    It is called anonymous because it is not bound to an identifier. This anonymous function
+    will be repeated by \emph{std.repeat}.
 \item On line 7 and 8, some calculations are done.
@@
 \item Lastly, line 13 contains another calculation in which the variable \emph{count}
-    is decremented by one. Since this is the last instruction in the inner function,
-    it is simultaneously its return value, e.g. the value of count is returned
+    is decremented by one. Since this is the last instruction of the anonymous function,
+    it simultaneously is its return value, e.g. the value of count is returned
     to the \emph{std.repeat} function. The \emph{std.repeat} function will
-    repeat the inner function until the returned value is zero or undefined.
+    repeat the inner function until the returned value is zero.
     This means the inner function is executed exactly ten times, because
@@
 \end{itemize}
+[Source \cte{anonymous}]
 
@@
 program opens a YouTube video in Firefox. Because the start command is not
-available in \emph{GNU Bash} this instruction will most likely not work
+available in \emph{GNU Bash}, this instruction will most likely not work
 on GNU+Linux.
@@
 will still work. This is because a hash symbol does not simply delete
-a variable, but rather make it unusable for future references. To sum up,
+a variable, but rather make it inaccessible for future references. To sum up,
 the function \emph{f} is still capable of using \emph{std.exec}, because
-at the time \emph{f} was defined, \emph{std.exec} was still accessible.
+at the time \emph{f} was defined, \emph{std.exec} was still available.
 
 This security concept forces developers to think about access restriction. First, safe to use aliases
-to dangerous functions have to be created. Next, access to all those exploitable functions is
+to dangerous functions have to be created. Next, access to all exploitable functions is
 revoked. This way, any code can only access variables and functions that are still available.
@@
 \section{Discussion}
-The following paragraphs will go over some positive results and some caveats of the
-final product.
-
-The name \name{} was chosen, because it implied this programming language
+The name \name{} was chosen, because it implies this programming language
 to be a learning project as well as a scripting language, which employ
@@
 at a time. The file name extension \emph{.nts} is used in the examples. This
-file extension is not required, but was chosen as it stands for \emph{Not TypeScript}, because
+file extension is not mandatory, but was chosen as it stands for \emph{Not TypeScript}, because
 the language neither enforces strict typing, nor is it Microsoft's \emph{TypeScript}.
+
+The following paragraphs will go over some positive results and some caveats of the
+final product.
 
@@
 \subsubsection{Numbers}
-Another goal that turned out not to be practical to achieve was to use just as much
+Another goal that turned out not to be practical to achieve was the use of just as much
 memory for each variable as needed and not more. This means every variable would
-have to be assigned a datatype, containing information about the size and possible values.
+have to be assigned a datatype, containing information about the size and domain of possible values.
 Taking C as a reference, the datatype \emph{char} is exactly one byte in size, and
@@
 the datatype can be discarded at compile time, whereas in an interpreter like \name{} such
-information needs to persist throughout the program's lifetime.
+information needs to persist throughout the program's lifetime. This is because interpreters
+do not know the entire source code of the program before it is executed, while compilers do.
 
@@
 Memory leaks happen when a program reserves memory to store data, but
-never frees up the space for use for other programs, even though the data
+never frees up the space to be used by other programs, even though the data
 has long been forgotten about and will no longer be accessed.
-In \name{}, memory leaks are a major problem. Even though barely noticeable in small
+In \name{}, memory leaks are a major problem. Even though they are barely noticeable in small
 programs it interprets, \name{} does not have proper setups in place to free up
@@
 free it up automatically. While \name{} also tries to make use of this technique,
-there are many flaws introduced while developing the interpreter that hinder 
+there are many flaws that were introduced while developing the interpreter that hinder 
 Garbage-Collection from working correctly.
@@
 Rust has an ownership system in place,
-in which any allocated memory is owned by a function. When the function
+in which any allocated memory is owned by one function and can be borrowed by 
+multiple other functions. When the owning function
 is done processing, the memory is automatically freed. Such a system
@@
 lost in the details.
-Inventing new ways to make the job of developers easier is what programming
+Coming up with new ways to make the job of developers easier is what programming
 language design is all about.
@@
 the executable program against an input and finds the origins of not freed data.
+Even though it was used to fix some memory leaks, many remain.
 
@@
 Since most languages used today allow increasing a
-program's throughput using multi-threading, not having it at hand in \name{}
+program's throughput using multi-threading, not having the feature at hand in \name{}
 is a major limitation.
@@
 can share code. Examples would be \emph{npm} or \emph{cargo}, 
-the 'Node Package Manager' for JavaScript applications and Rust's package 
+the \emph{Node Package Manager} for JavaScript applications and Rust's package 
 manager respectively.
@@
 project to find errors in the \name{} implementation.
-[Sources \cte{npm}, \cte{cargo}]
 
@@
 and it can be compiled to machine-native programs via LLVM. 
-[Source \cte{kotlin_compile}]
+[Sources \cte{npm}, \cte{cargo}, \cte{kotlin_compile}, \cte{llvm}]
 
@@
 that are used in other programming languages, but 
-could not be implemented into \name{}.
+could not be implemented in \name{}.
 
@@
 is a mechanism by which one programming language can call functions written 
-in another. This makes a language more attractive, as it can be used with
-existing libraries from another language, instead of functions having to be
-re-written. Enabling C function calls often allows a certain language to
-directly interact with system libraries or the operating system itself.
+in another. This makes a language more attractive to new users, as it can be used
+in conjunction with existing libraries from other languages. 
+Enabling calls to foreign C function could allow programms written in a new langugage
+to directly interact with system libraries or the operating system itself.
 
@@
 
-Foreign function interfaces are specially difficult to implement, or implement
-right. Let us assume a programming language interpreter \name{} written in C that
+Foreign function interfaces are very difficult to implement correctly.
+Let us assume a programming language interpreter \name{} written in C that
 would like to call some functions that are part of yet another C program or
@@
 structure of the functions to be called, say return- or argument-types, which
-is by the way where the term foreign functions originates from. This is why the
+is by the way where the term \emph{foreign functions} originates from. This is why the
 \name{} developer would have to provide that information by re-declaring the
-function inside \name{}, so that, with some trickery, data can be passed to and
+function inside of \name{}, so that, with some trickery, data can be passed to and
 received from the function in the correct format, preventing a segmentation
@@
 Python for this purpose, can be used to load and call such foreign functions.
+The library was not used in this paper's implementation, as the compilation process
+on Microsoft Windows would have become too difficult.
 [Sources \cte{FFI}, \cte{JNI}, \cte{ctypes}, \cte{libffi}]
@@
 of objects.
-The easiest and simplest way to go about implementing Reflection in \name{} is
-certainly to allow the interpretation of strings at runtime. That way, strings
+The easiest and simplest way to go about implementing reflection in \name{} would 
+certainly be to allow the interpretation of strings, i.e. text, at runtime. That way, strings
 holding program instructions can be created and modified and later executed.
+The difficulty of implementing reflection in the current state of \name{} 
+lies in the fact that either syntax trees that exist already would have to be 
+altered according to the injected source code, or syntactical compromises would have to be made.
+This is why reflection is not a feature of \name{}.
 [Source \cte{reflection}]
@@
 problems and solutions. The goal of this project to learn about 
-the inner workings of an interpreter and
-to get to know the C language has been achieved.
+the inner workings of an interpreter has been achieved.
+With the source code of the project having exceeded 3500 lines, getting to
+know the C programming language better was a success as well.
 
@@
 This section contains a bibliography which links to all online articles
-this project could profit from, and the list of figures.
+this project could profit from and the respective dates of last access, and the list of figures.
 The entire source code of \name{} can be found in the Appendix, section \ref{Appendix}.
-
-\subsection{Bibliography}
-\renewcommand*{\bibfont}{\normalsize}
-\pagelabel{bibliography}\printbibliography[heading=none]
 
@@
 
+\subsection{Bibliography}
+\renewcommand*{\bibfont}{\normalsize}
+\pagelabel{bibliography}\printbibliography[heading=none]
+
 \section{Declaration of Authenticity}
@@
 listed under the \emph{Bibliography} section on page \pageref{bibliography}.
-% TODO: check correctness of all file paths!! e.g. src/...
-
-\newpage\section{Appendix}\label{Appendix}
+
+\section{Appendix}\label{Appendix}
 The appendix contains the complete source code of \name{}.

### 2022-10-30 pseudocode/pratt-parser.c

@@ -1,36 +1,27 @@
-// implement functions
 stnode* parse() {
-    // just call expr
-    return expr(0);
+    return expr(0); // parse is a wrapper function for expr
 }
 
 stnode *expr(int precedence) {
-    // read secondary
-    stnode *left = secondary();
+    stnode *left = secondary(); // Generate a secondary node
     stnode *ret = left;
 
-    // reached end of file
-    if(left == NULL)
-        return NULL;
+    if(left == NULL) return NULL; // Reached the end of the source code
 
     token next;
     while(precedence_of(next = peek_token()) > precedence 
             && next.type == SYMBOL) {
-        // allocate a new node
-        ret = allocate_stnode(EXPR);
-        ret->left_child = sec;
+        ret = allocate_stnode(EXPR); // Allocate a new node
+        ret->left_child = left;
         ret->right_child = expr(presedence_of(next) - associativity_of(next));
         ret->operator = operator_of_token(next);
 
-        // throw away token
-        next = next_token();
-
-        sec = ret;
+        next = advance(); // Throw away the token, request a new one
+        left = ret;
     }
     return ret;
 }
 
 stnode *secondary() {
-    // dealing with brackets left out, see full source code...
-    return stnode_from(next_token()); 
+    return stnode_from(advance());  // See the full source code for details
 }

### 2022-10-30 paper.tex
final touches

@@
 
-\author{Florian M. Donnelly}
 \title{The Development of a Programming Language}
@@
     \vfill
-    Matura Paper written by\\ \textbf{Florian M. Donnelly}
+    Matura Paper written by\\ \ver{\textbf{Florian M. Donnelly}}
 
     \vspace{0.5cm}
-    Under Supervision of\\ \textbf{Adrian Lüthi}
+    Under Supervision of\\ \ver{\textbf{Adrian Lüthi}}
 
     \vspace{0.5cm}
-    Gymnasium Burgdorf\\ October 31, 2022
+    Gymnasium Burgdorf\ver{, 23e}\\ October 31, 2022
 
@@
 
-    \url{https://github.com/LeFlorians/TestScript}.
+    \ver{\url{https://github.com/LeFlorians/TestScript}.}
 
@@
 
-    \url{https://github.com/LeFlorians/TestScript/tree/\githash}.
+    \ver{\url{https://github.com/LeFlorians/TestScript/tree/\githash}.}
 
@@
 
-\newpage\subsection{A Project Overview}
+\subsection{A Project Overview}
 In the following paragraphs, the project in its entirety will be briefly
@@
 
-\section{Declaration of Authenticity}
-Hereby I, Florian M. Donnelly, declare to have written this paper independently, 
-also including the entire source code of \name{}, using resources 
-listed under the \emph{Bibliography} section on page \pageref{bibliography}.
+\section{Declaration of Authorship}
+By typing my full name below, I confirm that the matura paper I am submitting is entirely
+my own original work except where otherwise indicated. I am aware of the shool regulations
+concerning plagiarism.
+
+\ver{Florian Michael Donnelly}
 
